### 2025-02-01 products/manuscript/Manuscript.qmd
Part 1 project

---
title: "The effects of multi environmental exposure on Tuberculosis Incidence in Asian and African countris"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

* Muhammad Nasir$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
* Second Author$^{3, *}$
* Third Autor$^{3, \dagger}$
* Last Author$^{1, 2, \land}$

**Author affiliations**

1. College of Public Health, University of Georgia, Athens, GA, USA.
2. Another Affiliation, Athens, GA, USA.
3. Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information

Tuberculosis is old infectious disease which still exists and become major public health in the world, particularly in developing countries. Tuberculosis is linked to environmental factors. This study will analyze the effects of multi environmental factors on tuberculosis incidence in Asia and Afirca regions, where the majority of tuberculosis coming from these regions. The environmental factors and country economic and hman development index as covariates.

## Description of data and data source

The study will use multi sources of data including World Bank data sources, World Health Organization (WHO), United Nation for Children Fund (Unicef), and UN-HABITAT, Emissions Database for Global Atmospheric Researcg (EDGAR), and Institute for Health Metrics and Evaluation (IHME). The data contains informations of all variables in all countries globally from 2000 to 2020. The variables will include  proportion of urban opulation living in slum areas, percentage of popuation using safely managed sanitation services, percentage of population using at least basic drinking water sevice, fule for cooking, Population-weighted exposure to ambient PM2.5 pollution, total green houses, and will include GDP/ country economic status, human development index, and Universal Health Caverage as covariates.

Note: I have downloaded the dataset, but I have not conducted any cleaning or exploration on the dataset. 
List of data sourses and website is provided in an exel fine in my data folder. I will include them in my reference for the next steps.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

The main aim of this study is to look at the impacts of environmental factors on tuberculosis incidence.

RQ: What enironmental factors asociated with tuberculosis incidence?

H0: there is no relationship between proportion ofurban population living in slum areas and tuberculosis. 
H0: thee is no relationship between safely managed santation and tuberculosis incidence 
H0: there is no relatinship between drinking water and tuberculosis incidence 
H0: there is no relationship between fule for cooking and tuberculosis incidence 
H0: there is no relationship between PM 2.5 exposure and tuberclosis incidence 
H0: there is no association between total green gass houses and tuberculosis incidence

Ha: there is  relationship between proportion ofurban population living in slum areas and tuberculosis. 
Ha: thee is relationship between safely managed santation and tuberculosis incidence 
Ha: there is relatinship between drinking water and tuberculosis incidence 
Ha: there isrelationship between fule for cooking and tuberculosis incidence 
Ha: there is relationship between PM 2.5 exposure and tuberclosis incidence 
Ha: there is association between total green gass houses and tuberculosis incidence

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

I haven't decide about data analysis methods. However, I am thinking to perform time series analysis and multi linear regression (if suitable) or Generealized linear model. I am still in the process of reading more papers to get more informatin which methods best suit this data and answer the research questions.

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

I will add the code.

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-02-02 products/manuscript/Manuscript.qmd
Install renv() to make the project reproducible
fixed some typo

@@
 csl: ../../assets/american-journal-of-epidemiology.csl
+editor: 
+  markdown: 
+    wrap: 72
 ---
 
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
+The structure below is one possible setup for a manuscript, or a general
+data analysis project (including the course project). Adjust as needed.
+You don't need to have exactly these sections, but the content covering
+those sections should be addressed.
 
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
+This uses MS Word as output format. [See
+here](https://quarto.org/docs/output-formats/ms-word.html) for more
+information. You can switch to other formats, like html or pdf. See [the
+Quarto documentation](https://quarto.org/) for other formats.
 
@@
 
-* Muhammad Nasir$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
-* Second Author$^{3, *}$
-* Third Autor$^{3, \dagger}$
-* Last Author$^{1, 2, \land}$
+-   Muhammad Nasir$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
+-   Second Author$^{3, *}$
+-   Third Author$^{3, \dagger}$
+-   Last Author$^{1, 2, \land}$
 
@@
 1.  College of Public Health, University of Georgia, Athens, GA, USA.
-2. Another Affiliation, Athens, GA, USA.
+2.  Faculty of Public Health, Universitas Tadulako, Palu, Indonesia.
 3.  Yet another Affiliation.
@@
 
-$\land$ Corresponding author: some@email.com
+$\land$ Corresponding author: some\@email.com
 
@@
 # Summary/Abstract
-_Write a summary of your project._
+
+*Write a summary of your project.*
 
@@
 
-Tuberculosis is old infectious disease which still exists and become major public health in the world, particularly in developing countries.
-Tuberculosis is linked to environmental factors.
-This study will analyze the effects of multi environmental factors on tuberculosis incidence in Asia and Afirca regions, where the majority of tuberculosis coming from these regions.
-The environmental factors and country economic and hman development index as covariates.
+Tuberculosis is old infectious disease which still exists and become
+major public health in the world, particularly in developing countries.
+Tuberculosis is linked to environmental factors. This study will analyze
+the effects of multi environmental factors on tuberculosis incidence in
+Asia and Africa regions, where the majority of tuberculosis coming from
+these regions. The environmental factors and country economic and human
+development index as covariates.
 
@@
 
-The study will use multi sources of data including World Bank data sources, World Health Organization (WHO), United Nation for Children Fund (Unicef), and UN-HABITAT, Emissions Database for Global Atmospheric Researcg (EDGAR), and Institute for Health Metrics and Evaluation (IHME).
-The data contains informations of all variables in all countries globally from 2000 to 2020.
-The variables will include  proportion of urban opulation living in slum areas, percentage of popuation using safely managed sanitation services, percentage of population using at least basic drinking water sevice, fule for cooking, Population-weighted exposure to ambient PM2.5 pollution, total green houses, and will include GDP/ country economic status, human development index, and Universal Health Caverage as covariates.
+The study will use multi sources of data including World Bank, World
+Health Organization (WHO), United Nation for Children Fund (Unicef),
+UN-HABITAT, Emissions Database for Global Atmospheric Research (EDGAR),
+and Institute for Health Metrics and Evaluation (IHME). Different
+dataset contains different variable, however all dataset have same year,
+country name and country codes that can be used to combine the dataset.
+The data also contain information of countries globally from 2000 to
+2020. The variables will include proportion of urban population living
+in slum areas, percentage of population using safely managed sanitation
+services, percentage of population using at least basic drinking water
+service, fuel for cooking, Population-weighted exposure to ambient PM2.5
+pollution, total green houses, and will include GDP/ country economic
+status, human development index, and Universal Health Coverage as
+covariates.
 
-Note: I have downloaded the dataset, but I have not conducted any cleaning or exploration on the dataset. 
-List of data sourses and website is provided in an exel fine in my data folder. I will include them in my reference for the next steps.
+Note: I have downloaded the dataset, but I have not conducted any
+cleaning or exploration on the dataset. List of data sources and website
+is provided in an excel fine in my data folder. I will include them in
+my reference for the next steps.
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
 
-The main aim of this study is to look at the impacts of environmental factors on tuberculosis incidence.
+*State the research questions you plan to answer with this analysis.*
 
-RQ: What enironmental factors asociated with tuberculosis incidence?
+The main aim of this study is to look at the impacts of environmental
+factors on tuberculosis incidence.
 
-H0: there is no relationship between proportion ofurban population living in slum areas and tuberculosis. 
-H0: thee is no relationship between safely managed santation and tuberculosis incidence 
-H0: there is no relatinship between drinking water and tuberculosis incidence 
-H0: there is no relationship between fule for cooking and tuberculosis incidence 
-H0: there is no relationship between PM 2.5 exposure and tuberclosis incidence 
-H0: there is no association between total green gass houses and tuberculosis incidence
+RQ: What environmental factors associated with tuberculosis incidence?
 
-Ha: there is  relationship between proportion ofurban population living in slum areas and tuberculosis. 
-Ha: thee is relationship between safely managed santation and tuberculosis incidence 
-Ha: there is relatinship between drinking water and tuberculosis incidence 
-Ha: there isrelationship between fule for cooking and tuberculosis incidence 
-Ha: there is relationship between PM 2.5 exposure and tuberclosis incidence 
-Ha: there is association between total green gass houses and tuberculosis incidence
+H0: there is no relationship between proportion of urban population
+living in slum areas and tuberculosis.
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
+H0: thee is no relationship between safely managed sanitation and
+tuberculosis incidence
 
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+H0: there is no relationship between drinking water and tuberculosis
+incidence
+
+H0: there is no relationship between fuel for cooking and tuberculosis
+incidence
+
+H0: there is no relationship between PM 2.5 exposure and tuberculosis
+incidence
+
+H0: there is no association between total green gas houses and
+tuberculosis incidence
+
+Ha: there is a relationship between proportion of urban population
+living in slum areas and tuberculosis.
+
+Ha: there is a relationship between safely managed sanitation and
+tuberculosis incidence
+
+Ha: there is a relationship between drinking water and tuberculosis
+incidence
+
+Ha: there is a relationship between fuel for cooking and tuberculosis
+incidence
+
+Ha: there is a relationship between PM 2.5 exposure and tuberculosis
+incidence
+
+Ha: there is a association between total green gas houses and
+tuberculosis incidence
+
+To cite other work (important everywhere, but likely happens first in
+introduction), make sure your references are in the bibtex file
+specified in the YAML header above and have the right bibtex key. Then
+you can include like this:
+
+Examples of reproducible research projects can for instance be found in
+[@mckay2020; @mckay2020a].
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+*Describe your methods. That should describe the data, the cleaning
+processes, and the analysis approaches. You might want to provide a
+shorter description here and all the details in the supplement.*
 
-I haven't decide about data analysis methods.
-However, I am thinking to perform time series analysis and multi linear regression (if suitable) or Generealized linear model.
-I am still in the process of reading more papers to get more informatin which methods best suit this data and answer the research questions.
+I haven't decide about data analysis methods. However, I am thinking to
+perform time series analysis and multi linear regression (if suitable)
+or Gene-realized linear model (we will see after conducting data
+exploration). I am still in the process of reading more papers to get
+more information which methods best suit this data and answer the
+research questions.
 
@@
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+Sometimes you might want to show a schematic diagram/figure that was not
+created with code (if you can do it with code, do it). @fig-schematic is
+an example of some - completely random/unrelated - schematic that was
+generated with Biorender. We store those figures in the `assets` folder.
 
@@
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+
+Most of the data is downloaded from Worlbank Dataset. However, those
+data are originally from many different courses/ organizations, such as
+World Bank, WHA, UNICEF, UN-HABITAT, EDGAR, and IHME.
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
 
-I will add the code.
+*Write code that reads in the file and cleans it so it's ready for
+analysis. Since this will be fairly long code for most dataset, it might
+be a good idea to have it in one or several R scripts. If that is the
+case, explain here briefly what kind of cleaning/processing you do, and
+provide more details and well documented code somewhere (e.g. as
+supplement in a paper). All materials, including files that contain
+code, should be commented well so everyone can follow along.*
+
+Dataset are either in excel or CSV files. Several steps from loading to
+data analysis:
+
+-   to load the data set in this part using read_excel or read_csv.
+
+-   to combine them base on country name/ code and years.
+
+<!-- -->
+
+-   to drop unwanted variables
+
+-   to select Asian and African Countries only
+
+-   to drop N/A or missing value
+
+-   to conduct data exploration
+
+-   to decide suitable analysis method
+
+-   to perform data analysis
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+
+*Explain anything related to your statistical analyses.*
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+*Use a combination of text/tables/figures to explore and describe your
+data. Show the most important descriptive results here. Additional ones
+should go in the supplement. Even more can be in the R and Quarto files
+that are part of your project.*
 
@@
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+Note the loading of the data providing a **relative** path using the
+`../../` notation. (Two dots means a folder up). You never want to
+specify an **absolute** path like `C:\ahandel\myproject\results\`
+because if you share this with someone, it won't work for them since
+they don't have that path. You can also use the `here` R package to
+create paths. See examples of that below. I generally recommend the
+`here` package.
 
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+*To get some further insight into your data, if reasonable you could
+compute simple statistics (e.g. simple models with 1 predictor) to look
+for associations between your outcome(s) and each individual predictor
+variable. Though note that unless you pre-specified the outcome and main
+exposure, any "p\<0.05 means statistical significance" interpretation is
+not valid.*
 
@@
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
+*Use one or several suitable statistical/machine learning methods to
+analyze your data and to produce meaningful figures, tables, etc. This
+might again be code that is best placed in one or several separate R
+scripts that need to be well documented. You want the code to produce
+figures and data ready for display as tables, and save those. Then you
+load them here.*
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+
+*Summarize what you did, what you found and what it means.*
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+
+*Discuss what you perceive as strengths and limitations of your
+analysis.*
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+*What are the main take-home messages?*
+
+*Include citations in your Rmd file using bibtex, the list of references
+will automatically be placed at the end*
 
@@
 
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
+These papers [@mckay2020; @mckay2020a] are good examples of papers
+published using a fully reproducible setup similar to the one shown in
+this template.
 
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+Note that this cited reference will show up at the end of the document,
+the reference formatting is determined by the CSL file specified in the
+YAML header. Many more style files for almost any journal [are
+available](https://www.zotero.org/styles). You also specify the location
+of your bibtex reference file in the YAML. You can call your reference
+file anything you like.
 

### 2025-02-23 code/analysis-code/statistical-analysis.qdm.qmd
Working on Step two of Final Project

---
title: "statistical-analysis"
format: html
editor: visual
---

##

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
install.packages("maps")
library(maps)
install.packages("ggcorrplot")
library(ggcorrplot)
install.packages("plm")
library(plm)
install.packages("gt")
library(gt)
install.packages("car")
library(car)
install.packages("MASS")  # Jika belum terinstall
library(MASS)
install.packages("tidyverse")
library(tidyverse)
install.packages("reshape2")  # Install the package (only needed once)
library(reshape2)  # Load the package
install.packages("scales")
library(scales)

library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)
# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)

install.packages("gridExtra")  # If not installed
library(gridExtra)
```

Load the dataset
```{r}

asia_africa_location <- here("data", "processed-data", "asia_africa.rds")
readRDS(asia_africa_location)

colnames(asia_africa)
```

## Data Analysis

***Test the linearity***

```{r}

# Create a list of scatter plots for each independent variable against tbi
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
    geom_point() + 
    ggtitle(paste("tbi vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

# Combine all plots into a single frame
combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want

combineplot_location <- here( "code","analysis-code", "figures", "scatterplot.png") # to set up location for the pictures created 
ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(combined_plot)

```

```{r}

# Step 1: Ensure all numeric variables are correctly formatted
asia_africa <- asia_africa %>%
  mutate(across(c(tbi, pro_cleanfuel, pro_basicwater, pro_safewater, 
                  mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, 
                  pro_safesanitation, gdp, hdi), as.numeric)) %>%
  drop_na()  # Remove rows with missing values to avoid ggplot errors

asia_africa <- asia_africa %>%
  mutate(poly_greenhouses = as.numeric(poly(total_greenhouses, degree = 2, raw = TRUE)[,1]))

# Step 2: Reshape the dataset for visualization
asia_africa_long <- asia_africa %>%
  select(-country, -region, -year) %>%  # Exclude categorical variables
  pivot_longer(cols = -tbi, names_to = "predictor", values_to = "value")

# Step 3: Check if pivot_longer() worked correctly
print(head(asia_africa_long))

# Step 4: Generate scatterplots with loess smoothing to detect nonlinear patterns
plot_nonlinear <- ggplot(asia_africa_long, aes(x = value, y = tbi)) +
  geom_point(alpha = 0.5) +  # Scatterplot points
  geom_smooth(method = "loess", color = "red") +  # Loess smoothing for nonlinear trends
  facet_wrap(~predictor, scales = "free") +  # Create separate plots for each predictor
  theme_minimal()

# Step 5: Display the plot
print(plot_nonlinear)

nonlinear_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
ggsave(filename = nonlinear_location, plot=plot_nonlinear, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(combined_plot)
```

```{r}
glm_model <- glm(tbi ~ gdp + hdi + pro_basicsanitation + pro_basicwater + 
                 pro_safesanitation + pro_safewater + pro_cleanfuel + 
                 mean_exposure_pm2.5 + total_greenhouses, 
                 family = poisson(link = "log"), data = asia_africa)
summary(glm_model)

plot(glm_model)

```
Q-Q plot of residuals from the Poisson GLM suggests that the model is not fitting the data well. The deviation from the straight line, especially at higher quantiles (outliers labeled 78, 79, 80), indicates overdispersion—where the variance of tbi is larger than the mean, violating the Poisson assumption.

```{r}
dispersion_test <- sum(residuals(glm_model, type="pearson")^2) / glm_model$df.residual
dispersion_test  # If > 1.5, overdispersion exists

```
since the result is very high, which should be close to 1. We need a model that can hanlde overdispersion

```{r}
library(MASS)
glm_nb <- glm.nb(tbi ~ gdp + hdi + pro_basicsanitation + pro_basicwater + 
                 pro_safesanitation + pro_safewater + pro_cleanfuel + 
                 mean_exposure_pm2.5 + total_greenhouses, data = asia_africa)
summary(glm_nb)

```

Interpretation:

1. Overdispersion Check
The estimated Theta (4.246, SE = 0.430) suggests that overdispersion is present but is well-accounted for by the Negative Binomial model.
The Residual Deviance (196.14) is much lower than the Null Deviance (409.01), meaning the model fits the data better than a null model.
Dispersion parameter = 1, confirming that the Negative Binomial model is appropriate.

The predictors that significantly impact TBI (tbi) are:

- hdi (-7.642, p < 2e-16) Higher HDI is associated with lower tbi cases.
- pro_basicsanitation (0.00577, p = 0.0334)  Increased access to basic sanitation is linked to a slight increase in tbi.
- pro_basicwater (0.0318, p = 9.42e-16) → Higher access to basic water is positively associated with tbi.
- pro_safewater (-0.00856, p = 0.0414) → Increased safe water access reduces tbi.
- pro_cleanfuel (-0.005999, p = 0.0137) → More use of clean fuels is linked to lower tbi.
- mean_exposure_pm2.5 (-0.02698, p < 2e-16) → Higher PM2.5 exposure reduces tbi (unexpected; may need further investigation).

Data Transfomation

```{r}
asia_africa <- asia_africa %>%
  mutate(log_tbi = log(tbi + 1),
         log_cleanfuel = log(pro_cleanfuel + 1),
         log_basicwater = log(pro_basicwater + 1),
         log_basicsanitation = log(pro_basicsanitation + 1),
         log_safewater = log(pro_safewater + 1),
         log_safesanitation = log(pro_safesanitation + 1),
         poly_greenhouses = poly(total_greenhouses, 2))  # 2nd-degree polynomial

```

```{r}
library(mgcv)

gam_model <- gam(tbi ~ s(pro_cleanfuel) + s(pro_basicwater) + s(pro_safewater) +
                   s(mean_exposure_pm2.5) + s(total_greenhouses) +
                   s(pro_basicsanitation) + s(pro_safesanitation) +
                   s(gdp) + s(hdi), data = asia_africa, method = "REML")

summary(gam_model)
gamplot<- plot(gam_model, pages = 1)

```

```{r}
# Calculate correlations
cor(asia_africa$tbi, asia_africa$gdp)
cor(asia_africa$tbi, asia_africa$mean_exposure_pm2.5)
cor(asia_africa$tbi, asia_africa$hdi)
cor(asia_africa$tbi, asia_africa$pro_cleanfuel)
cor(asia_africa$tbi, asia_africa$pro_basicwater)
cor(asia_africa$tbi, asia_africa$pro_safewater)
cor(asia_africa$tbi, asia_africa$pro_basicsanitation)
cor(asia_africa$tbi, asia_africa$pro_safesanitation)
cor(asia_africa$tbi, asia_africa$mean_exposure_pm2.5)

```

Spearman's Rank Correlation (Recommended) Spearman correlation is a non-parametric method that measures monotonic relationships (whether the variables move together, not necessarily linearly).

```{r}

# Compute Spearman correlation matrix
cor_matrix_spearman <- cor(asia_africa[, c("tbi","gdp", "hdi", "pro_cleanfuel", "pro_safewater", 
                                           "pro_basicsanitation", "pro_safesanitation", "mean_exposure_pm2.5", "total_greenhouses")], method = "spearman", use = "complete.obs")

# Melt the matrix into long format for ggplot

cor_melted <- melt(cor_matrix_spearman)

# Plot heatmap
plot5 <- ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  geom_text(aes(label = round(value, 2)), size = 4) +  # Show correlation values
  theme_minimal() +
  labs(title = "Spearman Correlation Heatmap", x = "", y = "")
matrix_location <- here( "code","eda-code", "figures", "correlation_matrix.png") # to set up location for the pictures created 
ggsave(filename = matrix_location, plot=plot5, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(plot5)
```

```{r}

#| label: fig-Spearman's Correlation Matrix 
#| fig-cap: "Spearman's Correlation Matrix "
#| echo: FALSE
knitr::include_graphics(here("code", "eda-code", "figures", "correlation_matrix.png"))
```

I will performed longitudinal data analysis

```{r}
library(car)
#vif(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
       pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, data = pdata)) # The error might be occure due to multicolinearity

#alias(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
         pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, data = asia_africa))

stepwise_model <- stepAIC(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
                             pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, 
                             data = asia_africa), direction = "both") # drop variable with multicolinearity

summary(stepwise_model)

```

***Performing longitudinal analysis***

Test for mixed or fixed effect using Hausman Test

```{r}
pdata <- pdata.frame(asia_africa, index = c("country", "year"))

fe_model <- plm(tbi ~ pro_basicsanitation + mean_exposure_pm2.5, 
                data = pdata, model = "within")
re_model <- plm(tbi ~ pro_basicsanitation + mean_exposure_pm2.5, 
                data = pdata, model = "random")

hausman_test <- phtest(fe_model, re_model)
print(hausman_test)

```

Based on the Hausman Test, fixed effect is not consistent. Therefore, I will use mixed effect test.

```{r}
summary(re_model)
```

Interpretation: ***1. variance components Analyisis***

88% dari total variasi berasal dari perbedaan antar negara, menunjukkan bahwa perbedaan karakteristik tiap negara sangat berpengaruh terhadap insiden TBC.

12% berasal dari variasi idiosinkratik, atau faktor yang berubah dari waktu ke waktu dalam suatu negara.

***2. Theta***

Rata-rata theta = 0.8868, menunjukkan bahwa sebagian besar variasi dalam data disebabkan oleh perbedaan antar negara. Theta yang mendekati 1 berarti bahwa individu (negara) memiliki efek spesifik yang kuat dibandingkan variasi waktu.

### 2025-02-23 code/eda-code/eda.qmd
Working on Step two of Final Project

---
title: "An example exploratory analysis script"
date: "2024-02-07"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I \# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
install.packages("maps")
library(maps)
install.packages("ggcorrplot")
library(ggcorrplot)
install.packages("plm")
library(plm)
install.packages("gt")
library(gt)
install.packages("car")
library(car)
install.packages("MASS")  # Jika belum terinstall
library(MASS)
install.packages("tidyverse")
library(tidyverse)
install.packages("reshape2")  # Install the package (only needed once)
library(reshape2)  # Load the package
install.packages("scales")
library(scales)

library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)
# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","clean_noslum.rds")

#load data
mydata <- readRDS(data_location)

#subset for south east and Asia Africa only 
asia_africa <- mydata %>% 
  filter(region == c("South-East Asia", "Africa"))
unique(asia_africa$country)

saveRDS(asia_africa, here("data", "processed-data", "asia_africa.rds"))
```

# Data exploration through tables

```{r}

# Define distinct colors for each country
country_colors <- c(
  "Algeria" = "#E6194B",  # Red
  "Bangladesh" = "#3CB44B",  # Green
  "Central African Republic" = "#FFE119",  # Yellow
  "Chad" = "#4363D8",  # Blue
  "Ethiopia" = "#F58231",  # Orange
  "Ghana" = "#911EB4",  # Purple
  "Indonesia" = "#42D4F4",  # Light Blue
  "Lesotho" = "#F032E6",  # Pink
  "Madagascar" = "#FABEBE",  # Light Red
  "Malawi" = "#469990",  # Teal
  "Myanmar" = "#E6BEFF",  # Lavender
  "Nepal" = "#9A6324",  # Brown
  "Senegal" = "#FFFAC8",  # Cream
  "Sri Lanka" = "#800000",  # Maroon
  "Tanzania" = "#A9A9A9",  # Gray
  "Togo" = "#FFD8B1",  # Peach
  "Uganda" = "#000075",  # Dark Blue
  "Zimbabwe" = "#808000"  # Olive
)

# Convert country to a factor
asia_africa$country <- as.factor(asia_africa$country)

# Create boxplot with borders
plot1 <- ggplot(asia_africa, aes(x = country, y = tbi, fill = country)) +
  geom_boxplot(color = "black", size = 0.1) +  # Add black border with thicker lines
  scale_fill_manual(values = country_colors) +  # Manually assign colors
  labs(title = "Boxplot of TBI by Country",
       x = "Country",
       y = "TBI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate labels for readability
        panel.border = element_rect(color = "black", fill = NA, size = 1.5))  # Add outer frame

plot_box <- here( "code","eda-code", "figures", "boxplot.png") # to set up location for the pictures created 
ggsave(filename = plot_box, plot=plot1, width = 12, height = 8, units = "in", dpi = 300) # save the pictures created

plot(plot1)

```

```{r}

# Select only numeric columns (excluding country, year, and region)
asia_africa_long <- asia_africa %>%
  select(-region) %>%  # Exclude categorical variable
  pivot_longer(cols = -c(country, year),  # Convert all numeric variables except country & year
               names_to = "Variable", 
               values_to = "Value")

# Create line plots for all variables
plot2 <- ggplot(asia_africa_long, aes(x = year, y = Value, group = country, color = country)) +
  geom_line(size = 1, alpha = 0.8) +  # Alpha for transparency
  facet_wrap(~ Variable, scales = "free_y") +  # Separate plots for each variable
  labs(title = "Trends of All Variables Over Time",
       x = "Year",
       y = "Value",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "none",  # Hide legend for clarity
        axis.text.x = element_text(angle = 45, hjust = 1))

plot_spageti <- here( "code","eda-code", "figures", "spageti.png") # to set up location for the pictures created 
ggsave(filename = plot_spageti, plot=plot2, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created

plot(plot2)
```

```{r}
plot3 <- ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.7) +  # Transparency to see overlapping lines
  labs(title = "Spaghetti Plot of TBC Incidence",
       x = "Year",
       y = "TBC Incidence") +
  theme_minimal() +
  theme(legend.position = "none")

plot_spageti_all <- here( "code","eda-code", "figures", "spageti-tbi.png") # to set up location for the pictures created 
ggsave(filename = plot_spageti_all, plot=plot3, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created

plot(plot3)
```

***Creating plots to show the trend of all variables in each country over the year***

```{r}
# Find a scaling factor to adjust TBI to match percentage values
max_indicator <- max(asia_africa[, c("pro_safesanitation", "pro_basicsanitation", "pro_safewater", "pro_basicwater", "pro_cleanfuel")], na.rm = TRUE)
max_tbi <- max(asia_africa$tbi, na.rm = TRUE)
scale_factor <- max_indicator / max_tbi  # Scale TBI to fit

# Reshape the data from wide to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_safesanitation, pro_basicsanitation, pro_safewater, pro_basicwater, pro_cleanfuel),
               names_to = "Indicator", 
               values_to = "Value")

# Create the plot
plot4 <- ggplot() +
  # First, plot water, sanitation, and clean fuel indicators
  geom_line(data = asia_africa_long, 
            aes(x = year, y = Value, color = Indicator, group = Indicator), 
            size = 1) +
  # Add TBI as a separate line with a different scale
  geom_line(data = asia_africa, 
            aes(x = year, y = tbi * scale_factor, color = "TBI"), 
            size = 1, linetype = "dashed") + 
  facet_wrap(~ country) +  # Separate plots for each country
  theme_minimal() +
  labs(title = "Percentage of Population with Access to Water, Sanitation, Clean Fuel & TBI Trends",
       x = "Year",
       y = "Percentage (%)",
       color = "Indicator") +  # Label for legend
  scale_y_continuous(sec.axis = sec_axis(~ . / scale_factor, name = "TBI")) +  # Secondary Y-axis for TBI
  scale_color_manual(values = c("blue", "green", "purple", "orange", "brown", "red")) +  # Custom colors
  theme(legend.position = "bottom",  
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, size = 1))  # Rotate x-axis labels for readability

plot_location1 <- here( "code","eda-code", "figures", "line-graph-tbivsfactors.png") # to set up location for the pictures created 
ggsave(filename = plot_location1, plot=plot4, width = 12, height = 16, units = "in", dpi = 300) # save the pictures created 
plot(plot4)
```

Check the layout after saving the figures

```{r}

#| label: fig-multivariable 
#| fig-cap: "Line charts of all variables"
#| echo: FALSE
knitr::include_graphics(here("code", "eda-code", "figures", "line-graph-tbivsfactors.png"))
```

***I want to show the mean of each variable every year based on regions***

```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(kableExtra)
library(here)

# Define row colors for each category
row_colors <- c(
  "TBI" = "#FFD700", # Gold
  "GDP" = "#ADD8E6", # Light Blue
  "HDI" = "#90EE90", # Light Green
  "Clean Fuel" = "#FFB6C1", # Light Pink
  "Safe Water" = "#87CEEB", # Sky Blue
  "Basic Sanitation" = "#FFA07A", # Light Salmon
  "Safe Sanitation" = "#F08080", # Light Coral
  "PM2.5" = "#D3D3D3", # Light Gray
  "Greenhouse" = "#B0C4DE" # Light Steel Blue
)

# Create summary per year
summary_table <- asia_africa %>%
  group_by(year) %>%  
  summarise(
    Min_TBI = min(tbi, na.rm = TRUE),
    Mean_TBI = mean(tbi, na.rm = TRUE),
    Max_TBI = max(tbi, na.rm = TRUE),
    SD_TBI = sd(tbi, na.rm = TRUE),

    Min_GDP = min(gdp, na.rm = TRUE),
    Mean_GDP = mean(gdp, na.rm = TRUE),
    Max_GDP = max(gdp, na.rm = TRUE),
    SD_GDP = sd(gdp, na.rm = TRUE),

    Min_HDI = min(hdi, na.rm = TRUE),
    Mean_HDI = mean(hdi, na.rm = TRUE),
    Max_HDI = max(hdi, na.rm = TRUE),
    SD_HDI = sd(hdi, na.rm = TRUE),

    Min_CleanFuel = min(pro_cleanfuel, na.rm = TRUE),
    Mean_CleanFuel = mean(pro_cleanfuel, na.rm = TRUE),
    Max_CleanFuel = max(pro_cleanfuel, na.rm = TRUE),
    SD_CleanFuel = sd(pro_cleanfuel, na.rm = TRUE),

    Min_SafeWater = min(pro_safewater, na.rm = TRUE),
    Mean_SafeWater = mean(pro_safewater, na.rm = TRUE),
    Max_SafeWater = max(pro_safewater, na.rm = TRUE),
    SD_SafeWater = sd(pro_safewater, na.rm = TRUE),

    Min_BasicSanitation = min(pro_basicsanitation, na.rm = TRUE),
    Mean_BasicSanitation = mean(pro_basicsanitation, na.rm = TRUE),
    Max_BasicSanitation = max(pro_basicsanitation, na.rm = TRUE),
    SD_BasicSanitation = sd(pro_basicsanitation, na.rm = TRUE),

    Min_SafeSanitation = min(pro_safesanitation, na.rm = TRUE),
    Mean_SafeSanitation = mean(pro_safesanitation, na.rm = TRUE),
    Max_SafeSanitation = max(pro_safesanitation, na.rm = TRUE),
    SD_SafeSanitation = sd(pro_safesanitation, na.rm = TRUE),

    Min_PM2.5 = min(mean_exposure_pm2.5, na.rm = TRUE),
    Mean_PM2.5 = mean(mean_exposure_pm2.5, na.rm = TRUE),
    Max_PM2.5 = max(mean_exposure_pm2.5, na.rm = TRUE),
    SD_PM2.5 = sd(mean_exposure_pm2.5, na.rm = TRUE),

    Min_Greenhouse = min(total_greenhouses, na.rm = TRUE),
    Mean_Greenhouse = mean(total_greenhouses, na.rm = TRUE),
    Max_Greenhouse = max(total_greenhouses, na.rm = TRUE),
    SD_Greenhouse = sd(total_greenhouses, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -year, names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = year, values_from = Value) %>% # Convert to wide format
  rowwise() %>% 
  mutate(Accumulated = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~ sprintf("%.2f", .x))) %>% # Format numbers to 2 decimal places
  relocate(Accumulated, .after = Statistic) %>% # Move Accumulated after Statistic
  mutate(Row_Index = row_number(), # Create row indices for coloring
         Category = case_when(
           grepl("TBI", Statistic) ~ "TBI",
           grepl("GDP", Statistic) ~ "GDP",
           grepl("HDI", Statistic) ~ "HDI",
           grepl("CleanFuel", Statistic) ~ "Clean Fuel",
           grepl("SafeWater", Statistic) ~ "Safe Water",
           grepl("BasicSanitation", Statistic) ~ "Basic Sanitation",
           grepl("SafeSanitation", Statistic) ~ "Safe Sanitation",
           grepl("PM2.5", Statistic) ~ "PM2.5",
           grepl("Greenhouse", Statistic) ~ "Greenhouse"
         ))

# Generate table with colors
summary_table_display <- summary_table %>%
  select(-Category, -Row_Index) %>% # Remove Category before displaying
  kable(caption = "Summary Statistics by Year with Accumulated Column") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Apply row coloring dynamically
for (cat in names(row_colors)) {
  rows_to_color <- which(summary_table$Category == cat) # Get row indices
  summary_table_display <- summary_table_display %>%
    row_spec(rows_to_color, background = row_colors[cat])
}

saveRDS(summary_table_display, file = here("code", "eda-code", "tables", "summary_table.html"))
print(summary_table_display)

```

```{r}
#| label: tbl-resulttable1
#| tbl-cap: "Summary Statistics Table"
#| echo: FALSE

htmltools::includeHTML(here("code", "eda-code", "tables", "summary_table.html"))

```

## Data Analysis

***Test the linearity***

```{r}
gg1 <- ggplot(asia_africa, aes(x = gdp, y = tbi)) +
  geom_point() +  # Create scatter plot
  ggtitle("Scatter Plot: tbi vs GDP") +  # Add title
  xlab("GDP") +  # Label the x-axis
  ylab("TBI") +  # Label the y-axis
  theme_minimal()

gg2 <- 
```

```{r}

# Create a list of scatter plots for each independent variable against tbi
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
    geom_point() + 
    ggtitle(paste("tbi vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

# Combine all plots into a single frame
combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want
combined_plot

combineplot_location <- here( "code","eda-code", "figures", "scatterplot.png") # to set up location for the pictures created 
ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(combined_plot)

```

 Spearman's Rank Correlation (Recommended) Spearman correlation is a non-parametric method that measures monotonic relationships (whether the variables move together, not necessarily linearly).

```{r}

# Compute Spearman correlation matrix
cor_matrix_spearman <- cor(asia_africa[, c("tbi","gdp", "hdi", "pro_cleanfuel", "pro_safewater", 
                                           "pro_basicsanitation", "pro_safesanitation", "mean_exposure_pm2.5", "total_greenhouses")], method = "spearman", use = "complete.obs")

# Melt the matrix into long format for ggplot

cor_melted <- melt(cor_matrix_spearman)

# Plot heatmap
plot5 <- ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  geom_text(aes(label = round(value, 2)), size = 4) +  # Show correlation values
  theme_minimal() +
  labs(title = "Spearman Correlation Heatmap", x = "", y = "")
matrix_location <- here( "code","eda-code", "figures", "correlation_matrix.png") # to set up location for the pictures created 
ggsave(filename = matrix_location, plot=plot5, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(plot5)
```

```{r}

#| label: fig-Spearman's Correlation Matrix 
#| fig-cap: "Spearman's Correlation Matrix "
#| echo: FALSE
knitr::include_graphics(here("code", "eda-code", "figures", "correlation_matrix.png"))
```

I will performed longitudinal data analysis

```{r}
library(car)
#vif(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
       pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, data = pdata)) # The error might be occure due to multicolinearity

#alias(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
         pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, data = asia_africa))

stepwise_model <- stepAIC(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
                             pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, 
                             data = asia_africa), direction = "both") # drop variable with multicolinearity

summary(stepwise_model)

```

***Performing longitudinal analysis***

Test for mixed or fixed effect using Hausman Test

```{r}
pdata <- pdata.frame(asia_africa, index = c("country", "year"))

fe_model <- plm(tbi ~ pro_basicsanitation + mean_exposure_pm2.5, 
                data = pdata, model = "within")
re_model <- plm(tbi ~ pro_basicsanitation + mean_exposure_pm2.5, 
                data = pdata, model = "random")

hausman_test <- phtest(fe_model, re_model)
print(hausman_test)

```

Based on the Hausman Test, fixed effect is not consistent. Therefore, I will use mixed effect test.

```{r}
summary(re_model)
```

Interpretation: ***1. variance components Analyisis***

88% dari total variasi berasal dari perbedaan antar negara, menunjukkan bahwa perbedaan karakteristik tiap negara sangat berpengaruh terhadap insiden TBC.

12% berasal dari variasi idiosinkratik, atau faktor yang berubah dari waktu ke waktu dalam suatu negara.

***2. Theta***

Rata-rata theta = 0.8868, menunjukkan bahwa sebagian besar variasi dalam data disebabkan oleh perbedaan antar negara. Theta yang mendekati 1 berarti bahwa individu (negara) memiliki efek spesifik yang kuat dibandingkan variasi waktu.

### 2025-02-23 code/processing-code/processingfile.qmd
Working on Step two of Final Project

---
title: "An example cleaning script"
author: "Muhammad Nasir"
date: "02/21/2025"
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

# Processing script

This file contains processing data for the Final Project titlled " The effects of multi environmental exposures on Tuberculosis Incidence in Asian and African countris".

***Variables in the study*** - Tuberculosis Incidence (Response Variable)

Independent Variables - Basic water - Safety Water - Greenhouse - PM2.5 - Basic Sanitation - Safe Sanitation - Slum - Fule_cooking

Control variables: - Human Development Index - Health Expenditure - GDP

# Setup

Load needed packages.
make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(purrr)
install.packages("ggplot2")
library(ggplot2)

```

# Data loading

In this project, I combined many datasets from many different sources.
I will load them one by one, then combine the dataset based on country name.

***Tuberculosis Incidence Dataset*** Tuberculosis Insidence dataset contains Estimated incidence of all forms of tuberculosis (per 100.000 population), it was obtained from World Bank Website.
https://data.worldbank.org/indicator/SH.TBS.INCD

```{r}
# note the use of the here() package and not absolute paths
tbi_location <- here::here("data","raw-data","tbi.csv")
tbi <- read.csv(tbi_location)

colnames(tbi)
```

```{r}
tbi <- tbi %>% 
  rename(country = Entity, year = Year, tbi=Estimated.incidence.of.all.forms.of.tuberculosis) %>%
  select(-Code)

saveRDS(tbi, here("data", "processed-data", "processed-tbi.rds"))
```

***Basic Water Dataset*** Basic Water dataset contains percentatge of People using at least basic drinking water services (% of population).
The data is from WHO/UNICEF Joint Monitoring Programme ( JMP ) for Water Supply, Sanitation and Hygiene ( washdata.org ).
https://databank.worldbank.org/reports.aspx?s
ource=2&series=SH.H2O.BASW.ZS&country=

```{r}
waterbasic_location <- here::here("data","raw-data","basicwater.xlsx")
waterbasic_raw <- read_excel (waterbasic_location)

summary(waterbasic_raw)
skimr::skim(waterbasic_raw)
```

Creating long table to make the table shape consitent to allow us to combine tables in the next steps

```{r}
waterbasic_raw <- waterbasic_raw %>% 
  select(-`Country Code`, - `1990 [YR1990]`, -`Series Name`, -`Series Code`)# delete potention confusing columns

colnames(waterbasic_raw) <- gsub("\\[.*\\]", "", colnames(waterbasic_raw)) # to delete [YR.... ] in the columns

# Reshape from wide to long format
waterbasic <- waterbasic_raw %>%
  pivot_longer(cols = -`Country Name`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_basicwater") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023) %>% # Remove rows where year is 2023
  rename(country = `Country Name`)

saveRDS(waterbasic, here("data", "processed-data", "processed-waterbasic.rds"))
```

***safely manage drinking water*** Fafely managed drinking water dataset is the percentage of People using safely managed drinking water services (% of population).
it is originally from WHO/UNICEF Joint Monitoring Programme ( JMP ) for Water Supply, Sanitation and Hygiene ( washdata.org ).
the data downloaded in World bank Dataset https://databank.worldbank.org/reports.aspx?s
ource=2&series=SH.H2O.SMDW.ZS&country=#

```{r}
safewater_location <- here::here("data","raw-data","safewater.csv")
safewater <- read.csv (safewater_location)

summary(safewater)
skimr::skim(safewater)
head(safewater )
colnames(safewater)
```

Creating long table to make the table shape consitent to allow us to combine tables in the next steps

```{r}
safewater <- safewater %>% 
  select(-`Series.Code`,, -`Series.Name`, -`Country.Code`)# delete potention confusing columns

colnames(safewater) <-gsub("X(\\d+)..YR\\d+.", "\\1", colnames(safewater))  # to delete [YR.... ] in the columns 
safewater <- safewater %>%
  rename(country = Country.Name)

# Reshape from wide to long format
 safewater <- safewater %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_safewater") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

 saveRDS(safewater, here("data", "processed-data", "processed-safewater.rds"))
```

***Fuel for Cooking***

```{r}
cookingfuel_location <- here::here("data","raw-data","fule_cooking.xlsx")
cookingfuel <- read_excel (cookingfuel_location )

summary(cookingfuel)
skimr::skim(cookingfuel)
colnames(cookingfuel)
```

Creating long table to make the table shape consitent to allow us to combine tables in the next steps

```{r}
cookingfuel <- cookingfuel %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(cookingfuel) <- gsub("\\[.*\\]", "", colnames(cookingfuel))  # to delete [YR.... ] in the columns

cookingfuel <- cookingfuel %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 cookingfuel <- cookingfuel %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_cleanfuel") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

 saveRDS(cookingfuel, here("data", "processed-data", "processed-cookingfuel.rds"))
```

***Basic Sanitation***

```{r}
basicsanitation_location <- here::here("data","raw-data","basicsanitation.csv")
basicsanitation <- read.csv (basicsanitation_location)

summary(basicsanitation)
skimr::skim(basicsanitation)
```

```{r}
basicsanitation <- basicsanitation %>% 
  select(-`Series.Code`, , -`Series.Name`, -`Country.Code`)# delete potention confusing columns

colnames(basicsanitation) <-gsub("X(\\d+)..YR\\d+.", "\\1", colnames(basicsanitation))  # to delete [YR.... ] in the columns 
basicsanitation <- basicsanitation %>%
  rename(country = Country.Name)

# Reshape from wide to long format
 basicsanitation <- basicsanitation %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_basicsanitation") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(basicsanitation, here("data", "processed-data", "processed-basicsanitation.rds"))
```

***Safe Sanitation***

```{r}
safesanitation_location <- here::here("data","raw-data","safesanitation.csv")
safesanitation <- read.csv (basicsanitation_location)

summary(safesanitation)
skimr::skim(safesanitation)
```

```{r}
safesanitation <- safesanitation %>% 
  select(-`Series.Code`, , -`Series.Name`, -`Country.Code`)# delete potention confusing columns

colnames(safesanitation) <-gsub("X(\\d+)..YR\\d+.", "\\1", colnames(safesanitation))  # to delete [YR.... ] in the columns 
safesanitation <- safesanitation %>%
  rename(country = Country.Name)

# Reshape from wide to long format
 safesanitation <- safesanitation %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_safesanitation") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(safesanitation, here("data", "processed-data", "processed-safesanitation.rds"))
```

***Slum Area***

```{r}
slum_location <- here::here("data","raw-data","slum.xlsx")
slum <- read_excel (slum_location)

summary(slum)
skimr::skim(slum)
```

```{r}
slum <- slum %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(slum) <- gsub("\\[.*\\]", "", colnames(slum))  # to delete [YR.... ] in the columns

slum <- slum %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 slum <- slum %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_slum") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(slum, here("data", "processed-data", "processed-slum.rds"))
```

***Greenhouses***

```{r}
greenhouses_location <- here::here("data","raw-data","greenhouse.xlsx")
greenhouses <- read_excel (greenhouses_location)

summary(greenhouses)
skimr::skim(greenhouses)
```

```{r}
greenhouses <- greenhouses %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(greenhouses) <- gsub("\\[.*\\]", "", colnames(greenhouses))  # to delete [YR.... ] in the columns

greenhouses <- greenhouses %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 greenhouses <- greenhouses %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "total_greenhouses") %>% # Total greenhouse gas in million metric cubic ton including emissions and removals from Land Use, Land-Use Change, and Forestry (LULUCF)
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(greenhouses, here("data", "processed-data", "processed-greenhouses.rds"))
```

***PM2.5***

```{r}
pm25_location <- here::here("data","raw-data","pm25.xlsx")
pm25 <- read_excel (pm25_location)

summary(pm25)
skimr::skim(pm25)
```

```{r}
pm25 <- pm25 %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(pm25) <- gsub("\\[.*\\]", "", colnames(pm25))  # to delete [YR.... ] in the columns

pm25 <- pm25 %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 pm25 <- pm25 %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "mean_exposure_pm2.5") %>% #PM2.5 air pollution, mean annual exposure (micrograms per cubic meter)
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

 saveRDS(pm25, here("data", "processed-data", "processed-pm25.rds"))
```

***GDP***

```{r}
gdp_location <- here::here("data","raw-data","gdp.csv")
gdp <- read.csv (gdp_location)

summary(gdp)
skimr::skim(gdp)
```

```{r}
gdp <- gdp %>% 
  select(-`Series.Code`, , -`Series.Name`, -`Country.Code`)# delete potention confusing columns

colnames(gdp) <-gsub("X(\\d+)..YR\\d+.", "\\1", colnames(gdp))  # to delete [YR.... ] in the columns 
gdp <- gdp %>%
  rename(country = Country.Name)

# Reshape from wide to long format
 gdp <- gdp %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "gdp") %>% #US Dollar 
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(gdp, here("data", "processed-data", "processed-gdp.rds"))
```

***Human Development Index***

```{r}
hdi_location <- here::here("data","raw-data","hdi.csv")
hdi <- read.csv (hdi_location)

summary(hdi)
skimr::skim(hdi)

```

```{r}
hdi <- hdi %>%
  rename(country = Entity, year= Year, hdi= Human.Development.Index) %>%
  select(-Code) %>%
  filter(year >= 2000 & year <= 2020)

saveRDS(hdi, here("data", "processed-data", "processed-hdi.rds"))
```

***Combine dataset***

```{r}
#create list of dataset
datasets <- list(tbi, cookingfuel, waterbasic, safewater, slum, pm25, greenhouses, basicsanitation, safesanitation, gdp, hdi)

# Merge all datasets based on "country" and "year"
combined_df <- reduce(datasets, full_join, by = c("country", "year"))
```

***create region based on WHO Region***

```{r}

# Define regions
africa_countries <- c("Algeria", "Angola", "Benin", "Botswana", "Burkina Faso", "Burundi", "Cameroon", 
                      "Cape Verde", "Central African Republic", "Chad", "Comoros", "Ivory Coast", 
                      "Democratic Republic of the Congo", "Equatorial Guinea", "Eritrea", "Ethiopia", 
                      "Gabon", "Gambia", "Ghana", "Guinea", "Guinea-Bissau", "Kenya", "Lesotho", 
                      "Liberia", "Madagascar", "Malawi", "Mali", "Mauritania", "Mauritius", "Mozambique", 
                      "Namibia", "Niger", "Nigeria", "Republic of the Congo", "Rwanda", 
                      "São Tomé and Príncipe", "Senegal", "Seychelles", "Sierra Leone", "South Africa", 
                      "South Sudan", "Eswatini", "Togo", "Uganda", "Tanzania", "Zambia", "Zimbabwe")

sear_countries <- c("Bangladesh", "Bhutan", "Democratic People's Republic of Korea", "India", "Indonesia", 
                    "Maldives", "Myanmar", "Nepal", "Sri Lanka", "Thailand", "Timor-Leste")

eur_countries <- c("Albania", "Andorra", "Armenia", "Austria", "Azerbaijan", "Belarus", "Belgium", 
                   "Bosnia and Herzegovina", "Bulgaria", "Croatia", "Cyprus", "Czech Republic", "Denmark", 
                   "Estonia", "Finland", "France", "Georgia", "Germany", "Greece", "Hungary", "Iceland", 
                   "Ireland", "Israel", "Italy", "Kazakhstan", "Kyrgyzstan", "Latvia", "Lithuania", 
                   "Luxembourg", "Malta", "Moldova", "Monaco", "Montenegro", "Netherlands", "North Macedonia", 
                   "Norway", "Poland", "Portugal", "Romania", "Russia", "San Marino", "Serbia", "Slovakia", 
                   "Slovenia", "Spain", "Sweden", "Switzerland", "Tajikistan", "Turkey", "Turkmenistan", 
                   "Ukraine", "United Kingdom", "Uzbekistan")

emr_countries <- c("Afghanistan", "Bahrain", "Djibouti", "Egypt", "Iran", "Iraq", "Jordan", "Kuwait", 
                   "Lebanon", "Libya", "Morocco", "Oman", "Pakistan", "Qatar", "Saudi Arabia", "Somalia", 
                   "Sudan", "Syria", "Tunisia", "United Arab Emirates", "Yemen")

wpr_countries <- c("Australia", "Brunei", "Cambodia", "China", "Cook Islands", "Fiji", "Japan", "Kiribati", 
                   "Laos", "Malaysia", "Marshall Islands", "Micronesia", "Mongolia", "Nauru", "New Zealand", 
                   "Niue", "Palau", "Papua New Guinea", "Philippines", "Samoa", "Singapore", "Solomon Islands", 
                   "South Korea", "Taiwan", "Tonga", "Tuvalu", "Vanuatu", "Vietnam")

amr_countries <- c("Antigua and Barbuda", "Argentina", "Bahamas", "Barbados", "Belize", "Bolivia", "Brazil", 
                   "Canada", "Chile", "Colombia", "Costa Rica", "Cuba", "Dominica", "Dominican Republic", 
                   "Ecuador", "El Salvador", "Grenada", "Guatemala", "Guyana", "Haiti", "Honduras", "Jamaica", 
                   "Mexico", "Nicaragua", "Panama", "Paraguay", "Peru", "Saint Kitts and Nevis", "Saint Lucia", 
                   "Saint Vincent and the Grenadines", "Suriname", "Trinidad and Tobago", "Uruguay", "Venezuela")

# Add Region column
combined_df <- combined_df %>%
  mutate(region = case_when(
    country %in% africa_countries ~ "Africa",
    country %in% sear_countries ~ "South-East Asia",
    country %in% eur_countries ~ "Europe",
    country %in% emr_countries ~ "Eastern Mediterranean",
    country %in% wpr_countries ~ "Western Pacific",
    country %in% amr_countries ~ "Americas",
    TRUE ~ "Other"  # Assign "Other" to countries not in these regions
  ))

```

***Save the dataset to csv file***

```{r}

write.csv(combined_df, here("data", "processed-data", "combined-dataset.csv"), row.names = FALSE)
saveRDS(combined_df, here("data", "processed-data", "combined-dataset.rds"))
```

### Exluding slum dataset

Slum dataset is different from other dataset, the observation is every 2 years.
I want to create subset without slum data

```{r}
no_slum <- combined_df %>% 
  select(-pro_slum)

```

Now, I want to drop all NA observation

```{r}

no_slum <- no_slum %>%
  filter(if_all(everything(), ~ !is.na(.) & . != "" & . != ".."))
```

Now, I want to delete countries which have incomplete observation in all years.

```{r}
# Count the number of rows for each country
complete_countries <- no_slum %>%
  group_by(country) %>%
  summarise(row_count = n()) %>%
  filter(row_count == 21) %>%
  pull(country)  # Extract only countries with exactly 21 rows

# Keep only those countries in the dataset
clean_noslum <- no_slum %>%
  filter(country %in% complete_countries)

# Print first few rows to check
head(clean_noslum)

table(clean_noslum$year, clean_noslum$country) # to make sure that all countries have complete year 
dim(clean_noslum)

unique(clean_noslum$country)

```

```{r}
clean_noslum <- clean_noslum %>%
  mutate(across(c(pro_cleanfuel, pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, gdp), as.numeric))

summary(clean_noslum)

saveRDS(clean_noslum, here("data", "processed-data", "clean_noslum.rds")) # save the file

```

### Creating clean dataset inlcuding slum

Note: Slum dataset only contain observation every two years. In this dataset, I will follow the year where slum has observation. 
```{r}
# Define the required years
required_years_alldata <- c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020)

# Filter the dataset for the required years
all_data <- combined_df %>%
  filter(year %in% required_years_alldata) %>%
  group_by(country) %>%
  filter(all(required_years_alldata %in% year)) %>%
  ungroup()

all_data  <- all_data  %>%
  filter(year %in% required_years_alldata) %>%  # Keep only required years
  group_by(country) %>%
  filter(n_distinct(year) == length(required_years_alldata)) %>%  # Ensure all years exist for each country
  ungroup()

all_data <- all_data %>%
  mutate(across(where(is.character), ~ na_if(.x, ".."))) %>%  # Convert ".." to NA only for character columns
  drop_na()  # Remove rows with any NA values

# View the filtered data
print(all_data )

unique(all_data$country)
# Optionally, save the filtered data to a new CSV file
#write.csv(filtered_hdi, "filtered_hdi_selected_years.csv", row.names = FALSE)

```

```{r}
ggplot(all_data, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.7) +  # Transparency to see overlapping lines
  labs(title = "Spaghetti Plot of TBC Incidence",
       x = "Year",
       y = "TBC Incidence") +
  theme_minimal() +
  theme(legend.position = "none") 
```

```{r}
# convert variables into numeric

all_data <- all_data %>%
  mutate(across(c(pro_cleanfuel, pro_basicwater, pro_safewater, pro_slum, pro_basicsanitation, pro_safesanitation, mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, gdp), as.numeric))
summary(all_data)
str(all_data)

saveRDS(all_data, here("data", "processed-data", "clean_data_withslum.rds")) # save the file 
write.csv(all_data, here("data", "processed-data", "clean_data_withslum.csv"), row.names = FALSE)
unique(all_data$country)
```

```{r}
all_asiaafrica <- all_data %>% 
  filter(region %in% c("South-East Asia", "Africa"))

dim(all_asiaafrica)

unique(all_asiaafrica$country)

m2 <- lm(log(tbi) ~ pro_slum, data = all_asiaafrica)
summary(m2)

plot(m2)
```

```{r}
library(reshape2)  # To reshape data for ggplot

# Compute Spearman correlation matrix
cor_matrix_spearman_slum <- cor(all_asiaafrica[, c("tbi","gdp", "hdi", "pro_cleanfuel", "pro_safewater", 
                                           "pro_basicsanitation", "pro_safesanitation", "pro_slum")], 
                           method = "spearman", use = "complete.obs")

# Melt the matrix into long format for ggplot
cor_melted_slum <- melt(cor_matrix_spearman_slum)

# Plot heatmap
ggplot(cor_melted_slum, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  geom_text(aes(label = round(value, 2)), size = 4) +  # Show correlation values
  theme_minimal() +
  labs(title = "Spearman Correlation Heatmap", x = "", y = "")
```

```{r}
ggplot(all_asiaafrica, aes(x = year, y = pro_slum, group = country)) +
  geom_line() +
  facet_wrap(~ country) +  # Creates separate plots for each country
  theme_minimal()
```

### 2025-02-23 products/manuscript/Manuscript.qmd
Working on Step two of Final Project

@@
 ---
-title: "The effects of multi environmental exposure on Tuberculosis Incidence in Asian and African countris"
+title: "The effects of multi environmental exposures on Tuberculosis Incidence in South East Asian and African countris"
 author: ""
@@
 ---
-
-The structure below is one possible setup for a manuscript, or a general
-data analysis project (including the course project). Adjust as needed.
-You don't need to have exactly these sections, but the content covering
-those sections should be addressed.
 
@@
 
-*Describe your methods. That should describe the data, the cleaning
-processes, and the analysis approaches. You might want to provide a
-shorter description here and all the details in the supplement.*
+The datasets were loded one by one, then drop all unused variable before
+comined those datasets using reduce() function. Countries, which have at
+least one missing years from 2000 to 2020, were dropped from the study.
+To make sure that the data is clean and ready to analyse, all missing
+values were dropped. This study focuses on South East Asia and Africa
+Regions, where the major proportion of tuberculosis cases came.
+Therefore, a newsubset for South East Asia and African countries was
+created.
 
-I haven't decide about data analysis methods. However, I am thinking to
-perform time series analysis and multi linear regression (if suitable)
-or Gene-realized linear model (we will see after conducting data
-exploration). I am still in the process of reading more papers to get
-more information which methods best suit this data and answer the
-research questions.
+After data cleaning, multiple exploratory data analysis was forformed
+such us summary and vizualising the data. A bloxplot was created to show
+Tuberculosis Incidence in each country during the time frame. Spagheti
+plot was created to visualize the trend of tuberculosis incidence and
+other variables over the time. Lastly, line charts showing the trend of
+all variables in each country.
+
+The exploratory data analysis showed that the data were not normally
+distributed and there was no linear relationship between tuberculosis
+incidence and other predictors. Therefore, Spearman's correlation matrix
+was perfomed to see the correlations. To have better understanding, a
+heatmap matrix plot was created to show the correlation.
+
+The main statistical analysis in thie study is longitudinal data
+analysis (eithre fixed effect or mixed effect method). To see which one
+is better, Hausman Test was performed to determine which model is
+better.
+
+NOTE: I still need to discuss and need more advice about the data
+analysis from Dr. Handel
 
@@
 
-Sometimes you might want to show a schematic diagram/figure that was not
-created with code (if you can do it with code, do it). @fig-schematic is
-an example of some - completely random/unrelated - schematic that was
-generated with Biorender. We store those figures in the `assets` folder.
+NOTE: I AM STILL WORKING TO FIGUR THIS ONE OUT Sometimes you might want
+to show a schematic diagram/figure that was not created with code (if
+you can do it with code, do it). @fig-schematic is an example of some -
+completely random/unrelated - schematic that was generated with
+Biorender. We store those figures in the `assets` folder.
 
@@
 
-*Write code that reads in the file and cleans it so it's ready for
-analysis. Since this will be fairly long code for most dataset, it might
-be a good idea to have it in one or several R scripts. If that is the
-case, explain here briefly what kind of cleaning/processing you do, and
-provide more details and well documented code somewhere (e.g. as
-supplement in a paper). All materials, including files that contain
-code, should be commented well so everyone can follow along.*
+Data Import and data cleaning process have been explained in the methode
+part. To see the code for data import and data cleaning , please visit
+working directory ("code", "processing-code", "processingfile.qdm").
+Processed and clean data are saved at a folder ("data",
+"processed-data"). There are many .rds files in the folder icnlduing
+file for each variables, and file after combined all variables.
 
@@
 -   to combine them base on country name/ code and years.
-
-<!-- -->
 
@@
 
-*Use a combination of text/tables/figures to explore and describe your
-data. Show the most important descriptive results here. Additional ones
-should go in the supplement. Even more can be in the R and Quarto files
-that are part of your project.*
+There are some exploratory data analysis conducted and resulted table,
+and charts. Table 1 shows the summary statistics including min, mean,
+max, and standard deviation of all variables.
 
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the
-`../../` notation. (Two dots means a folder up). You never want to
-specify an **absolute** path like `C:\ahandel\myproject\results\`
-because if you share this with someone, it won't work for them since
-they don't have that path. You can also use the `here` R package to
-create paths. See examples of that below. I generally recommend the
-`here` package.
+NOTE: I still have issue in saving the stiled tables. I am still working
+on it. I will show it next step.
 
@@
 
+The line graph ( @fig-1 ) shows the trend of all variables over the
+period in each country. It can be seen that the majority of country
+showed the decreases trend in tuberculosis incidence over 21 years
+period, except Lesotho showing a sharp increase until the middle of the
+period before dramatic decrease until the end of the period. All
+countries showed the increase trend in percentage of population using
+using basic and safe drinking water, basic and safe sanitation, and
+clean fuel and technology for cooking, except Zimbabuwe showing the
+decrease in percentage of population using basic drinking water and safe
+sanitation, and Senegal showing the decrease of population using clean
+fuel and technology for cooking.
+
+```{r}
+
+#| label: fig-1
+#| fig-cap: "Trends of all variable over the period in each country"
+#| echo: FALSE
+knitr::include_graphics(here("code","eda-code","figures","line-graph-tbivsfactors.png"))
+
+```
+
+( @fig-2 ) shows spaghetti plots in each variables. We can see the
+trends of each country in every variables. It can be seen that generally
+most of the countries show the increase trends in water and sanitation,
+clean fuel use, human development index, total greenhouses, but
+fluctuated trends in mean of PM2.5 expousures.
+
+```{r}
+
+#| label: fig-2
+#| fig-cap: "Spaghetti plots in each variables"
+#| echo: FALSE
+knitr::include_graphics(here("code","eda-code","figures","spageti.png"))
+
+```
+
+Now lets take a more detail tuberculosis incidence in each country over
+the year using the following box plot. Lesotho leads the incidence
+followed by Myanmar and Tanzania.
+
+```{r}
+
+#| label: fig-3
+#| fig-cap: "Spaghetti plots in each variables"
+#| echo: FALSE
+knitr::include_graphics(here("code","eda-code","figures","boxplot.png"))
+
+```
+
 ## Basic statistical analysis
+
+As first step of statistical analysis.
+Using scatterplot to show the linearity between tuberculosis incidence and predictors.
+( @fig-4 ) shows the scatterplots.
+The scatterplots do not show linear relationship between tuberculosis Incidence and predictors.
+
+```{r}
+
+#| label: fig-4
+#| fig-cap: "Scatter plots between Tuberculosis Incidence and Predictors "
+#| echo: FALSE
+knitr::include_graphics(here("code","eda-code","figures","scatterplot.png"))
+
+```
+Before going further, we need to look at correlation between tuberculosis Incidence and all predictors.
+
+Since they show nonlinearity, spearman's correlation performed and hsowing in the following correlation matrix. See ( @fig-5 )
+
+```{r}
+
+#| label: fig-5
+#| fig-cap: "Spearman Correlation Matrix "
+#| echo: FALSE
+knitr::include_graphics(here("code","eda-code","figures","correlation_matrix.png"))
+
+```
+
+Since there is no linear relationship, nonlinear relationship test was performed.
+
+```{r}
+
+```
 

### 2025-02-23 code/analysis-code/statistical-analysis.qdm.qmd
Add some important things

@@
 
+library(splines)
 library(here)
@@
 Load the dataset
+
 ```{r}
@@
 asia_africa_location <- here("data", "processed-data", "asia_africa.rds")
-readRDS(asia_africa_location)
+asia_africa <- readRDS(asia_africa_location)
 
@@
 ```{r}
+
+# Define independent variables (exclude 'tbi' and categorical variables)
+independent_vars <- colnames(asia_africa)[!colnames(asia_africa) %in% c("tbi", "country", "region", "year")]
+
+# Create scatter plots for each independent variable against tbi
+plots <- lapply(independent_vars, function(var) {
+  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
+    geom_point() + 
+    ggtitle(paste("TBI vs", var)) + 
+    xlab(var) + 
+    ylab("TBI") + 
+    theme_minimal()
+})
 
@@
 
+Interpretation: The predictors have no linear relationship with Tuberculosis Incidence. Therefore, we would like to perform nonlinear relationship to decide further action and analysis.
+
+***Check Nonlinear relationship***
+
+```{r}
+
+# Reshape the dataset to long format
+asia_africa_long <- asia_africa %>%
+  pivot_longer(cols = c(pro_cleanfuel, pro_basicwater, pro_safewater, 
+                        mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, 
+                        pro_safesanitation, gdp, hdi),
+               names_to = "Predictor", values_to = "Value")
+
+# Plot nonlinear relationships using LOESS smoothing
+nonlinear <- ggplot(asia_africa_long, aes(x = Value, y = tbi)) +
+  geom_point(alpha = 0.5) +  # Scatter points
+  geom_smooth(method = "loess", color = "red") +  # LOESS curve
+  facet_wrap(~ Predictor, scales = "free_x") +  # Separate plots for each predictor
+  theme_minimal() +
+  labs(title = "Nonlinear Relationship between TBI and Predictors",
+       x = "Predictor Value",
+       y = "TBI")
+
+nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
+ggsave(filename = nonrelationshiplot_location, plot=nonlinear, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+plot(nonlinear)
+```
+
+Data transformation
+
+```{r}
+
+library(dplyr)
+
+asia_africa <- asia_africa %>%
+  mutate(
+    log_tbi = log(tbi + 1),
+    log_gdp = log(gdp + 1),
+    log_total_greenhouses = log(total_greenhouses + 1),
+    log_pm2.5 = log(mean_exposure_pm2.5 + 1),
+    sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
+    sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
+    poly_cleanfuel = as.numeric(poly(pro_cleanfuel, 2, raw = TRUE)[, 1]),
+    poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
+    poly_safesanitation = as.numeric(poly(pro_safesanitation, 2, raw = TRUE)[, 1])
+  )
+
+```
+
+```{r}
+asia_africa_transformed <- asia_africa %>%
+  dplyr::select(log_tbi, log_gdp, log_total_greenhouses, log_pm2.5, 
+                poly_cleanfuel, poly_basicsanitation, poly_safesanitation,
+                sqrt_basicwater, sqrt_safewater, hdi) %>%
+  pivot_longer(cols = -log_tbi, names_to = "Predictor", values_to = "Value")
+
+colnames(asia_africa)
+
+# Plot after transformation
+transformed <- ggplot(asia_africa_transformed, aes(x = Value, y = log_tbi)) +
+  geom_point(alpha = 0.3) +
+  geom_smooth(method = "lm", color = "red") +  # Use linear model for comparison
+  facet_wrap(~ Predictor, scales = "free_x") +
+  theme_minimal() +
+  labs(title = "Transformed Predictor Relationships with Log(TBI)",
+       x = "Transformed Predictor Value", y = "Log(TBI)")
+
+transformedplot_location <- here( "code","analysis-code", "figures", "transformed.png") # to set up location for the pictures created 
+ggsave(filename = transformedplot_location, plot=transformed, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+plot(transformed)
+```
+
+After transforming variables. They look more linear. Now GLM will be performed
+
+***GLMs***
+
+```{r}
+glm_model <- glm(log_tbi ~ log_gdp + log_total_greenhouses + log_pm2.5 +
+                      poly_cleanfuel + poly_basicsanitation + poly_safesanitation +
+                      sqrt_basicwater + sqrt_safewater + hdi,
+                 data = asia_africa, family = gaussian)
+summary(glm_model)
+
+```
+
+\*\*\*General Addictive Model (GAM)
+
+```{r}
+library(mgcv)
+
+gam_model <- gam(log_tbi ~ s(log_gdp) + s(log_total_greenhouses) + s(log_pm2.5) +
+                      s(poly_cleanfuel) + s(poly_basicsanitation) + s(poly_safesanitation) +
+                      s(sqrt_basicwater) + s(sqrt_safewater) + s(hdi),
+                 data = asia_africa, family = gaussian)
+summary(gam_model)
+
+```
+
+Interpretation:
+
+Generalized Additive Model (GAM) explains 94.5% of the deviance in log-transformed tuberculosis incidence (Log(TBI)), which suggests an excellent fit.
+Below is a step-by-step breakdown of the interpretation.
+
+Overall Model Performance Adjusted R² (R-sq.(adj) = 0.917) - This means 91.7% of the variability in Log(TBI) is explained by the predictors.
+- A very strong model since it explains most of the variance.
+
+Deviance Explained = 94.5%
+
+-   Similar to R² but for GAM, indicating excellent model fit.
+
+Generalized Cross-Validation (GCV) = 0.0685
+
+-   Lower GCV indicates better predictive performance.
+-   This suggests that the model generalizes well to new data.
+
+![](images/clipboard-4125844072.png)
+
+Now, I want to remove insignificant variables
+
+```{r}
+gam_optimized <- gam(log_tbi ~ s(log_gdp) + s(log_total_greenhouses) + s(log_pm2.5) + 
+                         s(poly_cleanfuel) + s(sqrt_basicwater) + 
+                         s(sqrt_safewater) + s(hdi),
+                     data = asia_africa, family = gaussian)
+summary(gam_optimized)
+
+```
+
+Comparing GAM models
+
+```{r}
+AIC(gam_model, gam_optimized)
+BIC(gam_model, gam_optimized)
+
+```
+
+Visualize significant predictors
+
+```{r}
+plot(gam_model, pages = 1, shade = TRUE)
+
+```
+
+```{r}
+ggplot(data = asia_africa, aes(x = exp(log_tbi), y = exp(predict(gam_model, type = "response")))) +
+  geom_point(alpha = 0.5) +
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  labs(title = "Actual vs. Predicted TBI",
+       x = "Actual TBI", y = "Predicted TBI")
+
+```
+
+***Comparing the model***
+
+```{r}
+AIC(glm_model, gam_model)
+BIC(glm_model, gam_model)
+```
+
+After comapring the models, based on AIC and BIC, GAM model is more appropriate
+
+Check Paryial effects of Predictors
+
+```{r}
+plot(gam_model, pages = 1, shade = TRUE)
+
+```
+
+```{r}
+predictions <- predict(gam_model, newdata = asia_africa, type = "response")
+head(predictions)
+
+```
+
+Compare actual vs predicted values
+
+```{r}
+ggplot(data = asia_africa, aes(x = log_tbi, y = predictions)) +
+  geom_point(alpha = 0.5) +
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  labs(title = "Actual vs. Predicted Log(TBI)", x = "Actual Log(TBI)", y = "Predicted Log(TBI)")
+
+asia_africa$predicted_tbi <- exp(predict(gam_model, newdata = asia_africa, type = "response"))
+
+ggplot(asia_africa, aes(x = tbi, y = predicted_tbi)) +
+  geom_point(alpha = 0.5) +
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  labs(title = "Actual vs. Predicted TBI",
+       x = "Actual TBI", y = "Predicted TBI")
+
+```
+
+Interpretation:
+
+The points are generally close to the line, indicating that the model is accurate.
+
 ```{r}
@@
 
+# Ensure it's a tibble
+asia_africa <- as_tibble(asia_africa)
+
+# Select variables (excluding categorical ones)
+asia_africa_selected <- asia_africa %>%
+  select(-all_of(c("country", "region", "year")))
+
+# Reshape the dataset
+asia_africa_long <- asia_africa_selected %>%
+  pivot_longer(cols = -tbi, names_to = "predictor", values_to = "value")
+
+# Print first few rows
+print(head(asia_africa_long))
+
 # Step 2: Reshape the dataset for visualization
+asia_africa <- as_tibble(asia_africa)
 asia_africa_long <- asia_africa %>%
-  select(-country, -region, -year) %>%  # Exclude categorical variables
+  select(-c(country, region, year)) %>%  # Use `-c()` for multiple column exclusion
   pivot_longer(cols = -tbi, names_to = "predictor", values_to = "value")
@@
 ```
+
 Q-Q plot of residuals from the Poisson GLM suggests that the model is not fitting the data well.
@@
 ```
+
 since the result is very high, which should be close to 1. We need a model that can hanlde overdispersion
@@
 
-1. Overdispersion Check
-The estimated Theta (4.246, SE = 0.430) suggests that overdispersion is present but is well-accounted for by the Negative Binomial model.
+1.
+Overdispersion Check The estimated Theta (4.246, SE = 0.430) suggests that overdispersion is present but is well-accounted for by the Negative Binomial model.
 The Residual Deviance (196.14) is much lower than the Null Deviance (409.01), meaning the model fits the data better than a null model.
@@
 
-- hdi (-7.642, p < 2e-16) Higher HDI is associated with lower tbi cases.
+-   hdi (-7.642, p \< 2e-16) Higher HDI is associated with lower tbi cases.
 -   pro_basicsanitation (0.00577, p = 0.0334) Increased access to basic sanitation is linked to a slight increase in tbi.
@@
 -   pro_cleanfuel (-0.005999, p = 0.0137) → More use of clean fuels is linked to lower tbi.
-- mean_exposure_pm2.5 (-0.02698, p < 2e-16) → Higher PM2.5 exposure reduces tbi (unexpected; may need further investigation).
+-   mean_exposure_pm2.5 (-0.02698, p \< 2e-16) → Higher PM2.5 exposure reduces tbi (unexpected; may need further investigation).
 

### 2025-02-23 products/manuscript/Manuscript.qmd
Add some important things

@@
 
+#| label: fig-6
+#| fig-cap: "Nonlinear relationship"
+#| echo: FALSE
+knitr::include_graphics(here("code","analysis-code","figures","nonlinear.png"))
+
 ```
 
-*To get some further insight into your data, if reasonable you could
-compute simple statistics (e.g. simple models with 1 predictor) to look
-for associations between your outcome(s) and each individual predictor
-variable. Though note that unless you pre-specified the outcome and main
-exposure, any "p\<0.05 means statistical significance" interpretation is
-not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+After performing nonlinear model, data transformation used based on the shave or distribution of the plots from nonlinear relationship plots ( @fig-6 ) above.
+( @fig-7 ) shows the data distribution after data transformation.
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
+
+#| label: fig-7
+#| fig-cap: "transformed data plots"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
+knitr::include_graphics(here("code","analysis-code","figures","transformed.png"))
+
 ```
@@
 
-*Use one or several suitable statistical/machine learning methods to
-analyze your data and to produce meaningful figures, tables, etc. This
-might again be code that is best placed in one or several separate R
-scripts that need to be well documented. You want the code to produce
-figures and data ready for display as tables, and save those. Then you
-load them here.*
+For the final model, Generalized linear models (GLMs) and Generalized Addictive Models (GAMs). I need to discuss with Dr. Handal to get advice about this.
+
+I actually have performed both GLMs and GAMs, but I still need to learn both models in depth.
+Specially for GAM, I am very new in the models.
+Initially I performed GLMs since the Transformed plots show linears line, but aftrer comparing the GLMs and GAMs model, GAMs perfoming better.
 
@@
 # References
+
+I will include reference for the sourse of data in the next step.

### 2025-03-19 code/analysis-code/statistical-analysis.qdm.qmd
Data Analysis (Part 3 of the project)

@@
 library(gridExtra)
+
+install.packages("lmerTest")
+library(lmerTest)
+install.packages("stargazer")
+library(stargazer)
+
 ```
@@
 
+# In the study, I will focus on Water and Sanitation and remove Lesotho as it considered as outlier
+asia_africa <- asia_africa%>%
+  filter(country != "Lesotho") %>% #Drop Lesotho from the observation
+  select("tbi", "year", "pro_basicwater", "pro_safewater", "pro_basicsanitation", "pro_safesanitation", "hdi", "country", "region")
+
 colnames(asia_africa)
@@
 ```{r}
+# Adding Fixed-Effect to Make Conditional Models ---
+ggplot(asia_africa, aes(x = year, y = tbi)) + 
+  geom_line(aes(group=country), col="grey") + 
+  stat_smooth(method="loess", col="black", lwd=1.5, se=FALSE)+
+  scale_x_continuous(name = "years") + 
+  scale_y_continuous(name = "Tuberculosis Incidence") +
+  theme_bw() + 
+  theme(axis.text=element_text(size=14, colour="black"), 
+        axis.title=element_text(size=14,face="bold")) +
+  theme(strip.text.x = element_text(size = 14))+ theme(legend.position="none") 
+```
 
-# Define independent variables (exclude 'tbi' and categorical variables)
-independent_vars <- colnames(asia_africa)[!colnames(asia_africa) %in% c("tbi", "country", "region", "year")]
+```{r}
+# Convert dataset to long format for faceting
+asia_africa_long <- asia_africa %>%
+  pivot_longer(cols = c(pro_basicwater, pro_safewater,
+                        pro_basicsanitation, pro_safesanitation, hdi),
+               names_to = "variable", values_to = "value")
 
-# Create scatter plots for each independent variable against tbi
-plots <- lapply(independent_vars, function(var) {
-  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
-    geom_point() + 
-    ggtitle(paste("TBI vs", var)) + 
-    xlab(var) + 
-    ylab("TBI") + 
-    theme_minimal()
-})
+# Create scatter plot faceted by independent variable
+scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
+  geom_point(alpha = 0.6) +  # Scatter points, colored by country
+  stat_smooth(method = "loess", col = "black", lwd = 1.2, se = FALSE) +  # LOESS (non-linear) in black
+  facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
+  labs(x = "Independent Variable Value", y = "Tuberculosis Incidence (TBI)") +
+  theme_bw() + 
+  theme(axis.text = element_text(size = 12, colour = "black"), 
+        axis.title = element_text(size = 14, face = "bold"),
+        strip.text = element_text(size = 14, face = "bold"), 
+        legend.position = "none")  # Hides the legend
 
-# Create a list of scatter plots for each independent variable against tbi
-plots <- lapply(independent_vars, function(var) {
-  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
-    geom_point() + 
-    ggtitle(paste("tbi vs", var)) + 
-    xlab(var) + 
-    ylab("TBI") + 
-    theme_minimal()
-})
+nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
+ggsave(filename = nonrelationshiplot_location, plot=scatter_faceted, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+# Show plot
+print(scatter_faceted)
 
-# Combine all plots into a single frame
-combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want
+```
 
-combineplot_location <- here( "code","analysis-code", "figures", "scatterplot.png") # to set up location for the pictures created 
-ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
-plot(combined_plot)
+Non-linearity test 
+```{r}
+# Install package if not already installed
+install.packages("lmtest")
+library(lmtest)
 
+# Fit a linear model
+lm_model <- lm(tbi ~ pro_cleanfuel + pro_basicwater + pro_safewater + 
+               mean_exposure_pm2.5 + total_greenhouses + 
+               pro_basicsanitation + pro_safesanitation + gdp + hdi, 
+               data = asia_africa)
+
+# Run the RESET test for linearity
+resettest(lm_model, power = 2:3)
+
+```
+
+```{r}
+
+# Fit a linear model (LM)
+lm_model <- lm(tbi ~ pro_cleanfuel + pro_basicwater + pro_safewater + 
+               mean_exposure_pm2.5 + total_greenhouses + 
+               pro_basicsanitation + pro_safesanitation + gdp + hdi, 
+               data = asia_africa)
+
+# Fit a non-linear model (GAM)
+gam_model <- gam(tbi ~ s(pro_cleanfuel) + s(pro_basicwater) + s(pro_safewater) + 
+                  s(mean_exposure_pm2.5) + s(total_greenhouses) + 
+                  s(pro_basicsanitation) + s(pro_safesanitation) + 
+                  s(gdp) + s(hdi), 
+                 data = asia_africa, method = "REML")
+
+# Compare AIC values
+AIC(lm_model, gam_model)
+
+# Plot residuals vs. fitted values for the linear model
+plot(lm_model$fitted.values, resid(lm_model), 
+     xlab = "Fitted Values", ylab = "Residuals", 
+     main = "Residuals vs. Fitted Plot")
+abline(h = 0, col = "red")  # Horizontal line at 0
+
+```
+
+```{r}
+gamm_model <-gam(tbi ~ s(year) + s(pro_cleanfuel) + s(hdi) + ti(pro_cleanfuel, hdi) + s(country, bs="re"), 
+    family = nb(), method = "REML", data = asia_africa)
+
+```
+
+```{r}
+library(car)
+vif(gamm_model)
+```
+
+```{r}
+cor_matrix <- cor(asia_africa[, c("year", "pro_cleanfuel", "hdi", "gdp")], use="pairwise.complete.obs")
+print(cor_matrix)
+
+```
+
+### Longitudinal analysis with Mixed-effect models
+```{r}
+# Pastikan tidak ada NA dalam variabel yang dianalisis
+asia_africa_clean <- asia_africa %>%
+  drop_na(tbi, pro_cleanfuel, mean_exposure_pm2.5, pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, total_greenhouses, gdp, hdi, year, country)
+
+# Model Mixed-Effects: TBI is predicted by predictors with random effect per each country 
+model_lmm <- lmer(tbi ~ pro_cleanfuel+ mean_exposure_pm2.5 + pro_basicwater + pro_safewater + pro_basicsanitation + pro_safesanitation + total_greenhouses + gdp + hdi + year + (1 | country),  # Random intercept dressed variation in each country
+                   data = asia_africa_clean, REML = FALSE)
+
+# Ringkasan hasil model
+summary(model_lmm)
 ```
@@
 asia_africa_long <- asia_africa %>%
-  pivot_longer(cols = c(pro_cleanfuel, pro_basicwater, pro_safewater, 
-                        mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, 
-                        pro_safesanitation, gdp, hdi),
+  pivot_longer(cols = c( pro_basicwater, pro_safewater, pro_basicsanitation, 
+                        pro_safesanitation,hdi),
                names_to = "Predictor", values_to = "Value")
@@
 
-Data transformation
+It can be seen that some variables show u-shaped (quadratic), however some variables looks more complex.
+In this case, we are going to try to fit two models, first model will fit Mixed effects model with Polynomial Terms, and addictive mixed model (GAMM).
+
+I will perform two steps here: 
+1. Data transformation, then looking at possible model, 
+2. Directly performing Mixed effects model with polynomial terms, and Performing GAMM
 
@@
     log_tbi = log(tbi + 1),
-    log_gdp = log(gdp + 1),
-    log_total_greenhouses = log(total_greenhouses + 1),
-    log_pm2.5 = log(mean_exposure_pm2.5 + 1),
     sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
     sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
-    poly_cleanfuel = as.numeric(poly(pro_cleanfuel, 2, raw = TRUE)[, 1]),
     poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
@@
 asia_africa_transformed <- asia_africa %>%
-  dplyr::select(log_tbi, log_gdp, log_total_greenhouses, log_pm2.5, 
-                poly_cleanfuel, poly_basicsanitation, poly_safesanitation,
+  dplyr::select(log_tbi, poly_basicsanitation, poly_safesanitation,
                 sqrt_basicwater, sqrt_safewater, hdi) %>%
@@
 plot(transformed)
+```
+
+*Generalized Linear mixed model (GLMM)*
+After transformation, the data show linear relationship, therefore we perform Linear Mixed Models (LMM) 
+```{r}
+
+#Generalized Linear mixed model (GLMM)
+#without HDI
+library(lme4)
+model1<- glmer(log(tbi) ~ sqrt(pro_basicwater) + sqrt(pro_safewater) + 
+                          poly(pro_basicsanitation, 2) + poly(pro_safesanitation, 2) +
+                          (1 | country), data = asia_africa)
+summary(model1)
+
+```
+
+Interpretation:
+✔ Random intercept for country is moderate, meaning tbi varies across countries.
+✔ Residual variance is small, indicating that the model explains much of the variability
+
+✔ Intercept (5.24113) is highly significant, representing the baseline log-transformed tbi when all predictors are at their reference values.
+✔ poly(pro_basicsanitation, 2)1 (first-order term) is significant, indicating that basic sanitation has a linear effect on tbi.
+❌ poly(pro_basicsanitation, 2)2 (second-order term) is not significant, suggesting that the quadratic effect is weak or unnecessary.
+❌ pro_basicwater and pro_safewater are not significant, meaning they may not be strong predictors of tbi.
+
+```{r}
+#with HDI
+library(lme4)
+model2<- glmer(log(tbi) ~ sqrt(pro_basicwater) + sqrt(pro_safewater) + 
+                          poly(pro_basicsanitation, 2) + poly(pro_safesanitation, 2) +
+                          (1 | country) + hdi, data = asia_africa)
+summary(model2)
+
+```
+Interpretation:
+
+✔ Random intercept for country is moderate, meaning tbi varies across countries.
+✔ Residual variance is very small, indicating that the model explains most of the variability, but this may suggest overfitting.
+
+✔ HDI (hdi) is highly significant (p < 0.001), meaning higher HDI is associated with lower tbi.
+✔ poly(pro_basicsanitation, 2)1 (first-degree term) is significant, indicating that basic sanitation has a non-linear effect on tbi.
+❌ poly(pro_basicsanitation, 2)2 (second-degree term) is not significant, suggesting that the quadratic effect is weak or unnecessary.
+❌ pro_basicwater and pro_safewater are not significant, meaning they may not be useful in predicting tbi.
+
+*check wulticolinearity*
+```{r}
+vif(model_lmm)
+```
+The Vif test shows that Pro_safesanitation is >5. We need to remove.
+
+*fit model without non-significant predictor*
+
+```{r}
+#refit model without non-significant predictor using lmm 
+model3 <- glmer(log(tbi) ~ hdi + poly(pro_basicsanitation, 2) + (1 | country), 
+                     data = asia_africa)
+summary(model3)
+
+```
+Interpretation:
+
+The country-level variance (0.42162) is relatively high, indicating that TB incidence varies significantly between countries.
+The residual variance (0.01748) is much smaller, meaning a significant portion of TB variation is explained by country-level differences.
+
+Key Findings
+✅ HDI is highly significant (Estimate = -3.67, p < 2e-16)
+
+Higher HDI is strongly associated with lower TB incidence.
+A 1-unit increase in HDI decreases log(TBI) by ~3.67.
+✅ Second-degree polynomial term for sanitation (pro_basicsanitation²) is highly significant (p = 4.15e-06)
+
+Suggests a nonlinear relationship between basic sanitation and TB incidence.
+Diminishing or increasing effects at different levels of sanitation.
+❌ First-degree polynomial term for sanitation (pro_basicsanitation¹) is NOT significant (p = 0.286)
+
+This indicates that a linear effect of sanitation is weak.
+The nonlinear (quadratic) effect is more important.
+
+*Performing Generalized Additive Models (GAM)*
+
+```{r}
+model4 <- gamm(log(tbi) ~ s(hdi, bs="ps") + s(pro_basicwater, bs="ps") +
+                         s(pro_safewater, bs="ps") + s(pro_basicsanitation, bs="ps") +
+                         s(country, bs="re"),
+              data = asia_africa_clean, method = "REML")
+# bam() from mgcv, which is more efficient for mixed models
+summary(model4$gam)
+plot(model4$gam)
+
+```
+
+```{r}
+AIC(model1)
+AIC(model2)
+AIC(model3)
+AIC(model4)
+
+BIC(model1)
+BIC(model2)
+BIC(model3)
+BIC(model4)
+```
+```{r}
+# Create a data frame for model comparison
+model_comparison <- data.frame(
+  Model = c("Model 1: Main Predictors", "Model 2: Including Covariates", "Model 3: Significant variables only", "Model 4: GAMM"),
+  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
+  BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4))
+)
+
+# Print the table
+print(model_comparison)
+
+library(here)
+
+# Define the file location using here()
+criteria_location <- here("code", "analysis-code", "tables", "criteria_output.rds")
+
+# Save the model
+saveRDS(model_comparison, file = criteria_location)
+
+# Save as a nicely formatted text file
+sink("model_comparison.txt")
+print(model_comparison)
+sink()
+
+```
+
+Model4 (GAM) Performs better.
+
+Lets visualize model4
+
+```{r}
+
+# Define the file path using here()
+plot_location <- here("code", "analysis-code", "figures", "gam_diagnostics.jpg")
+
+# Open JPEG device
+jpeg(plot_location, width = 2000, height = 1500, res = 300)
+
+# Arrange the plots in a grid (2 rows, 3 columns)
+par(mfrow = c(2, 3))  
+plot(model4$gam, pages = 1)  # Generate all GAM diagnostic plots
+
+# Close the graphics device to save the file
+dev.off()
+
+# Confirm file saved
+print(paste("Plot saved at:", plot_location))
+
+```
+
+```{r}
+# Create summary table
+gam_summary <- summary(model4$gam)
+
+# Extract relevant statistics
+gam_table <- data.frame(
+  Term = rownames(gam_summary$s.table),
+  EDF = gam_summary$s.table[, 1],
+  F_Value = gam_summary$s.table[, 3],
+  P_Value = gam_summary$s.table[, 4]
+)
+
+# Define the file location using here()
+summary_location <- here("code", "analysis-code", "tables", "summary_Gamm.rds")
+
+# Save the model
+saveRDS(gam_table, file = summary_location)
 ```

### 2025-03-19 code/eda-code/eda.qmd
Data Analysis (Part 3 of the project)

@@
 library(GGally)
+
 # Install and load the patchwork package
@@
 library(patchwork)
+
+install.packages("gtsummary")
+library(gtsummary)
 ```
@@
   filter(region == c("South-East Asia", "Africa"))
+
+asia_africa <- asia_africa%>%
+  filter(country != "Lesotho") #Drop Lesotho from the observation
+
 unique(asia_africa$country)
@@
 
-# Select only numeric columns (excluding country, year, and region)
+# Reshape data to long format for faceting
 asia_africa_long <- asia_africa %>%
-  select(-region) %>%  # Exclude categorical variable
-  pivot_longer(cols = -c(country, year),  # Convert all numeric variables except country & year
-               names_to = "Variable", 
-               values_to = "Value")
+  pivot_longer(cols = c(tbi, pro_basicwater, pro_safewater, pro_basicsanitation, 
+                        pro_safesanitation, gdp, hdi),
+               names_to = "variable", values_to = "value")
 
-# Create line plots for all variables
-plot2 <- ggplot(asia_africa_long, aes(x = year, y = Value, group = country, color = country)) +
-  geom_line(size = 1, alpha = 0.8) +  # Alpha for transparency
-  facet_wrap(~ Variable, scales = "free_y") +  # Separate plots for each variable
-  labs(title = "Trends of All Variables Over Time",
-       x = "Year",
-       y = "Value",
-       color = "Country") +
-  theme_minimal() +
-  theme(legend.position = "none",  # Hide legend for clarity
-        axis.text.x = element_text(angle = 45, hjust = 1))
+# Create spaghetti plot with facets
+plot2 <- ggplot(asia_africa_long, aes(x = year, y = value, color = country)) + 
+  geom_line(aes(group = country), alpha = 0.7) +  # Different color for each country
+  stat_smooth(method = "loess", col = "black", lwd = 1.5, se = FALSE) +  # Mean line in black
+  scale_x_continuous(name = "Year") + 
+  scale_y_continuous(name = "Value") + 
+  facet_wrap(~ variable, scales = "free_y") +  # Create facets for each variable
+  theme_bw() + 
+  theme(axis.text = element_text(size = 12, colour = "black"), 
+        axis.title = element_text(size = 14, face = "bold"),
+        strip.text = element_text(size = 14, face = "bold"),  # Style facet titles
+        legend.position = "none")  # Remove legend if not needed
 
@@
 plot(plot2)
+
 ```
@@
 ```{r}
-plot3 <- ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
-  geom_line(alpha = 0.7) +  # Transparency to see overlapping lines
-  labs(title = "Spaghetti Plot of TBC Incidence",
-       x = "Year",
-       y = "TBC Incidence") +
-  theme_minimal() +
-  theme(legend.position = "none")
 
+plot3 <-  ggplot(asia_africa, aes(x = year, y = tbi, color = country)) + 
+  geom_line(aes(group = country), alpha = 0.7) +  # Different color for each country
+  stat_smooth(method = "loess", col = "black", lwd = 1.5, se = FALSE) +  # Mean line in black
+  scale_x_continuous(name = "Year") + 
+  scale_y_continuous(name = "Tuberculosis Incidence") +
+  theme_bw() + 
+  theme(axis.text = element_text(size = 14, colour = "black"), 
+        axis.title = element_text(size = 14, face = "bold"),
+        strip.text.x = element_text(size = 14),
+        legend.position = "none")  # Remove legend if not needed
 plot_spageti_all <- here( "code","eda-code", "figures", "spageti-tbi.png") # to set up location for the pictures created 
@@
 ```{r}
+
 # Find a scaling factor to adjust TBI to match percentage values
-max_indicator <- max(asia_africa[, c("pro_safesanitation", "pro_basicsanitation", "pro_safewater", "pro_basicwater", "pro_cleanfuel")], na.rm = TRUE)
+max_indicator <- max(asia_africa[, c("pro_safesanitation", "pro_basicsanitation", "pro_safewater", "pro_basicwater")], na.rm = TRUE)
 max_tbi <- max(asia_africa$tbi, na.rm = TRUE)
@@
 asia_africa_long <- asia_africa %>%
-  pivot_longer(cols = c(pro_safesanitation, pro_basicsanitation, pro_safewater, pro_basicwater, pro_cleanfuel),
+  pivot_longer(cols = c(pro_safesanitation, pro_basicsanitation, pro_safewater, pro_basicwater, tbi),
                names_to = "Indicator", 
@@
   geom_line(data = asia_africa, 
-            aes(x = year, y = tbi * scale_factor, color = "TBI"), 
+            aes(x = year, y = tbi * scale_factor, color = "tbi"), 
             size = 1, linetype = "dashed") + 
@@
   theme_minimal() +
-  labs(title = "Percentage of Population with Access to Water, Sanitation, Clean Fuel & TBI Trends",
+  labs(title = "Percentage of Population with Access to Water, Sanitation & TBI Trends",
        x = "Year",
@@
   scale_y_continuous(sec.axis = sec_axis(~ . / scale_factor, name = "TBI")) +  # Secondary Y-axis for TBI
-  scale_color_manual(values = c("blue", "green", "purple", "orange", "brown", "red")) +  # Custom colors
+  scale_color_manual(values = c("blue", "green", "purple", "orange", "red")) +  # Custom colors
   theme(legend.position = "bottom",  
@@
 ```{r}
-# Load required libraries
-library(dplyr)
-library(tidyr)
-library(kableExtra)
-library(here)
 
-# Define row colors for each category
-row_colors <- c(
-  "TBI" = "#FFD700", # Gold
-  "GDP" = "#ADD8E6", # Light Blue
-  "HDI" = "#90EE90", # Light Green
-  "Clean Fuel" = "#FFB6C1", # Light Pink
-  "Safe Water" = "#87CEEB", # Sky Blue
-  "Basic Sanitation" = "#FFA07A", # Light Salmon
-  "Safe Sanitation" = "#F08080", # Light Coral
-  "PM2.5" = "#D3D3D3", # Light Gray
-  "Greenhouse" = "#B0C4DE" # Light Steel Blue
-)
+summary_table <- tbl_summary(asia_africa, include = -c(country, region))
 
-# Create summary per year
-summary_table <- asia_africa %>%
-  group_by(year) %>%  
-  summarise(
-    Min_TBI = min(tbi, na.rm = TRUE),
-    Mean_TBI = mean(tbi, na.rm = TRUE),
-    Max_TBI = max(tbi, na.rm = TRUE),
-    SD_TBI = sd(tbi, na.rm = TRUE),
-
-    Min_GDP = min(gdp, na.rm = TRUE),
-    Mean_GDP = mean(gdp, na.rm = TRUE),
-    Max_GDP = max(gdp, na.rm = TRUE),
-    SD_GDP = sd(gdp, na.rm = TRUE),
-
-    Min_HDI = min(hdi, na.rm = TRUE),
-    Mean_HDI = mean(hdi, na.rm = TRUE),
-    Max_HDI = max(hdi, na.rm = TRUE),
-    SD_HDI = sd(hdi, na.rm = TRUE),
-
-    Min_CleanFuel = min(pro_cleanfuel, na.rm = TRUE),
-    Mean_CleanFuel = mean(pro_cleanfuel, na.rm = TRUE),
-    Max_CleanFuel = max(pro_cleanfuel, na.rm = TRUE),
-    SD_CleanFuel = sd(pro_cleanfuel, na.rm = TRUE),
-
-    Min_SafeWater = min(pro_safewater, na.rm = TRUE),
-    Mean_SafeWater = mean(pro_safewater, na.rm = TRUE),
-    Max_SafeWater = max(pro_safewater, na.rm = TRUE),
-    SD_SafeWater = sd(pro_safewater, na.rm = TRUE),
-
-    Min_BasicSanitation = min(pro_basicsanitation, na.rm = TRUE),
-    Mean_BasicSanitation = mean(pro_basicsanitation, na.rm = TRUE),
-    Max_BasicSanitation = max(pro_basicsanitation, na.rm = TRUE),
-    SD_BasicSanitation = sd(pro_basicsanitation, na.rm = TRUE),
-
-    Min_SafeSanitation = min(pro_safesanitation, na.rm = TRUE),
-    Mean_SafeSanitation = mean(pro_safesanitation, na.rm = TRUE),
-    Max_SafeSanitation = max(pro_safesanitation, na.rm = TRUE),
-    SD_SafeSanitation = sd(pro_safesanitation, na.rm = TRUE),
-
-    Min_PM2.5 = min(mean_exposure_pm2.5, na.rm = TRUE),
-    Mean_PM2.5 = mean(mean_exposure_pm2.5, na.rm = TRUE),
-    Max_PM2.5 = max(mean_exposure_pm2.5, na.rm = TRUE),
-    SD_PM2.5 = sd(mean_exposure_pm2.5, na.rm = TRUE),
-
-    Min_Greenhouse = min(total_greenhouses, na.rm = TRUE),
-    Mean_Greenhouse = mean(total_greenhouses, na.rm = TRUE),
-    Max_Greenhouse = max(total_greenhouses, na.rm = TRUE),
-    SD_Greenhouse = sd(total_greenhouses, na.rm = TRUE)
-  ) %>%
-  pivot_longer(cols = -year, names_to = "Statistic", values_to = "Value") %>%
-  pivot_wider(names_from = year, values_from = Value) %>% # Convert to wide format
-  rowwise() %>% 
-  mutate(Accumulated = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
-  ungroup() %>%
-  mutate(across(where(is.numeric), ~ sprintf("%.2f", .x))) %>% # Format numbers to 2 decimal places
-  relocate(Accumulated, .after = Statistic) %>% # Move Accumulated after Statistic
-  mutate(Row_Index = row_number(), # Create row indices for coloring
-         Category = case_when(
-           grepl("TBI", Statistic) ~ "TBI",
-           grepl("GDP", Statistic) ~ "GDP",
-           grepl("HDI", Statistic) ~ "HDI",
-           grepl("CleanFuel", Statistic) ~ "Clean Fuel",
-           grepl("SafeWater", Statistic) ~ "Safe Water",
-           grepl("BasicSanitation", Statistic) ~ "Basic Sanitation",
-           grepl("SafeSanitation", Statistic) ~ "Safe Sanitation",
-           grepl("PM2.5", Statistic) ~ "PM2.5",
-           grepl("Greenhouse", Statistic) ~ "Greenhouse"
-         ))
-
-# Generate table with colors
-summary_table_display <- summary_table %>%
-  select(-Category, -Row_Index) %>% # Remove Category before displaying
-  kable(caption = "Summary Statistics by Year with Accumulated Column") %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
-
-# Apply row coloring dynamically
-for (cat in names(row_colors)) {
-  rows_to_color <- which(summary_table$Category == cat) # Get row indices
-  summary_table_display <- summary_table_display %>%
-    row_spec(rows_to_color, background = row_colors[cat])
-}
-
-saveRDS(summary_table_display, file = here("code", "eda-code", "tables", "summary_table.html"))
-print(summary_table_display)
+saveRDS(summary_table, file = here("code", "eda-code", "tables", "summary_table.rds"))
+print(summary_table)
 
@@
 ```{r}
-#| label: tbl-resulttable1
-#| tbl-cap: "Summary Statistics Table"
+#| label: tbl-summarytable
+#| tbl-cap: "Data summary table."
 #| echo: FALSE
-
-htmltools::includeHTML(here("code", "eda-code", "tables", "summary_table.html"))
+resulttable= read_rds(here("code", "eda-code", "tables", "summary_table.rds"))
+knitr::kable(resulttable, format = "pipe")
 
@@
 ```{r}
-gg1 <- ggplot(asia_africa, aes(x = gdp, y = tbi)) +
-  geom_point() +  # Create scatter plot
-  ggtitle("Scatter Plot: tbi vs GDP") +  # Add title
-  xlab("GDP") +  # Label the x-axis
-  ylab("TBI") +  # Label the y-axis
-  theme_minimal()
 
-gg2 <- 
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+# Convert dataset to long format for faceting
+asia_africa_long <- asia_africa %>%
+  pivot_longer(cols = c( pro_basicwater, pro_safewater, 
+                        pro_basicsanitation, pro_safesanitation, hdi),
+               names_to = "variable", values_to = "value")
+
+# Create scatter plot faceted by independent variable
+scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
+  geom_point(alpha = 0.6) +  # Scatter points, colored by country
+  stat_smooth(method = "loess", col = "black", lwd = 1.2, se = FALSE) +  # LOESS (non-linear) in black
+  facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
+  labs(x = "Independent Variable Value", y = "Tuberculosis Incidence (TBI)") +
+  theme_bw() + 
+  theme(axis.text = element_text(size = 12, colour = "black"), 
+        axis.title = element_text(size = 14, face = "bold"),
+        strip.text = element_text(size = 14, face = "bold"), 
+        legend.position = "none")  # Hides the legend
+
+# Show plot
+print(scatter_faceted)
+
 ```
@@
 
-# Create a list of scatter plots for each independent variable against tbi
+library(ggplot2)
+library(patchwork)  # For arranging multiple plots
+
+# Define independent variables
+independent_vars <- c("hdi", "pro_basicsanitation", "pro_safesanitation", 
+                      "pro_basicwater", "pro_safewater")
+
+# Create individual scatter plots
 plots <- lapply(independent_vars, function(var) {
   ggplot(asia_africa, aes_string(x = var, y = "tbi")) +  
-    geom_point() + 
-    ggtitle(paste("tbi vs", var)) + 
+    geom_point(color = "black") +  # Black points only, no line
+    ggtitle(paste("TBI vs", var)) + 
     xlab(var) + 
@@
 
-# Combine all plots into a single frame
-combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want
-combined_plot
+combineplot_location <- here( "code","eda-code", "figures", "scatterplot.png") # to set up location for the pictures created 
+ggsave(filename = combineplot_location, plot=final_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+plot(final_plot)
 
-combineplot_location <- here( "code","eda-code", "figures", "scatterplot.png") # to set up location for the pictures created 
-ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+```
+
+```{r}
+# Remove rows where tbi or any independent variable has missing values
+asia_africa_clean <- asia_africa %>%
+  filter(!is.na(tbi)) %>% # Remove missing tbi values
+  drop_na(all_of(independent_vars))  # Remove missing values in independent variables
+
+# Create scatter plots again with cleaned data
+plots <- lapply(independent_vars, function(var) {
+  ggplot(asia_africa_clean, aes_string(x = var, y = "tbi", color = "country")) +
+    geom_point(alpha = 0.7) +
+    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "solid", size = 1) +
+    ggtitle(paste("TBI vs", var)) + 
+    xlab(var) + 
+    ylab("TBI") + 
+    theme_minimal() +
+    theme(legend.position = "none")  
+})
+
+# Combine all plots
+combined_plot <- wrap_plots(plots, ncol = 3)
+
+# Save the plot
+#combineplot_location <- here("code", "eda-code", "figures", "scatterplot.png") 
+#ggsave(filename = combineplot_location, plot = combined_plot, width = 16, height = 12, units = "in", dpi = 300)
+
+# Display the plot
 plot(combined_plot)

### 2025-03-19 products/manuscript/Manuscript.qmd
Data Analysis (Part 3 of the project)

@@
 ---
-title: "The effects of multi environmental exposures on Tuberculosis Incidence in South East Asian and African countris"
+title: "The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countris"
 author: ""
@@
 Tuberculosis is linked to environmental factors. This study will analyze
-the effects of multi environmental factors on tuberculosis incidence in
-Asia and Africa regions, where the majority of tuberculosis coming from
-these regions. The environmental factors and country economic and human
-development index as covariates.
+the effects of multi environmental factors, specially water and
+sanitation on tuberculosis incidence in Asia and Africa regions, where
+the majority of tuberculosis coming from these regions. The water and
+sanitation will be the main exposures and Human Development Index as a
+control variable.
 
@@
 The study will use multi sources of data including World Bank, World
-Health Organization (WHO), United Nation for Children Fund (Unicef),
-UN-HABITAT, Emissions Database for Global Atmospheric Research (EDGAR),
-and Institute for Health Metrics and Evaluation (IHME). Different
-dataset contains different variable, however all dataset have same year,
-country name and country codes that can be used to combine the dataset.
-The data also contain information of countries globally from 2000 to
-2020. The variables will include proportion of urban population living
-in slum areas, percentage of population using safely managed sanitation
-services, percentage of population using at least basic drinking water
-service, fuel for cooking, Population-weighted exposure to ambient PM2.5
-pollution, total green houses, and will include GDP/ country economic
-status, human development index, and Universal Health Coverage as
-covariates.
-
-Note: I have downloaded the dataset, but I have not conducted any
-cleaning or exploration on the dataset. List of data sources and website
-is provided in an excel fine in my data folder. I will include them in
-my reference for the next steps.
+Health Organization (WHO), United Nation for Children Fund (Unicef).
+Different dataset contains different variable, however all dataset have
+same year, country name and country codes that can be used to combine
+the dataset. The data also contain information of countries globally
+from 2000 to 2020. The variables will include percentage of population
+using safely managed sanitation services, percentage of population using
+at least basic drinking water service, percentage of population using
+well-managed water, percentage of population using basic sanitation,
+percentage of population using well-managed sanitation, and human
+development index.
 
@@
 
-The main aim of this study is to look at the impacts of environmental
-factors on tuberculosis incidence.
+The main aim of this study is to look at the effect of Water and
+Sanitation on on tuberculosis incidence.
 
-RQ: What environmental factors associated with tuberculosis incidence?
+RQ: Does access to water and sanitation effect Tuberculosis incidence in
+Asian and African Regions?
 
-H0: there is no relationship between proportion of urban population
-living in slum areas and tuberculosis.
+Null Hypotheisis
 
-H0: thee is no relationship between safely managed sanitation and
+H0: thee is no effect of access to basic sanitation on tuberculosis
+incidence
+
+H0: thee is no effect of access to safely managed sanitation and on
 tuberculosis incidence
 
-H0: there is no relationship between drinking water and tuberculosis
-incidence
-
-H0: there is no relationship between fuel for cooking and tuberculosis
-incidence
-
-H0: there is no relationship between PM 2.5 exposure and tuberculosis
-incidence
-
-H0: there is no association between total green gas houses and
+H0: there is no effect of access to access basic drinking water and
 tuberculosis incidence
 
-Ha: there is a relationship between proportion of urban population
-living in slum areas and tuberculosis.
-
-Ha: there is a relationship between safely managed sanitation and
+H0: there is no effect of access to safely managed drinking water on
 tuberculosis incidence
 
-Ha: there is a relationship between drinking water and tuberculosis
-incidence
+Alternative Hypothesis Ha: thee is an effect of access to basic
+sanitation on tuberculosis incidence
 
-Ha: there is a relationship between fuel for cooking and tuberculosis
-incidence
+Ha: thee is an effect of access to safely managed sanitation and on
+tuberculosis incidence
 
-Ha: there is a relationship between PM 2.5 exposure and tuberculosis
-incidence
+Ha: there is an effect of access to access basic drinking water and
+tuberculosis incidence
 
-Ha: there is a association between total green gas houses and
+Ha: there is an effect of access to safely managed drinking water on
 tuberculosis incidence
@@
 
-The datasets were loded one by one, then drop all unused variable before
-comined those datasets using reduce() function. Countries, which have at
-least one missing years from 2000 to 2020, were dropped from the study.
-To make sure that the data is clean and ready to analyse, all missing
-values were dropped. This study focuses on South East Asia and Africa
-Regions, where the major proportion of tuberculosis cases came.
+The datasets were loaded one by one, then drop all unused variable
+before combined those datasets using reduce() function. Countries, which
+have at least one missing years from 2000 to 2020, were dropped from the
+study. To make sure that the data is clean and ready to analyse, all
+missing values were dropped. This study focuses on South East Asia and
+Africa Regions, where the major proportion of tuberculosis cases came.
 Therefore, a newsubset for South East Asia and African countries was
@@
 
-After data cleaning, multiple exploratory data analysis was forformed
-such us summary and vizualising the data. A bloxplot was created to show
-Tuberculosis Incidence in each country during the time frame. Spagheti
+After data cleaning, multiple exploratory data analysis was performed
+such us summary and visualizing the data. A bloxplot was created to show
+Tuberculosis Incidence in each country during the time frame. Spaghetti
 plot was created to visualize the trend of tuberculosis incidence and
@@
 incidence and other predictors. Therefore, Spearman's correlation matrix
-was perfomed to see the correlations. To have better understanding, a
-heatmap matrix plot was created to show the correlation.
+was performed to see the correlations. To have better understanding, a
+heat map matrix plot was created to show the correlation.
 
-The main statistical analysis in thie study is longitudinal data
-analysis (eithre fixed effect or mixed effect method). To see which one
-is better, Hausman Test was performed to determine which model is
-better.
+The main statistical analysis in the study is longitudinal data analysis
+(either fixed effect or mixed effect method). To see which one is
+better, Hausman Test was performed to determine which model is better.
 
@@
 
-## Data aquisition
+## Data acquisition
 
-Most of the data is downloaded from Worlbank Dataset. However, those
-data are originally from many different courses/ organizations, such as
-World Bank, WHA, UNICEF, UN-HABITAT, EDGAR, and IHME.
+The data was downloaded from Worlbank Dataset. However, those data are
+originally from many different courses/ organizations, such as World
+Bank, WHA, UNICEF, UN-HABITAT, EDGAR, and IHME.
 
@@
 
-Data Import and data cleaning process have been explained in the methode
+Data Import and data cleaning process have been explained in the method
 part. To see the code for data import and data cleaning , please visit
@@
 Processed and clean data are saved at a folder ("data",
-"processed-data"). There are many .rds files in the folder icnlduing
+"processed-data"). There are many .rds files in the folder indulging
 file for each variables, and file after combined all variables.
@@
 
+The study applied Generalized Additive Mixed Model (GAMM). The model can
+capture mixed effect of countries on Tuberculosis Incidence with non
+linear relationship. The data analysis started by looking at the
+relationship between Tuberculosis Incidence and Predictors. Non linear
+relationship was discovered.Next step is to perform non-linear
+relationship to looking at form of relationship. Human Development Index
+(HDI), Access to basic water and well-managed water show curvature
+relationship, while access to basic sanitation and well-managed
+sanitation show polynomial. Data transformation was performed based.
+After data transformation, we fit model1, which only includes water and
+sanitation variables (as main variables in this study), model2 includes
+all predictors (HDI, access to water and sanitation), model3 only
+includes significant variables in the previous model. Model 1, 2 and 3
+use Generalized Linear Mixed Model (GLMM). We also performed VIF test to
+look at multicollinearity. Model 4 uses Generalized Additive Mixed
+Models. Moreover, BIC and AIC tests were performed to look at the best
+model. BIC and AIC tests found that Model 4 (GAMM) performs better with
+the smallest BIC and AIC. To be able to interpret GAMM better, we
+visualize the result into plots.
+
 {{< pagebreak >}}
@@
 
-NOTE: I still have issue in saving the stiled tables. I am still working
-on it. I will show it next step.
+NOTE: I still have issue in saving the tables. I am still working on it.
+I will show it next step.
 
@@
 #| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
+resulttable=readRDS (here("code", "eda-code", "tables", "summary_table.rds"))
 knitr::kable(resulttable)
@@
 showed the decreases trend in tuberculosis incidence over 21 years
-period, except Lesotho showing a sharp increase until the middle of the
-period before dramatic decrease until the end of the period. All
-countries showed the increase trend in percentage of population using
-using basic and safe drinking water, basic and safe sanitation, and
-clean fuel and technology for cooking, except Zimbabuwe showing the
-decrease in percentage of population using basic drinking water and safe
-sanitation, and Senegal showing the decrease of population using clean
-fuel and technology for cooking.
+period. All countries showed the increase trend in percentage of
+population using using basic and safe drinking water, basic and safe
+sanitation, except Zimbabuwe showing the decrease in percentage of
+population using basic drinking water and safe sanitation.
 
@@
 most of the countries show the increase trends in water and sanitation,
-clean fuel use, human development index, total greenhouses, but
-fluctuated trends in mean of PM2.5 expousures.
+and human development index
 
@@
 #| label: fig-3
-#| fig-cap: "Spaghetti plots in each variables"
+#| fig-cap: "Boxplot plots for each Country"
 #| echo: FALSE
@@
 
-As first step of statistical analysis.
-Using scatterplot to show the linearity between tuberculosis incidence and predictors.
-( @fig-4 ) shows the scatterplots.
-The scatterplots do not show linear relationship between tuberculosis Incidence and predictors.
+As first step of statistical analysis. Using scatter plot to show the
+linearity between tuberculosis incidence and predictors. ( @fig-4 )
+shows the scatter plots. The scatter plots do not show linear
+relationship between tuberculosis Incidence and predictors.
 
@@
 ```
-Before going further, we need to look at correlation between tuberculosis Incidence and all predictors.
 
-Since they show nonlinearity, spearman's correlation performed and hsowing in the following correlation matrix. See ( @fig-5 )
+Before going further, we need to look at correlation between
+tuberculosis Incidence and all predictors.
+
+Since they show non linearity, spearman's correlation performed and
+showing in the following correlation matrix. See ( @fig-5 )
 
@@
 
-Since there is no linear relationship, nonlinear relationship test was performed.
+Now, we want to assess non linear relationship.
 
@@
 
-After performing nonlinear model, data transformation used based on the shave or distribution of the plots from nonlinear relationship plots ( @fig-6 ) above.
-( @fig-7 ) shows the data distribution after data transformation.
+After performing nonlinear model, data transformation used based on the
+shave or distribution of the plots from nonlinear relationship plots (
+@fig-6 ) above. ( @fig-7 ) shows the data distribution after data
+transformation.
 
@@
 
-For the final model, Generalized linear models (GLMs) and Generalized Addictive Models (GAMs). I need to discuss with Dr. Handal to get advice about this.
-
-I actually have performed both GLMs and GAMs, but I still need to learn both models in depth.
-Specially for GAM, I am very new in the models.
-Initially I performed GLMs since the Transformed plots show linears line, but aftrer comparing the GLMs and GAMs model, GAMs perfoming better.
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+There are four models build; fitting model with only water and
+sanitation variables, then fitting models including co-variate (HDI).
+The third model is to fit only significant variables. The first three
+models use Generalized Linear Mixed Models (GLMM), while the last model
+used Generalized Additive Mixed Models (GAMM). Those models were
+compared using AIC and BIC criteria to select the best/ the final
+models.
 
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| label: tbl-summarytable
+#| tbl-cap: "Model Criteria"
 #| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
+resulttable2=readRDS (here("code", "analysis-code", "tables", "criteria_output.rds"))
 knitr::kable(resulttable2)
+```
+
+Based on BIC and AIC criteria, the best model fit the data is
+Generalized Additive Mixed Models.
+
+\$ \log(\text{TBI}) \sim s(\text{hdi}) + s(\text{pro\_basicwater}) +
+s(\text{pro\_safewater}) + s(\text{pro\_basicsanitation}) +
+s(\text{country}, \text{bs}="re") \$
+
+```{r}
+#| label: tbl-summarytable
+#| tbl-cap: "Generalized Additive Mixed Models"
+#| echo: FALSE
+resulttable3=readRDS (here("code", "analysis-code", "tables", "summary_Gamm.rds"))
+knitr::kable(resulttable3)
+```
+
+This study employed a Generalized Additive Model (GAM) with a Gaussian
+family and identity link function to examine the association between
+tuberculosis incidence and key environmental and socioeconomic factors
+across multiple countries. The model incorporated smooth functions for
+continuous predictors and a random effect for country to account for
+within-country variability. The results indicate that HDI and basic
+sanitation are significant determinants of tuberculosis incidence, while
+access to basic and safe water did not show strong associations.
+
+The smooth function for Human Development Index (HDI) was highly
+significant (p \< 2e-16), with an estimated effective degrees of freedom
+(EDF) = 4.283, confirming a non-linear relationship between HDI and
+tuberculosis incidence. The effect was strongest at lower HDI values,
+suggesting that improvements in socioeconomic conditions may have the
+greatest impact in reducing tuberculosis burden in low-HDI countries.
+
+Among the water and sanitation variables, basic water access was not
+statistically significant (p = 0.73379), indicating no strong evidence
+of an association between water coverage and tuberculosis incidence.
+Similarly, safe water access was not significant at α = 0.05 (p =
+0.05605), suggesting only a weak relationship. However, basic sanitation
+access was statistically significant (p = 0.00542, EDF = 3.845),
+suggesting that higher access to sanitation is associated with a
+reduction in tuberculosis incidence. This highlights the importance of
+improving sanitation infrastructure as a potential intervention for
+tuberculosis control.
+
+The random effect of country (s(country)) was highly significant (p \<
+2e-16, EDF = 16.768), indicating substantial between-country variation
+in tuberculosis incidence. This justifies the inclusion of random
+effects to account for country-level heterogeneity and underscores the
+need for context-specific policy interventions tailored to national
+conditions.
+
+The model fit was strong, with an adjusted R² of 0.974, indicating that
+97.4% of the variance in tuberculosis incidence was explained by the
+model. The scale estimate was 0.014598, suggesting a low residual
+variance, which reflects a well-fitting model.
+
+```{r}
+#| label: fig-7
+#| fig-cap: "Generalized Additive Mixed Models"
+#| echo: FALSE
+knitr::include_graphics(here("code","analysis-code","figures","gam_diagnostics.jpg"))
+
 ```
@@
 *What are the main take-home messages?*
+
+These findings underscore the significant role of socioeconomic
+conditions and sanitation infrastructure in tuberculosis control. While
+access to basic and safe water was not a significant predictor,
+improvements in sanitation and overall socioeconomic development (HDI)
+were strongly associated with lower tuberculosis incidence. Given the
+significant country-level differences, tuberculosis prevention
+strategies should be tailored to national contexts, emphasizing
+improvements in living standards and sanitation access. These results
+provide valuable insights for policymakers aiming to design targeted
+public health interventions in high-burden settings.
 

### 2025-04-08 assets/statistical-analysis.qmd
Edit manuscript

--- 2025-02-23 code/analysis-code/statistical-analysis.qdm.qmd
@@
 #load needed packages. make sure they are installed.
-library(here) #for data loading/saving
+library(here)
 library(dplyr)
@@
 library(ggplot2)
-install.packages("maps")
+#install.packages("maps")
 library(maps)
-install.packages("ggcorrplot")
+#install.packages("ggcorrplot")
 library(ggcorrplot)
-install.packages("plm")
+#install.packages("plm")
 library(plm)
-install.packages("gt")
+#install.packages("gt")
 library(gt)
-install.packages("car")
+#install.packages("car")
 library(car)
-install.packages("MASS")  # Jika belum terinstall
+#install.packages("MASS")
 library(MASS)
@@
 library(tidyverse)
-install.packages("reshape2")  # Install the package (only needed once)
-library(reshape2)  # Load the package
-install.packages("scales")
+#install.packages("reshape2") 
+library(reshape2) 
+#install.packages("scales")
 library(scales)
-
 library(splines)
@@
 library(knitr)
-install.packages("kableExtra")
 library(kableExtra)
-install.packages("GGally")
+#install.packages("GGally")
 library(GGally)
-# Install and load the patchwork package
-install.packages("patchwork")
+#install.packages("patchwork")
 library(patchwork)
-
-install.packages("gridExtra")  # If not installed
+#install.packages("gridExtra")
 library(gridExtra)
+#install.packages("lmerTest")
+library(lmerTest)
+#install.packages("stargazer")
+library(stargazer)
+#install.packages("lmtest")
+library(lmtest)
+#install.packages("mgcv") 
+library(mgcv)  
+install.packages("gamm4")
+library(gamm4)
+install.packages("glmnet")
+library(glmnet)
+install.packages("brms")
+library(brms)
 ```
@@
 
-asia_africa_location <- here("data", "processed-data", "asia_africa.rds")
-asia_africa <- readRDS(asia_africa_location)
+seaadf_location <- here("data", "processed-data", "seaa.rds")
+seaa_df <- readRDS(seaadf_location)
 
-colnames(asia_africa)
+str(seaa_df)
+
+seaa_df <- seaa_df %>%
+  mutate(across(c(pro_basicwater, pro_safewater, 
+                  pro_basicsanitation, pro_safesanitation), 
+                ~ as.numeric(.)))
 ```
@@
 
-***Test the linearity***
+First, we perform linear regression. Due to high multicolinearity, percentage of population using safe sanitation is eliminated from the model. 
+```{r}
+# Fit a linear model (without Interaction)
+model_lm1 <- lm(tbi ~ pro_basicwater + pro_safewater + 
+               pro_basicsanitation +  hdi, 
+               data = seaa_df)
+summary(model_lm1)
+```
+```{r}
+# Fit a linear model (with Interaction)
+model_lm_inte <- lm(tbi ~ (pro_basicwater + pro_safewater + 
+               pro_basicsanitation +  hdi)^2, 
+               data = seaa_df)
+summary(model_lm_inte )
+```
+
+```{r}
+# Plot residuals vs. fitted values for the linear model
+plot(model_lm1$fitted.values, resid(model_lm1), 
+     xlab = "Fitted Values", ylab = "Residuals", 
+     main = "Residuals vs. Fitted Plot")
+abline(h = 0, col = "red")  # Horizontal line at 0
+```
+
+# Fit a non-linear model (GAM)
+```{r}
+
+model_gam1 <- gam(tbi ~ s(pro_basicwater) + s(pro_safewater) + s(pro_basicsanitation) + s(hdi), 
+                 data = seaa_df, method = "REML")
+
+```
+
+compared the linear model and non linear model using AIC
 
@@
 
-# Define independent variables (exclude 'tbi' and categorical variables)
-independent_vars <- colnames(asia_africa)[!colnames(asia_africa) %in% c("tbi", "country", "region", "year")]
+```
 
-# Create scatter plots for each independent variable against tbi
-plots <- lapply(independent_vars, function(var) {
-  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
-    geom_point() + 
-    ggtitle(paste("TBI vs", var)) + 
-    xlab(var) + 
-    ylab("TBI") + 
-    theme_minimal()
-})
+```{r}
+model_gam_inte <- gam(tbi ~ 
+                    s(pro_basicwater) + 
+                    s(pro_safewater) + 
+                    s(pro_basicsanitation) + 
+                    s(hdi) + 
+                    te(pro_basicwater, pro_safewater) +
+                    te(pro_basicwater, pro_basicsanitation) +
+                    te(pro_basicwater, hdi) +
+                    te(pro_safewater, pro_basicsanitation) +
+                    te(pro_safewater, hdi) +
+                    te(pro_basicsanitation, hdi),
+                  data = seaa_df, method = "REML")
+summary(model_gam_inte)
+```
+```{r}
+# Compare AIC values
+AIC(model_lm1, model_lm_inte, model_gam1, model_gam_inte)
+```
+It can be seen that GAM model with two way interaction performs best.
 
-# Create a list of scatter plots for each independent variable against tbi
-plots <- lapply(independent_vars, function(var) {
-  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
-    geom_point() + 
-    ggtitle(paste("tbi vs", var)) + 
-    xlab(var) + 
-    ylab("TBI") + 
-    theme_minimal()
-})
+### Longitudinal analysis with Mixed-effect models
 
-# Combine all plots into a single frame
-combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol to the number of columns you want
+```{r}
+# make sure to deal with NA data
+seaa_df <- seaa_df %>%
+  drop_na(tbi,  pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, hdi, year, country)
 
-combineplot_location <- here( "code","analysis-code", "figures", "scatterplot.png") # to set up location for the pictures created 
-ggsave(filename = combineplot_location, plot=combined_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
-plot(combined_plot)
+# Model Mixed-Effects: TBI is predicted by predictors with random effect per each country 
+model_me <- lmer(tbi ~ pro_basicwater + pro_safewater + pro_basicsanitation + pro_safesanitation + hdi + year + (1 | country),  # Random intercept dressed variation in each country
+                   data = seaa_df, REML = FALSE)
 
+# Ringkasan hasil model
+summary(model_me)
 ```
@@
 # Reshape the dataset to long format
-asia_africa_long <- asia_africa %>%
-  pivot_longer(cols = c(pro_cleanfuel, pro_basicwater, pro_safewater, 
-                        mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, 
-                        pro_safesanitation, gdp, hdi),
+asia_africa_long <- seaa_df %>%
+  pivot_longer(cols = c( pro_basicwater, pro_safewater, pro_basicsanitation, 
+                        pro_safesanitation,hdi),
                names_to = "Predictor", values_to = "Value")
@@
 
-nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
+nonrelationshiplot_location <- here( "results", "figures", "analysis", "nonlinear.png") # to set up location for the pictures created 
 ggsave(filename = nonrelationshiplot_location, plot=nonlinear, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
@@
 
-Data transformation
+It can be seen that some variables show u-shaped (quadratic), however some variables looks more complex.
+In this case, we are going to try to fit two models, first model will fit Mixed effects model with Polynomial Terms, and addictive mixed model (GAMM).
+
+I will perform two steps here: 1. Data transformation, then looking at possible model, 2. Directly performing Mixed effects model with polynomial terms, and Performing GAMM
 
@@
 
-library(dplyr)
-
-asia_africa <- asia_africa %>%
+seaa_df<- seaa_df %>%
   mutate(
     log_tbi = log(tbi + 1),
-    log_gdp = log(gdp + 1),
-    log_total_greenhouses = log(total_greenhouses + 1),
-    log_pm2.5 = log(mean_exposure_pm2.5 + 1),
     sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
     sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
-    poly_cleanfuel = as.numeric(poly(pro_cleanfuel, 2, raw = TRUE)[, 1]),
     poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
@@
 
+Visualize the linearity test after transfomation
 ```{r}
-asia_africa_transformed <- asia_africa %>%
-  dplyr::select(log_tbi, log_gdp, log_total_greenhouses, log_pm2.5, 
-                poly_cleanfuel, poly_basicsanitation, poly_safesanitation,
+seaa_transformed <- seaa_df %>%
+  dplyr::select(log_tbi, poly_basicsanitation, poly_safesanitation,
                 sqrt_basicwater, sqrt_safewater, hdi) %>%
@@
 
-colnames(asia_africa)
+colnames(seaa_transformed)
 
 # Plot after transformation
-transformed <- ggplot(asia_africa_transformed, aes(x = Value, y = log_tbi)) +
+plot_transformed <- ggplot(seaa_transformed, aes(x = Value, y = log_tbi)) +
   geom_point(alpha = 0.3) +
-  geom_smooth(method = "lm", color = "red") +  # Use linear model for comparison
+  geom_smooth(method = "loess", color = "red") +  # Use linear model for comparison
   facet_wrap(~ Predictor, scales = "free_x") +
@@
 
-transformedplot_location <- here( "code","analysis-code", "figures", "transformed.png") # to set up location for the pictures created 
-ggsave(filename = transformedplot_location, plot=transformed, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
-plot(transformed)
+transformedplot_location <- here("results", "figures",  "analysis", "plot_transformed.png") # to set up location for the pictures created 
+ggsave(filename = transformedplot_location, plot=plot_transformed, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+plot(plot_transformed)
+```
+*Performing Generalized Additive Models (GAM)*
+```{r}
+
+#  Fit the Bayesian GAMM
+model_brms_full <- brm(
+  formula = tbi ~ 
+    s(pro_basicwater) + 
+    s(pro_safewater) + 
+    s(pro_basicsanitation) + 
+    s(pro_safesanitation) + 
+    s(hdi) + 
+    s(year) + 
+    (1 | country),  # random intercept for longitudinal structure
+
+  data = seaa_df,
+  family = negbinomial())
+
+```
+
+```{r}
+# Model summary
+summary(model_brms_full)
+
+# Check model convergence and diagnostics
+plot(model_brms_full)           # smooth terms & convergence
+pp_check(model_brms_full)       # posterior predictive check
+conditional_effects(model_brms_full)  # visualize effects
+
+# Check random effects
+ranef(model_brms_full)
+
+# Save fitted values or residuals if needed
+seaa_df$fitted_tbi <- fitted(model_brms_full)[, "Estimate"]
+
+```
+
+```{r}
+model_brms_full <- brms(
+  formula = tbi ~ 
+    s(pro_basicwater) + 
+    s(pro_safewater) + 
+    s(pro_basicsanitation) + 
+    s(pro_safesanitation) + 
+    s(hdi) + 
+    s(year) + 
+    (1 | country),
+
+  data = seaa_df,
+  family = negbinomial(),
+
+  control = list(adapt_delta = 0.99),  # Fix for divergent transitions
+
+  chains = 4,
+  cores = 4,
+  iter = 4000
+)
+
+)
+
+```
+
+```{r}
+
+wash_vars <- seaa_df[, c("pro_basicwater", "pro_safewater", "pro_basicsanitation", "pro_safesanitation")]
+pca_result <- prcomp(wash_vars, scale. = TRUE)
+
+# Add first component to your data
+seaa_df$PC1 <- pca_result$x[, 1]
+
+model <- gamm4(
+  tbi ~ s(PC1) + s(hdi) + s(year),
+  random = ~(1 | country),
+  family = nb(),  
+  data = seaa_df
+)
+
+# Define the model formula with smooth terms
+formula <- tbi ~ s(hdi) + s(pro_basicwater) + s(pro_safewater) + s(pro_basicsanitation) + s(pro_safesanitation) + s(year)
+
+# Fit the GAMM with a negative binomial family to account for overdispersion
+gamm_model <- gamm(formula, random = list(country = ~1), family = nb(), data = seaa_df)
+
+# Extract the design matrix and response variable
+X <- model.matrix(gamm_model$gam)
+y <- seaa_df$tbi
+
+# Apply elastic net regularization (alpha = 0.5 for elastic net, 0 for ridge, 1 for lasso)
+cv_fit <- cv.glmnet(X, y, alpha = 0.5, family = "poisson")
+
+# Obtain the coefficients at the optimal lambda
+coef(cv_fit, s = "lambda.min")
+
+```
+
+after performing data transformation, the data is stil not linear, therefore, Geralized additive mixed morel, which can handle non linearity and account mixed effect of the ciuntries.
+
+```{r}
+#convert country into factors
+
+seaa_df$country <- as.factor(seaa_df$country)
+
+model_gamm <- gamm(
+  tbi ~ 
+    s(hdi) + 
+    s(pro_safewater) + 
+    s(pro_basicsanitation),
+  random = list(country = ~1),
+  data = seaa_df,
+  method = "REML"
+)
+
+model_gamm4 <- gamm4(
+  log_tbi ~ 
+    s(hdi) +
+    s(sqrt_basicwater) +
+    s(sqrt_safewater) +
+    s(poly_basicsanitation) +
+    s(poly_safesanitation),
+  random = ~(1 | country),
+  data = seaa_df
+)
+
+summary(model_gamm4$gam)  # for fixed effects (smooth terms)
+summary(model_gamm4$mer)  # for random effects
+plot(model_gamm4$gam, pages = 1)
+
+summary(model_gamm_trans$gam)
+plot(model_gamm_trans$gam)
+
+```
+
+```{r}
+AIC(model1)
+AIC(model2)
+AIC(model3)
+AIC(model4)
+
+BIC(model1)
+BIC(model2)
+BIC(model3)
+BIC(model4)
+```
+
+```{r}
+# Create a data frame for model comparison
+model_comparison <- data.frame(
+  Model = c("Model 1: Main Predictors", "Model 2: Including Covariates", "Model 3: Significant variables only", "Model 4: GAMM"),
+  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
+  BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4))
+)
+
+# Print the table
+print(model_comparison)
+
+library(here)
+
+# Define the file location using here()
+criteria_location <- here("code", "analysis-code", "tables", "criteria_output.rds")
+
+# Save the model
+saveRDS(model_comparison, file = criteria_location)
+
+# Save as a nicely formatted text file
+sink("model_comparison.txt")
+print(model_comparison)
+sink()
+
+```
+
+Model4 (GAM) Performs better.
+
+Lets visualize model4
+
+```{r}
+
+# Define the file path using here()
+plot_location <- here("code", "analysis-code", "figures", "gam_diagnostics.jpg")
+
+# Open JPEG device
+jpeg(plot_location, width = 2000, height = 1500, res = 300)
+
+# Arrange the plots in a grid (2 rows, 3 columns)
+par(mfrow = c(2, 3))  
+plot(model4$gam, pages = 1)  # Generate all GAM diagnostic plots
+
+# Close the graphics device to save the file
+dev.off()
+
+# Confirm file saved
+print(paste("Plot saved at:", plot_location))
+
+```
+
+```{r}
+# Create summary table
+gam_summary <- summary(model4$gam)
+
+# Extract relevant statistics
+gam_table <- data.frame(
+  Term = rownames(gam_summary$s.table),
+  EDF = gam_summary$s.table[, 1],
+  F_Value = gam_summary$s.table[, 3],
+  P_Value = gam_summary$s.table[, 4]
+)
+
+# Define the file location using here()
+summary_location <- here("code", "analysis-code", "tables", "summary_Gamm.rds")
+
+# Save the model
+saveRDS(gam_table, file = summary_location)
 ```
@@
 
-gam_model <- gam(log_tbi ~ s(log_gdp) + s(log_total_greenhouses) + s(log_pm2.5) +
-                      s(poly_cleanfuel) + s(poly_basicsanitation) + s(poly_safesanitation) +
+gam_model <- gam(log_tbi ~ s(poly_basicsanitation) + s(poly_safesanitation) +
                       s(sqrt_basicwater) + s(sqrt_safewater) + s(hdi),
-                 data = asia_africa, family = gaussian)
+                 data = seaa_df, family = gaussian)
 summary(gam_model)

### 2025-04-08 code/analysis-code/statistical-analysis.qmd
Edit manuscript

---
title: "statistical-analysis"
format: html
editor: visual
---

##

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
#install.packages("maps")
library(maps)
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("plm")
library(plm)
#install.packages("gt")
library(gt)
#install.packages("car")
library(car)
#install.packages("MASS") 
library(MASS)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("reshape2")  
library(reshape2) 
#install.packages("scales")
library(scales)
library(splines)
library(here)
library(knitr)
#install.packages("kableExtra")
library(kableExtra)
#install.packages("GGally")
library(GGally)
#install.packages("patchwork")
library(patchwork)
#install.packages("gridExtra") 
library(gridExtra)
#install.packages("lmerTest")
library(lmerTest)
#install.packages("stargazer")
library(stargazer)
#install.packages("lmtest")
library(lmtest)
#install.packages("mgcv")
library(mgcv)
#install.packages("Metrics")
library(Metrics)
#install.packages("reticulate")
library(reticulate)
#install.packages("xgboost")
library(xgboost)
#install.packages("tidymodels")y
library(tidymodels)
library(Metrics)
library(gamm4)
library(broom)
install.packages("gratia")
library(gratia)
library(dplyr)

```

Load the dataset

```{r}

asia_africa_location <- here("data", "processed-data", "seaa.rds")
asia_africa <- readRDS(asia_africa_location)

asia_africa <- asia_africa %>%
  filter(country != "Lesotho") #drop Lesotho, it is potential influential outlier

colnames(asia_africa)

asia_africa$pro_basicwater <- as.numeric(asia_africa$pro_basicwater)
asia_africa$pro_safewater <- as.numeric(asia_africa$pro_safewater)
asia_africa$pro_basicsanitation <- as.numeric(asia_africa$pro_basicsanitation)
asia_africa$pro_safesanitation <- as.numeric(asia_africa$pro_safesanitation)
asia_africa$country <- as.factor(asia_africa$country)
str(asia_africa)
```

## Data Analysis

```{r}
# Adding Fixed-Effect to Make Conditional Models ---
ggplot(asia_africa, aes(x = year, y = tbi)) + 
  geom_line(aes(group=country), col="grey") + 
  stat_smooth(method="loess", col="black", lwd=1.5, se=FALSE)+
  scale_x_continuous(name = "years") + 
  scale_y_continuous(name = "Tuberculosis Incidence") +
  theme_bw() + 
  theme(axis.text=element_text(size=14, colour="black"), 
        axis.title=element_text(size=14,face="bold")) +
  theme(strip.text.x = element_text(size = 14))+ theme(legend.position="none") 
```

```{r}
# Create scatter plot faceted by independent variable
scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
  geom_point(alpha = 0.6) +  # Scatter points, colored by country
  stat_smooth(method = "loess", col = "black", lwd = 1.2, se = FALSE) +  # LOESS (non-linear) in black
  facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
  labs(
    title = "TBI and Its Association with Water, Sanitation, and HDI Indicators",
    x = "Independent Variable Value",
    y = "Tuberculosis Incidence (TBI)"
  ) +
  theme_bw() + 
  theme(axis.text = element_text(size = 12, colour = "black"), 
        axis.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"), 
        legend.position = "none")  # Hides the legend

#save the plot 
plots_tbi_vs_allvars_loc <- here( "results", "figures", "analysis", "plots_tbi_vs_allvars.png") # to set up location for the pictures created 
ggsave(filename = plots_tbi_vs_allvars_loc  , plot=scatter_faceted, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created

print(scatter_faceted)

```

Non-linearity test

```{r}
# Compute the correlation matrix for selected variables
cor(asia_africa[, c("tbi", "pro_basicwater", "pro_safewater", 
                    "pro_basicsanitation", "hdi")])

# Fit a linear model
lm_model <- lm(tbi ~  pro_basicwater + pro_safewater + 
               pro_basicsanitation + hdi, 
               data = asia_africa)

# Run the RESET test for linearity
resettest(lm_model, power = 2:3)

```

```{r}
# Fit a linear model (LM)
lm_model <- lm(tbi ~ pro_basicwater + pro_safewater + 
               pro_basicsanitation +  hdi, 
               data = asia_africa)

# Fit a non-linear model (GAM)
gam_model <- gam(tbi ~ s(pro_basicwater) + s(pro_safewater) + 
                  s(pro_basicsanitation)  + s(hdi), 
                 data = asia_africa, method = "REML")

# Compare AIC values
AIC(lm_model, gam_model)

# Plot residuals vs. fitted values for the linear model
plot(lm_model$fitted.values, resid(lm_model), 
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs. Fitted Plot")
abline(h = 0, col = "red")  # Horizontal line at 0

```

```{r}

gamm_model_interactions<- gam(
  tbi ~ 
    s(year) + 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) +  
    ti(pro_basicwater, hdi) + 
    ti(pro_safewater, hdi) +
    ti(pro_basicsanitation, hdi) + 
    ti(pro_safesanitation, hdi) +
    ti(pro_safesanitation, pro_basicwater) +
    ti(pro_safewater, pro_basicwater) +
    s(country, bs = "re"),         
  family = nb(),                  
  method = "REML",                 
  data = asia_africa
)

summary(gamm_model_interactions)
AIC(gamm_model_interactions)

```

Note:

The warning indicate that the model is too complex, I will delete the least significant interactions:

```{r}

gamm_model_interactions_reduced<- gam(
  tbi ~ 
    s(year) + 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) +  
    ti(pro_basicwater, hdi) + 
    ti(pro_safewater, hdi) +
    ti(pro_safesanitation, pro_basicwater) +
    ti(pro_safewater, pro_basicwater) +
    s(country, bs = "re"),         
  family = nb(),                  
  method = "REML",                 
  data = asia_africa
)

AIC(gamm_model_interactions_reduced)

summary(gamm_model_interactions_reduced)

```

### Visualize the result of the final model

```{r}
# Extract summary of the model
model_summary <- summary(gamm_model_interactions_reduced)

# Parametric terms
param_tbl <- as.data.frame(model_summary$p.table) %>%
  tibble::rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    z.value = `z value`,
    p.value = `Pr(>|z|)`
  ) %>%
  mutate(
    type = "Parametric",
    edf = NA, 
    chi_sq = NA
  )

# Smooth terms
smooth_tbl <- as.data.frame(model_summary$s.table) %>%
  tibble::rownames_to_column("term") %>%
  rename(
    edf = edf,
    chi_sq = `Chi.sq`,
    p.value = `p-value`
  ) %>%
  mutate(
    type = "Smooth",
    estimate = NA, 
    std.error = NA, 
    z.value = NA
  )

# Combine parametric and smooth tables
model_results_tbl <- bind_rows(param_tbl, smooth_tbl) %>%
  select(type, term, estimate, std.error, edf, chi_sq, p.value)

# Display table with gt
model_results_tbl %>%
  gt(groupname_col = "type") %>%
  fmt_number(columns = where(is.numeric), decimals = 3) %>%
  tab_header(
    title = "Summary of GAMM Estimates",
    subtitle = "Parametric and Smooth Terms"
  ) %>%
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. Error",
    edf = "EDF",
    chi_sq = "Chi-square",
    p.value = "p-value"
  ) %>%
  sub_missing(everything(), missing_text = "-")

#save the table

# Save as RDS
saveRDS(model_results_tbl, file = here::here("results", "tables", "analysis", "gamm_summary_table.rds"))

```

NULL Model

```{r}
model_null_gamm <- gamm(tbi ~ 1, random = list(country = ~1), data = asia_africa)
```

```{r}

AIC(lm_model)
AIC(gam_model)
AIC(gamm_model_interactions)
AIC(gamm_model_interactions_reduced)

```

Based on the table, although Generalized model with full interaction results the smallest AIC, its df is negative indicating the model instability. Therefore, we used the GAMM with reduced model, which least significant interaction is reduced.

***Visualize the AIC***

```{r}
# create AIC data frame
aic_results <- data.frame(
  Model = c("LM", "GAM", "GAMM + Interactions", "GAMM + Interactions (Reduced)"),
  AIC = c(4735.692, 4568.658, 2956.188, 2954.157)
)

# create plot 
plot_aic <- ggplot(aic_results, aes(x = reorder(Model, AIC), y = AIC)) +
  geom_point(size = 4, color = "darkgreen") +
  geom_segment(aes(xend = Model, yend = 0), linetype = "dashed") +
  geom_text(aes(label = round(AIC, 1)), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(title = "AIC by Model", x = "Model", y = "AIC") +
  theme_minimal()

plot(plot_aic)
```

\*\*\* Print RMSE from those models\*\*\*

```{r}

# 1. LM Model
pred_lm <- predict(lm_model)
rmse_lm <- sqrt(mean((asia_africa$tbi - pred_lm)^2))

# 2. GAM Model
pred_gam <- predict(gam_model, type = "response")
rmse_gam <- sqrt(mean((asia_africa$tbi - pred_gam)^2))

# 4. GAMM Model with Interactions
pred_gamm_interactions <- predict(gamm_model_interactions, type = "response")
rmse_gamm_interactions <- sqrt(mean((asia_africa$tbi - pred_gamm_interactions)^2))

# 5. Reduced GAMM Model with Interactions
pred_gamm_interactions_red <- predict(gamm_model_interactions_reduced, type = "response")
rmse_gamm_interactions_red <- sqrt(mean((asia_africa$tbi - pred_gamm_interactions_red)^2))

# 6. NULL Model (GAMM null model)
pred_null <- predict(model_null_gamm$gam, type = "response")
rmse_null <- sqrt(mean((asia_africa$tbi - pred_null)^2))

# Print results
rmse_results <- data.frame(
  Model = c("Null Model", "LM", "GAM",  "GAMM + Interactions", "GAMM + Interactions (Reduced)" ),
  RMSE = c(rmse_null, rmse_lm, rmse_gam, rmse_gamm_interactions, rmse_gamm_interactions_red )
)

saveRDS(rmse_results, file = here("results", "tables", "analysis", "rmse_results.rds"))

print(rmse_results)
```

Althought GAMM with full interaction has smallest RMSE, based on the negative value of DF. we chose Gamm with interactons (reduced)

```{r}

#| label: tbl-rmse
#| tbl-cap: "RMSE Comparison of Models"
#| echo: false

# Load the table
rmse_results <- readRDS(here("results", "tables", "analysis", "rmse_results.rds"))

# Display the table
kable(rmse_results, digits = 2)
```

Vizualise the RMSE

```{r}

plot_RMSE <- ggplot(rmse_results, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_point(size = 4, color = "darkred") +
  geom_segment(aes(xend = Model, yend = 0), linetype = "dashed") +
  geom_text(aes(label = round(RMSE, 2)), 
            hjust = -0.2, size = 4, color = "black") +
  coord_flip() +
  labs(title = "RMSE by Model", x = "Model", y = "RMSE") +
  theme_minimal()

ggsave(
  filename = here("results", "figures", "analysis", "rmse_by_model.png"),
  plot = plot_RMSE,
  width = 7, height = 5, dpi = 300
)

plot(plot_RMSE)

# facet the AIC and RMSE plots 
Plot_AIC_RMSE <- plot_aic + plot_RMSE + plot_layout(ncol = 2)

# Save the plot 
ggsave(
  here("results", "figures", "analysis", "plot_aic_rmse.png"),
  plot = Plot_AIC_RMSE,
  width = 12, height = 8, dpi = 300
)

plot(Plot_AIC_RMSE)
```

Predicted and observed TBI from original dat

```{r}

asia_africa <- asia_africa %>%
  mutate(predicted_tbi = predict(gamm_model_interactions_reduced, newdata = asia_africa, type = "response"))

#Create plot
plot_observed_predicted_original <- ggplot(asia_africa, aes(x = predicted_tbi, y = tbi)) +
  geom_point(alpha = 0.6, color = "#1f77b4", size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(
    title = "Predicted vs Observed TBI (original data)",
    x = "Predicted TBI",
    y = "Observed TBI"
  ) +
  theme_minimal(base_size = 14)

# Save the plot 
ggsave(
  here("results", "figures", "analysis", "plot_observed_predicted_original.png"),
  plot = plot_observed_predicted_original,
  width = 10, height = 8, dpi = 300
)

plot(plot_observed_predicted_original)

```

#Machine Learning read the dataset

*Splitted Dataset* 
```{r}
# create setseed for reproductibility 
ml_seed <- set.seed(2345)

# Split dataset into a 75% train set and a 25% test set

splitted_ml <- initial_split(asia_africa, prop= .75)
df_train <- training(splitted_ml)
df_test <- testing(splitted_ml)

```

***fit the models in to machine learning***

The study is supposed to use Mixed Effect Random Forest (MERF). Since there is no a specific package for MERF in R,I will use xgboost

***Linear Model***

```{r}
# Fit a linear model (LM) with train data
lm_model_train <- lm(tbi ~ pro_basicwater + pro_safewater + 
               pro_basicsanitation + pro_safesanitation +  hdi, 
               data = df_train)

# Fit a non-linear model (GAM)
gam_model_train <- gam(tbi ~ s(pro_basicwater) + s(pro_safewater) + 
                  s(pro_basicsanitation) + s(pro_safesanitation) + s(hdi), 
                 data = df_train, method = "REML")

# Fit Gamm Model without interaction with training data 
gamm_model_train <- gam(
  tbi ~ 
    s(year) + 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) +  
    s(country, bs = "re"),         
  family = nb(),                  
  method = "REML",                 
  data = df_train
)

summary(gamm_model_train)

# Fit GamM Model with interaction  with training data

gamm_model_interactions_train <- gam(
  tbi ~ 
    s(year) + 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) +  
    ti(pro_basicwater, hdi) + 
    ti(pro_safewater, hdi) +
    ti(pro_basicsanitation, hdi) + 
    ti(pro_safesanitation, hdi) +
    ti(pro_safesanitation, pro_basicwater) +
    ti(pro_safewater, pro_basicwater) +
    s(country, bs = "re"),         
  family = nb(),                  
  method = "REML",                 
  data = df_train
)

summary(gamm_model_interactions_train)

```
***Fit generated data with  final model***
```{r}
# Fit GamM Model with interaction (reduced) with training data 
gamm_model_interactions_reduced_train  <- gam(
  tbi ~ 
    s(year) + 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) +  
    ti(pro_basicwater, hdi) + 
    ti(pro_safewater, hdi) +
    ti(pro_safesanitation, pro_basicwater) +
    ti(pro_safewater, pro_basicwater) +
    s(country, bs = "re"),         
  family = nb(),                  
  method = "REML",                 
  data = df_train
)

summary(gamm_model_interactions_reduced_train  )

```

**Fit Xgboost with train data**\*

```{r}

# Manually create interaction terms (to mimic GAMM)
df_train <- df_train %>%
  mutate(
    int_basicwater_hdi        = pro_basicwater * hdi,
    int_safewater_hdi         = pro_safewater * hdi,
    int_safesanitation_basic  = pro_safesanitation * pro_basicwater,
    int_safewater_basic       = pro_safewater * pro_basicwater
  )

df_test<- df_test %>%
  mutate(
    int_basicwater_hdi        = pro_basicwater * hdi,
    int_safewater_hdi         = pro_safewater * hdi,
    int_safesanitation_basic  = pro_safesanitation * pro_basicwater,
    int_safewater_basic       = pro_safewater * pro_basicwater
  )

# One-hot encode 'country' and create full model matrix (train data) 
X_train <- model.matrix(
  tbi ~ year + pro_basicwater + pro_safewater + pro_basicsanitation +
    pro_safesanitation + hdi + country + 
    int_basicwater_hdi + int_safewater_hdi +
    int_safesanitation_basic + int_safewater_basic,
  data = df_train
)[, -1]  # remove intercept

# Test Data 
X_test <- model.matrix(
  tbi ~ year + pro_basicwater + pro_safewater + pro_basicsanitation +
    pro_safesanitation + hdi + country + 
    int_basicwater_hdi + int_safewater_hdi +
    int_safesanitation_basic + int_safewater_basic,
  data = df_test
)[, -1]

# Extract outcome
y_train <- df_train$tbi
y_test <- df_test$tbi

# Fit XGBoost with count data (train data) 
xgb_model_train <- xgboost(
  data = X_train,
  label = y_train,
  objective = "count:poisson",
  nrounds = 100,
  eta = 0.1,
  max_depth = 6,
  subsample = 0.8,
  colsample_bytree = 0.8,
  verbose = 0
)

# Fit XGBoost with count data (testdata) 
xgb_model_test <- xgboost(
  data = X_test,
  label = y_test,
  objective = "count:poisson",  # appropriate for count outcomes
  nrounds = 100,
  eta = 0.1,
  max_depth = 6,
  subsample = 0.8,
  colsample_bytree = 0.8,
  verbose = 0
)

# Predict and evaluate

preds_xgb_train <- predict(xgb_model_train, newdata = X_train)
preds_xgb_test <- predict(xgb_model_test, newdata = X_test)

df_rmse_train <- tibble(truth = y_train, estimate = preds_xgb_train)
df_rmse_test  <- tibble(truth = y_test, estimate = preds_xgb_test)

#RMSE 
rmse(df_rmse_train, truth = truth, estimate = estimate)
rmse(df_rmse_test, truth = truth, estimate = estimate)

```

Create a plot

```{r}
# Combine training data
df_train_plot <- tibble(
  actual = y_train,
  predicted = preds_xgb_train,
  dataset = "Train"
)

plot(df_train_plot)

# Combine test data
df_test_plot <- tibble(
  actual = y_test,
  predicted = preds_xgb_test,
  dataset = "Test"
)
plot(df_test_plot)

# Combine both for plotting
df_plot <- bind_rows(df_train_plot, df_test_plot)

# Plot predicted vs observed
plot_obs_pred_ml <- ggplot(df_plot, aes(x = actual, y = predicted, color = dataset)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~ dataset) +
  labs(
    title = "Predicted vs Observed TBI",
    x = "Observed TBI",
    y = "Predicted TBI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Save the plot 
ggsave(
  here("results", "figures", "analysis", "plot_obs_pred_ml.png"),
  plot = plot_obs_pred_ml,
  width = 10, height = 8, dpi = 300
)

plot(plot_obs_pred_ml)
```

***Predict and Evaluate the models using test data***

```{r}
# Predict
lm_pred_test <- predict(lm_model_train, newdata = df_test, type = "response")
gam_pred_test <- predict(gam_model_train, newdata = df_test, type = "response")
gamm_pred_test <- predict(gamm_model_train, newdata = df_test, type = "response")
gamm_inte_pred_test <- predict(gamm_model_interactions_train, newdata = df_test, type = "response")
gamm_inte_pred_reduced_test <- predict(gamm_model_interactions_reduced_train, newdata = df_test, type = "response")
xgb_pred_test <- predict(xgb_model_train, newdata = X_test)

# Evaluate RMSE
rmse_ml_lm <- rmse(tibble(truth = df_test$tbi, estimate = lm_pred_test), truth, estimate)
rmse_ml_gam <- rmse(tibble(truth = df_test$tbi, estimate = gam_pred_test), truth, estimate)
rmse_ml_gamm <- rmse(tibble(truth = df_test$tbi, estimate = gamm_pred_test), truth, estimate)
rmse_ml_gamm_inte <- rmse(tibble(truth = df_test$tbi, estimate = gamm_inte_pred_test), truth, estimate)
rmse_ml_gamm_inte_reduced <- rmse(tibble(truth = df_test$tbi, estimate = gamm_inte_pred_reduced_test), truth, estimate)
rmse_ml_xgb <- rmse(tibble(truth = df_test$tbi, estimate = xgb_pred_test), truth, estimate)

```

```{r}

# Your RMSE results — replace with your actual values if different
rmse_original <- tibble(
  Model = c("Null Model", "LM", "GAM", "GAMM", "GAMM + Interactions", "GAMM + Interactions (Reduced)"),
  RMSE = c(148.58, 125.13, 85.17, 18.29, 7.77, 8.05),
  Data = "Full Data"
)

rmse_test <- tibble(
  Model = c("Null Model", "LM", "GAM", "GAMM", "GAMM + Interactions", "GAMM + Interactions (Reduced)"),
  RMSE = c(NA, 129.34, 102.64, 24.48, 10.73, 10.54),
  Data = "Test Data"
)

# Combine both into one dataframe
rmse_combined <- bind_rows(rmse_original, rmse_test)

```

```{r}

ggplot(rmse_combined, aes(x = Model, y = RMSE, fill = Data)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.6) +
  labs(
    title = "RMSE Comparison: Full Data vs Test Data",
    y = "RMSE (Lower is Better)",
    x = "Model"
  ) +
  scale_fill_manual(values = c("Full Data" = "#4E79A7", "Test Data" = "#F28E2B")) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

```

Now I want to plot the predicted value vs observed value between training data and test data

```{r}
# Let's say your best model is this
gamm_model_interactions_test <- gam(
  tbi ~ 
    s(year) + s(pro_basicwater) + s(pro_safewater) +
    s(pro_basicsanitation) + s(pro_safesanitation) + s(hdi) +
    ti(pro_basicwater, hdi) + ti(pro_safewater, hdi) +
    ti(pro_safesanitation, pro_basicwater) + ti(pro_safewater, pro_basicwater) +
    s(country, bs = "re"),
  data = df_test,
  family = nb(),
  method = "REML"
)

# Predicted values
train_data_pred <- df_train %>%
  mutate(Source = "Train", .fitted = predict(gamm_model_interactions_reduced_train, type = "response"))

test_data_pred <- df_test%>%
  mutate(Source = "Test", .fitted = predict(gamm_model_interactions_test, newdata = df_test, type = "response"))

# Combine
pred_obs_df <- bind_rows(train_data_pred, test_data_pred)

# Create Plot

plot_observed_predicted <- ggplot(pred_obs_df, aes(x = .fitted, y = tbi, color = Source, shape = Source)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Observed vs Predicted TBI (Training vs Test Data)",
    x = "Predicted TBI",
    y = "Observed TBI"
  ) +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Train" = "#1f77b4", "Test" = "#ff7f0e"))

#Save the plot
ggsave(
  here("results", "figures", "analysis", "plot_observed_predicted_train_test .png"),
  plot = plot_observed_predicted ,
  width = 10, height = 8, dpi = 300
)

plot(plot_observed_predicted )

#Facet the observed vs predictied for ML data and original data

plot_ob_pre_ml_vs_ori <- 
plot(plot_observed_predicted_original + plot_observed_predicted + plot_layout(nrow  = 2))

# Save the plot 
ggsave(
  here("results", "figures", "analysis", "plot_ob_pre_ml_vs_ori.png"),
  plot =plot_ob_pre_ml_vs_ori,
  width = 12, height = 8, dpi = 300
)

plot(plot_ob_pre_ml_vs_ori)
```

1.  Excellent fit Both training and test points fall very close to the diagonal

That means your model accurately predicts TBI, both on data it has seen and on data it hasn’t seen

This is a sign of a well-generalizing model, not overfittin

2.  Low bias & low variance Predictions are neither consistently above nor below the true values

There’s minimal scatter, indicating good precision

3.  Consistent performance across train and test The orange (test) points closely track the blue (train) points

This tells you that your model doesn't just memorize, it learns the pattern and generalizes it

#Tunning the model

```{r}

set.seed(2345)

# Cross-validation folds (same as with LASSO)
folds <- vfold_cv(asia_africa, v = 10)

#Define tunable XGBoost model
xgb_model <- boost_tree(
  trees = 1000,
  learn_rate = tune(),
  tree_depth = tune(),
  mtry = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) %>%
  set_engine("xgboost", objective = "count:poisson") %>%
  set_mode("regression")

xgb_recipe <- recipe(tbi ~ ., data = asia_africa) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Create workflow

xgb_wf <- workflow() %>%
  add_model(xgb_model) %>%
  add_recipe(xgb_recipe)

# Define Tuning Grid 
xgb_grid <- grid_space_filling(
  tree_depth(),
  learn_rate(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), df_train),
  size = 20
)

# tune with grid search 
xgb_tuned <- tune_grid(
  xgb_wf,
  resamples = folds,
  grid = xgb_grid,
  metrics = metric_set(rmse)
)

# Visualize 
autoplot_xgb <- autoplot(xgb_tuned)

# Save to file using here()
ggsave(
  filename = here::here("results", "figures", "analysis", "autoplot_xgb_tuning.png"),
  plot = autoplot_xgb,
  width = 10,
  height = 8,
  dpi = 300
)
```

### 2025-04-08 code/eda-code/eda.qmd
Edit manuscript

@@
 library(ggplot2)
-install.packages("maps")
+#install.packages("maps")
 library(maps)
-install.packages("ggcorrplot")
+#install.packages("ggcorrplot")
 library(ggcorrplot)
-install.packages("plm")
+#install.packages("plm")
 library(plm)
-install.packages("gt")
+#install.packages("gt")
 library(gt)
-install.packages("car")
+#install.packages("car")
 library(car)
-install.packages("MASS")  # Jika belum terinstall
+#install.packages("MASS")  
 library(MASS)
-install.packages("tidyverse")
+#install.packages("tidyverse")
 library(tidyverse)
-install.packages("reshape2")  # Install the package (only needed once)
+#install.packages("reshape2")
 library(reshape2)  # Load the package
-install.packages("scales")
+#install.packages("scales")
 library(scales)
-
 library(here)
 library(knitr)
-install.packages("kableExtra")
+#install.packages("kableExtra")
 library(kableExtra)
-install.packages("GGally")
+#install.packages("GGally")
 library(GGally)
-
-# Install and load the patchwork package
-install.packages("patchwork")
+#install.packages("patchwork")
 library(patchwork)
-
-install.packages("gtsummary")
+#install.packages("gtsummary")
 library(gtsummary)
@@
 #Path to data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","clean_noslum.rds")
+aa_location <- here::here("data","processed-data","seaa.rds")
 
 #load data
-mydata <- readRDS(data_location)
+seaa_df <- readRDS(aa_location)
+unique(seaa_df$country)
 
-#subset for south east and Asia Africa only 
-asia_africa <- mydata %>% 
-  filter(region == c("South-East Asia", "Africa"))
+str(seaa_df)
+#convert some caracters variables into numerics
 
-asia_africa <- asia_africa%>%
-  filter(country != "Lesotho") #Drop Lesotho from the observation
+seaa_df <- seaa_df %>%
+  mutate(across(c(pro_basicwater, pro_safewater, 
+                  pro_basicsanitation, pro_safesanitation), 
+                ~ as.numeric(.)))
 
-unique(asia_africa$country)
-
-saveRDS(asia_africa, here("data", "processed-data", "asia_africa.rds"))
 ```
 
-# Data exploration through tables
+# Data exploration Analysis
+
+I want to create boxplot for TBI
 
@@
 # Define distinct colors for each country
+# Define selected countries
+selected_countries <- c(
+  "Algeria", "Bangladesh", "Central African Republic", "Chad", "Ethiopia", "Ghana",
+  "Indonesia", "Madagascar", "Malawi", "Myanmar", "Nepal", "Senegal",
+  "Sierra Leone", "Sri Lanka", "Tanzania", "Togo", "Uganda", "Zimbabwe"
+)
+
+# Custom colors for selected countries (excluding Sierra Leone for now)
 country_colors <- c(
@@
   "Indonesia" = "#42D4F4",  # Light Blue
-  "Lesotho" = "#F032E6",  # Pink
   "Madagascar" = "#FABEBE",  # Light Red
@@
   "Uganda" = "#000075",  # Dark Blue
-  "Zimbabwe" = "#808000"  # Olive
+  "Zimbabwe" = "#808000",  # Olive,
+  "Sierra Leone" = "#008080"  # Custom Teal (added for missing color)
 )
@@
 # Convert country to a factor
-asia_africa$country <- as.factor(asia_africa$country)
+seaa_df$country <- as.factor(seaa_df$country)
 
 # Create boxplot with borders
-plot1 <- ggplot(asia_africa, aes(x = country, y = tbi, fill = country)) +
+boxplot_tbi <- ggplot(seaa_df, aes(x = country, y = tbi, fill = country)) +
   geom_boxplot(color = "black", size = 0.1) +  # Add black border with thicker lines
@@
 
-plot_box <- here( "code","eda-code", "figures", "boxplot.png") # to set up location for the pictures created 
-ggsave(filename = plot_box, plot=plot1, width = 12, height = 8, units = "in", dpi = 300) # save the pictures created
+tbi_plot_box <- here( "results", "figures", "eda", "tbi_boxplot.png") # to set up location for the pictures created 
+ggsave(filename = tbi_plot_box, plot=boxplot_tbi, width = 12, height = 8, units = "in", dpi = 300) # save the pictures created
 
-plot(plot1)
+plot(boxplot_tbi)
 
@@
 
+*** create spagethi plot*** 
 ```{r}
 
-# Reshape data to long format for faceting
-asia_africa_long <- asia_africa %>%
-  pivot_longer(cols = c(tbi, pro_basicwater, pro_safewater, pro_basicsanitation, 
-                        pro_safesanitation, gdp, hdi),
-               names_to = "variable", values_to = "value")
-
-# Create spaghetti plot with facets
-plot2 <- ggplot(asia_africa_long, aes(x = year, y = value, color = country)) + 
-  geom_line(aes(group = country), alpha = 0.7) +  # Different color for each country
-  stat_smooth(method = "loess", col = "black", lwd = 1.5, se = FALSE) +  # Mean line in black
-  scale_x_continuous(name = "Year") + 
-  scale_y_continuous(name = "Value") + 
-  facet_wrap(~ variable, scales = "free_y") +  # Create facets for each variable
-  theme_bw() + 
-  theme(axis.text = element_text(size = 12, colour = "black"), 
-        axis.title = element_text(size = 14, face = "bold"),
-        strip.text = element_text(size = 14, face = "bold"),  # Style facet titles
-        legend.position = "none")  # Remove legend if not needed
-
-plot_spageti <- here( "code","eda-code", "figures", "spageti.png") # to set up location for the pictures created 
-ggsave(filename = plot_spageti, plot=plot2, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created
-
-plot(plot2)
-
-```
-
-```{r}
-
-plot3 <-  ggplot(asia_africa, aes(x = year, y = tbi, color = country)) + 
+# Line plot of TBI over time by country
+tbi_spaghety <- ggplot(seaa_df , aes(x = year, y = tbi, color = country)) + 
   geom_line(aes(group = country), alpha = 0.7) +  # Different color for each country
@@
   scale_y_continuous(name = "Tuberculosis Incidence") +
+  labs(title = "Tuberculosis Incidence Trends Over Time by Country") +  # Add title here
   theme_bw() + 
@@
         legend.position = "none")  # Remove legend if not needed
-plot_spageti_all <- here( "code","eda-code", "figures", "spageti-tbi.png") # to set up location for the pictures created 
-ggsave(filename = plot_spageti_all, plot=plot3, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created
 
-plot(plot3)
+tbi_plot_spageti <- here( "results", "figures", "eda", "tbi_spageti.png") # to set up location for the pictures created 
+ggsave(filename = tbi_plot_spageti, plot=tbi_spaghety, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created
+
+plot(tbi_spaghety)
+
 ```
@@
 ```{r}
+# Define scaling factor for TBI to align with percentage indicators
+tbi_scale <- 100 / max(seaa_df$tbi, na.rm = TRUE)
 
-# Find a scaling factor to adjust TBI to match percentage values
-max_indicator <- max(asia_africa[, c("pro_safesanitation", "pro_basicsanitation", "pro_safewater", "pro_basicwater")], na.rm = TRUE)
-max_tbi <- max(asia_africa$tbi, na.rm = TRUE)
-scale_factor <- max_indicator / max_tbi  # Scale TBI to fit
+# Reshape the percentage indicators into long format
+seaa_long <- seaa_df %>%
+  pivot_longer(
+    cols = c(pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation),
+    names_to = "indicator",
+    values_to = "value"
+  )
 
-# Reshape the data from wide to long format
-asia_africa_long <- asia_africa %>%
-  pivot_longer(cols = c(pro_safesanitation, pro_basicsanitation, pro_safewater, pro_basicwater, tbi),
-               names_to = "Indicator", 
-               values_to = "Value")
+# Prepare TBI data, scaled for plotting
+tbi_long <- seaa_df %>%
+  mutate(indicator = "TBI", value = tbi * tbi_scale) %>%
+  select(country, year, indicator, value)
 
-# Create the plot
-plot4 <- ggplot() +
-  # First, plot water, sanitation, and clean fuel indicators
-  geom_line(data = asia_africa_long, 
-            aes(x = year, y = Value, color = Indicator, group = Indicator), 
-            size = 1) +
-  # Add TBI as a separate line with a different scale
-  geom_line(data = asia_africa, 
-            aes(x = year, y = tbi * scale_factor, color = "tbi"), 
-            size = 1, linetype = "dashed") + 
-  facet_wrap(~ country) +  # Separate plots for each country
+# Combine percentage indicators and TBI
+combined_long <- bind_rows(seaa_long, tbi_long)
+
+# Set factor levels for consistent legend and color mapping
+combined_long$indicator <- factor(
+  combined_long$indicator,
+  levels = c("pro_basicwater", "pro_safewater", "pro_basicsanitation", "pro_safesanitation", "TBI"),
+  labels = c("Basic Water", "Safe Water", "Basic Sanitation", "Safe Sanitation", "TBI")
+)
+
+# Plot
+plot_all_vars <- ggplot(combined_long, aes(x = year, y = value, group = indicator)) +
+  geom_line(aes(color = indicator, linetype = indicator), size = 1) +
+  facet_wrap(~ country) +
+  scale_y_continuous(
+    name = "Water, Sanitation (%)",
+    sec.axis = sec_axis(~ . / tbi_scale, name = "TBI (per 100k)")
+  ) +
+  scale_color_manual(
+    values = c(
+      "Basic Water" = "#1f77b4",
+      "Safe Water" = "#ff7f0e",
+      "Basic Sanitation" = "#2ca02c",
+      "Safe Sanitation" = "#d62728",
+      "TBI" = "black"
+    )
+  ) +
+  scale_linetype_manual(
+    values = c(
+      "Basic Water" = "solid",
+      "Safe Water" = "solid",
+      "Basic Sanitation" = "solid",
+      "Safe Sanitation" = "solid",
+      "TBI" = "dashed"
+    )
+  ) +
+  labs(
+    title = "Trends of Water, Sanitation, HDI , and TBI (per 100k) by Country",
+    x = "Year",
+    color = "Indicator",
+    linetype = "Indicator"
+  ) +
   theme_minimal() +
-  labs(title = "Percentage of Population with Access to Water, Sanitation & TBI Trends",
-       x = "Year",
-       y = "Percentage (%)",
-       color = "Indicator") +  # Label for legend
-  scale_y_continuous(sec.axis = sec_axis(~ . / scale_factor, name = "TBI")) +  # Secondary Y-axis for TBI
-  scale_color_manual(values = c("blue", "green", "purple", "orange", "red")) +  # Custom colors
-  theme(legend.position = "bottom",  
+  theme(
+    panel.border = element_blank(),  # No small borders around each panel
+    plot.background = element_rect(color = "black", fill = NA, size = 1),  # Outer border
+    panel.background = element_blank(),
+    strip.text = element_text(size = 10, face = "bold"),
     axis.text.x = element_text(angle = 45, hjust = 1),
-        panel.border = element_rect(color = "black", fill = NA, size = 1))  # Rotate x-axis labels for readability
+    legend.position = "bottom"
+  )
 
-plot_location1 <- here( "code","eda-code", "figures", "line-graph-tbivsfactors.png") # to set up location for the pictures created 
-ggsave(filename = plot_location1, plot=plot4, width = 12, height = 16, units = "in", dpi = 300) # save the pictures created 
-plot(plot4)
+#save the plot 
+plot_allvars <- here( "results", "figures", "eda", "plot_all_vars.png") # to set up location for the pictures created 
+ggsave(filename = plot_allvars , plot=plot_all_vars, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created
+
+plot(plot_all_vars)
 ```
@@
 #| echo: FALSE
-knitr::include_graphics(here("code", "eda-code", "figures", "line-graph-tbivsfactors.png"))
+knitr::include_graphics(here("results",  "figures", "eda", "plot_all_vars.png"))
 ```
+From the plots, it can be seen that percentage of population in those countries increased while the Tuberculosis Incidence decreased during the period.
 
@@
 
-summary_table <- tbl_summary(asia_africa, include = -c(country, region))
+summary_table <- tbl_summary(seaa_df, include = -c(country, region))
 
-saveRDS(summary_table, file = here("code", "eda-code", "tables", "summary_table.rds"))
+saveRDS(summary_table, file = here("results", "tables", "eda", "summary_table.rds"))
 print(summary_table)
+
+```
+
+```{r}
+
+# Calculate mean of all relevant variables by year and region
+summary_table <- seaa_df %>%
+  group_by(region, year) %>%
+  summarise(
+    mean_tbi = mean(tbi, na.rm = TRUE),
+    mean_pro_basicwater = mean(pro_basicwater, na.rm = TRUE),
+    mean_pro_safewater = mean(pro_safewater, na.rm = TRUE),
+    mean_pro_basicsanitation = mean(pro_basicsanitation, na.rm = TRUE),
+    mean_pro_safesanitation = mean(pro_safesanitation, na.rm = TRUE),
+    mean_hdi = mean(hdi, na.rm = TRUE),
+    .groups = "drop"
+  )
+
+# View the first few rows
+head(summary_table)
 
@@
 
+***Create a plot to show the trend of each variables by region***
+```{r}
+
+# Create summary table: mean of each variable per region and year
+summary_df <- seaa_df %>%
+  group_by(region, year) %>%
+  summarise(
+    mean_tbi = mean(tbi, na.rm = TRUE),
+    mean_pro_basicwater = mean(pro_basicwater, na.rm = TRUE),
+    mean_pro_safewater = mean(pro_safewater, na.rm = TRUE),
+    mean_pro_basicsanitation = mean(pro_basicsanitation, na.rm = TRUE),
+    mean_pro_safesanitation = mean(pro_safesanitation, na.rm = TRUE),
+    mean_hdi = mean(hdi, na.rm = TRUE),
+    .groups = "drop"
+  )
+
+summary_long <- summary_df %>%
+  pivot_longer(
+    cols = starts_with("mean_"),
+    names_to = "indicator",
+    values_to = "mean_value"
+  ) %>%
+  mutate(indicator = dplyr::recode(as.character(indicator),
+    "mean_tbi" = "TBI",
+    "mean_pro_basicwater" = "Basic Water",
+    "mean_pro_safewater" = "Safe Water",
+    "mean_pro_basicsanitation" = "Basic Sanitation",
+    "mean_pro_safesanitation" = "Safe Sanitation",
+    "mean_hdi" = "HDI"
+  ))
+
+# Create line plot
+plot_all_vars_region <- ggplot(summary_long, aes(x = year, y = mean_value, color = region)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  facet_wrap(~ indicator, scales = "free_y") +
+  labs(
+    title = "Trends of TBI, HDI, Water and Sanitation Indicators by Region (Mean per Year)",
+    x = "Year",
+    y = "Mean Value",
+    color = "Region"
+  ) +
+  theme_minimal() +
+  theme(
+    strip.text = element_text(size = 11, face = "bold"),
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    legend.position = "bottom"
+  )
+
+#save the plot 
+plot_allvars_region <- here( "results", "figures", "eda", "plot_all_vars_region.png") # to set up location for the pictures created 
+ggsave(filename = plot_allvars_region  , plot=plot_all_vars_region, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created
+
+plot(plot_all_vars_region)
+```
+
 ## Data Analysis
@@
 
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-
 # Convert dataset to long format for faceting
-asia_africa_long <- asia_africa %>%
+seaa_df_long <- seaa_df %>%
   pivot_longer(cols = c( pro_basicwater, pro_safewater, 
@@
 # Create scatter plot faceted by independent variable
-scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
+plots_tbi_vs_allvars <- ggplot(seaa_df_long, aes(x = value, y = tbi, color = country)) + 
   geom_point(alpha = 0.6) +  # Scatter points, colored by country
@@
   facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
-  labs(x = "Independent Variable Value", y = "Tuberculosis Incidence (TBI)") +
+  labs(
+    title = "Relationship Between Tuberculosis Incidence and Independent Variables",
+    x = "Independent Variable Value",
+    y = "Tuberculosis Incidence (TBI)"
+  ) +
   theme_bw() + 
@@
 
+#save the plot 
+plots_tbi_vs_allvars_loc <- here( "results", "figures", "eda", "plots_tbi_vs_allvars.png") # to set up location for the pictures created 
+ggsave(filename = plots_tbi_vs_allvars_loc  , plot=plots_tbi_vs_allvars, width = 16, height = 12, units = "in", dpi = 300) # save the pictures created
+
 # Show plot
-print(scatter_faceted)
+print(plots_tbi_vs_allvars)
 
 ```
+Note:
 
-```{r}
+The plot shows non-linear relationship betwen Tuberculosis Incidence and other indevendent variables.
 
-library(ggplot2)
-library(patchwork)  # For arranging multiple plots
-
-# Define independent variables
-independent_vars <- c("hdi", "pro_basicsanitation", "pro_safesanitation", 
-                      "pro_basicwater", "pro_safewater")
-
-# Create individual scatter plots
-plots <- lapply(independent_vars, function(var) {
-  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +  
-    geom_point(color = "black") +  # Black points only, no line
-    ggtitle(paste("TBI vs", var)) + 
-    xlab(var) + 
-    ylab("TBI") + 
-    theme_minimal()
-})
-
-combineplot_location <- here( "code","eda-code", "figures", "scatterplot.png") # to set up location for the pictures created 
-ggsave(filename = combineplot_location, plot=final_plot, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
-plot(final_plot)
-
-```
-
-```{r}
-# Remove rows where tbi or any independent variable has missing values
-asia_africa_clean <- asia_africa %>%
-  filter(!is.na(tbi)) %>% # Remove missing tbi values
-  drop_na(all_of(independent_vars))  # Remove missing values in independent variables
-
-# Create scatter plots again with cleaned data
-plots <- lapply(independent_vars, function(var) {
-  ggplot(asia_africa_clean, aes_string(x = var, y = "tbi", color = "country")) +
-    geom_point(alpha = 0.7) +
-    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "solid", size = 1) +
-    ggtitle(paste("TBI vs", var)) + 
-    xlab(var) + 
-    ylab("TBI") + 
-    theme_minimal() +
-    theme(legend.position = "none")  
-})
-
-# Combine all plots
-combined_plot <- wrap_plots(plots, ncol = 3)
-
-# Save the plot
-#combineplot_location <- here("code", "eda-code", "figures", "scatterplot.png") 
-#ggsave(filename = combineplot_location, plot = combined_plot, width = 16, height = 12, units = "in", dpi = 300)
-
-# Display the plot
-plot(combined_plot)
-
-```
+***Correlation matrix***
 
@@
 # Compute Spearman correlation matrix
-cor_matrix_spearman <- cor(asia_africa[, c("tbi","gdp", "hdi", "pro_cleanfuel", "pro_safewater", 
-                                           "pro_basicsanitation", "pro_safesanitation", "mean_exposure_pm2.5", "total_greenhouses")], method = "spearman", use = "complete.obs")
+cor_matrix_spearman <- cor(seaa_df[, c("tbi", "hdi", "pro_safewater", "pro_basicwater", 
+                                           "pro_basicsanitation", "pro_safesanitation")], method = "spearman", use = "complete.obs") %>%
+  print()
 
@@
 # Plot heatmap
-plot5 <- ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
+plot_cor_matrix <- ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
   geom_tile() +
@@
   labs(title = "Spearman Correlation Heatmap", x = "", y = "")
-matrix_location <- here( "code","eda-code", "figures", "correlation_matrix.png") # to set up location for the pictures created 
-ggsave(filename = matrix_location, plot=plot5, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
-plot(plot5)
+matrix_location <- here("results", "figures", "eda", "correlation_matrix.png") # to set up location for the pictures created 
+ggsave(filename = matrix_location, plot=plot_cor_matrix, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
+plot(plot_cor_matrix)
 ```
@@
 #| echo: FALSE
-knitr::include_graphics(here("code", "eda-code", "figures", "correlation_matrix.png"))
+knitr::include_graphics(here("results", "figures","eda",  "correlation_matrix.png"))
 ```
@@
 ```{r}
-library(car)
-#vif(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
-       pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, data = pdata)) # The error might be occure due to multicolinearity
 
-#alias(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
+vif(lm(tbi ~ hdi + pro_safewater +pro_basicwater + pro_basicsanitation + pro_safesanitation, data = seaa_df)) # The error might be occure due to multicolinearity
+
+alias(lm(tbi ~ gdp + hdi + pro_cleanfuel + pro_safewater + pro_basicsanitation + 
          pro_safesanitation + mean_exposure_pm2.5 + total_greenhouses, data = asia_africa))
@@
 
-Interpretation: ***1. variance components Analyisis***
+Interpretation: 
+***1. variance components Analyisis***
 

### 2025-04-08 code/processing-code/processingfile.qmd
Edit manuscript

@@
 
-This file contains processing data for the Final Project titlled " The effects of multi environmental exposures on Tuberculosis Incidence in Asian and African countris".
+This file contains processing data for the Final Project titlled " The effects of water and sanitation on Tuberculosis Incidence in South East Asian and African countris".
 
-***Variables in the study*** - Tuberculosis Incidence (Response Variable)
+***Variables in the study*** 
+- Tuberculosis Incidence (Response Variable)
 
-Independent Variables - Basic water - Safety Water - Greenhouse - PM2.5 - Basic Sanitation - Safe Sanitation - Slum - Fule_cooking
-
-Control variables: - Human Development Index - Health Expenditure - GDP
+- Basic water, Safety Water, Basic Sanitation, Safe Sanitation, Human Development Index (Independent Variables )
 
@@
 library(ggplot2)
+install.packages("viridis")
+library(viridisLite)
+
+install.packages("tidyverse")
+library(tidyverse)
+install.packages("rsample") 
+library(rsample)
 
@@
 tbi <- read.csv(tbi_location)
-
+```
+```{r}
 colnames(tbi)
@@
 
-***Fuel for Cooking***
-
-```{r}
-cookingfuel_location <- here::here("data","raw-data","fule_cooking.xlsx")
-cookingfuel <- read_excel (cookingfuel_location )
-
-summary(cookingfuel)
-skimr::skim(cookingfuel)
-colnames(cookingfuel)
-```
-
-Creating long table to make the table shape consitent to allow us to combine tables in the next steps
-
-```{r}
-cookingfuel <- cookingfuel %>% 
-  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns
-
-colnames(cookingfuel) <- gsub("\\[.*\\]", "", colnames(cookingfuel))  # to delete [YR.... ] in the columns
-
-cookingfuel <- cookingfuel %>%
-  rename(country = `Country Name`)
-
-# Reshape from wide to long format
- cookingfuel <- cookingfuel %>%
-  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
-               names_to = "year", 
-               values_to = "pro_cleanfuel") %>%
-  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
-  filter(year != 2023)  # Remove rows where year is 2023
-
- saveRDS(cookingfuel, here("data", "processed-data", "processed-cookingfuel.rds"))
-```
-
 ***Basic Sanitation***
@@
 
-***Slum Area***
-
-```{r}
-slum_location <- here::here("data","raw-data","slum.xlsx")
-slum <- read_excel (slum_location)
-
-summary(slum)
-skimr::skim(slum)
-```
-
-```{r}
-slum <- slum %>% 
-  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns
-
-colnames(slum) <- gsub("\\[.*\\]", "", colnames(slum))  # to delete [YR.... ] in the columns
-
-slum <- slum %>%
-  rename(country = `Country Name`)
-
-# Reshape from wide to long format
- slum <- slum %>%
-  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
-               names_to = "year", 
-               values_to = "pro_slum") %>%
-  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
-  filter(year != 2023)  # Remove rows where year is 2023
-
-saveRDS(slum, here("data", "processed-data", "processed-slum.rds"))
-```
-
-***Greenhouses***
-
-```{r}
-greenhouses_location <- here::here("data","raw-data","greenhouse.xlsx")
-greenhouses <- read_excel (greenhouses_location)
-
-summary(greenhouses)
-skimr::skim(greenhouses)
-```
-
-```{r}
-greenhouses <- greenhouses %>% 
-  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns
-
-colnames(greenhouses) <- gsub("\\[.*\\]", "", colnames(greenhouses))  # to delete [YR.... ] in the columns
-
-greenhouses <- greenhouses %>%
-  rename(country = `Country Name`)
-
-# Reshape from wide to long format
- greenhouses <- greenhouses %>%
-  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
-               names_to = "year", 
-               values_to = "total_greenhouses") %>% # Total greenhouse gas in million metric cubic ton including emissions and removals from Land Use, Land-Use Change, and Forestry (LULUCF)
-  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
-  filter(year != 2023)  # Remove rows where year is 2023
-
-saveRDS(greenhouses, here("data", "processed-data", "processed-greenhouses.rds"))
-```
-
-***PM2.5***
-
-```{r}
-pm25_location <- here::here("data","raw-data","pm25.xlsx")
-pm25 <- read_excel (pm25_location)
-
-summary(pm25)
-skimr::skim(pm25)
-```
-
-```{r}
-pm25 <- pm25 %>% 
-  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns
-
-colnames(pm25) <- gsub("\\[.*\\]", "", colnames(pm25))  # to delete [YR.... ] in the columns
-
-pm25 <- pm25 %>%
-  rename(country = `Country Name`)
-
-# Reshape from wide to long format
- pm25 <- pm25 %>%
-  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
-               names_to = "year", 
-               values_to = "mean_exposure_pm2.5") %>% #PM2.5 air pollution, mean annual exposure (micrograms per cubic meter)
-  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
-  filter(year != 2023)  # Remove rows where year is 2023
-
- saveRDS(pm25, here("data", "processed-data", "processed-pm25.rds"))
-```
-
-***GDP***
-
-```{r}
-gdp_location <- here::here("data","raw-data","gdp.csv")
-gdp <- read.csv (gdp_location)
-
-summary(gdp)
-skimr::skim(gdp)
-```
-
-```{r}
-gdp <- gdp %>% 
-  select(-`Series.Code`, , -`Series.Name`, -`Country.Code`)# delete potention confusing columns
-
-colnames(gdp) <-gsub("X(\\d+)..YR\\d+.", "\\1", colnames(gdp))  # to delete [YR.... ] in the columns 
-gdp <- gdp %>%
-  rename(country = Country.Name)
-
-# Reshape from wide to long format
- gdp <- gdp %>%
-  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
-               names_to = "year", 
-               values_to = "gdp") %>% #US Dollar 
-  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
-  filter(year != 2023)  # Remove rows where year is 2023
-
-saveRDS(gdp, here("data", "processed-data", "processed-gdp.rds"))
-```
-
 ***Human Development Index***
@@
 #create list of dataset
-datasets <- list(tbi, cookingfuel, waterbasic, safewater, slum, pm25, greenhouses, basicsanitation, safesanitation, gdp, hdi)
+datasets <- list(tbi,  waterbasic, safewater,  basicsanitation, safesanitation, hdi)
 
@@
 combined_df <- reduce(datasets, full_join, by = c("country", "year"))
+
+saveRDS(hdi, here("data", "processed-data", "combined-data.rds"))
 ```
@@
 
-### Exluding slum dataset
-
-Slum dataset is different from other dataset, the observation is every 2 years.
-I want to create subset without slum data
-
-```{r}
-no_slum <- combined_df %>% 
-  select(-pro_slum)
-
-```
-
 Now, I want to drop all NA observation
@@
 
-no_slum <- no_slum %>%
+combined_df <- combined_df %>%
   filter(if_all(everything(), ~ !is.na(.) & . != "" & . != ".."))
@@
 # Count the number of rows for each country
-complete_countries <- no_slum %>%
+complete_countries <- combined_df %>%
   group_by(country) %>%
@@
 # Keep only those countries in the dataset
-clean_noslum <- no_slum %>%
+combined_df <- combined_df %>%
   filter(country %in% complete_countries)
@@
 # Print first few rows to check
-head(clean_noslum)
+head(combined_df)
 
-table(clean_noslum$year, clean_noslum$country) # to make sure that all countries have complete year 
-dim(clean_noslum)
+table(combined_df$year, combined_df$country) # to make sure that all countries have complete year 
+dim(combined_df)
 
-unique(clean_noslum$country)
+unique(combined_df$country)
 
@@
 ```{r}
-clean_noslum <- clean_noslum %>%
-  mutate(across(c(pro_cleanfuel, pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, gdp), as.numeric))
-
-summary(clean_noslum)
-
-saveRDS(clean_noslum, here("data", "processed-data", "clean_noslum.rds")) # save the file
+write.csv(combined_df, here("data", "processed-data", "combined-df.csv"), row.names = FALSE)
 
@@
 
-### Creating clean dataset inlcuding slum
+***Only including countries from Couth East Asia and African COuntries***
 
-Note: Slum dataset only contain observation every two years. In this dataset, I will follow the year where slum has observation. 
 ```{r}
-# Define the required years
-required_years_alldata <- c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020)
+asia_africa <- combined_df %>% 
+  filter(region %in% c("South-East Asia", "Africa"))
 
-# Filter the dataset for the required years
-all_data <- combined_df %>%
-  filter(year %in% required_years_alldata) %>%
-  group_by(country) %>%
-  filter(all(required_years_alldata %in% year)) %>%
-  ungroup()
+dim(asia_africa)
 
-all_data  <- all_data  %>%
-  filter(year %in% required_years_alldata) %>%  # Keep only required years
-  group_by(country) %>%
-  filter(n_distinct(year) == length(required_years_alldata)) %>%  # Ensure all years exist for each country
-  ungroup()
+unique(asia_africa$region)
 
-all_data <- all_data %>%
-  mutate(across(where(is.character), ~ na_if(.x, ".."))) %>%  # Convert ".." to NA only for character columns
-  drop_na()  # Remove rows with any NA values
+# save the asia africa dataset
 
-# View the filtered data
-print(all_data )
+write.csv(combined_df, here("data", "processed-data", "asia_africa.csv"), row.names = FALSE)
+saveRDS(combined_df, here("data", "processed-data", "asia_africa.rds"))
+```
 
-unique(all_data$country)
-# Optionally, save the filtered data to a new CSV file
-#write.csv(filtered_hdi, "filtered_hdi_selected_years.csv", row.names = FALSE)
+***Visualize TBI*** 
+```{r}
+
+# Line plot of TBI over time by country
+ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
+  geom_line(alpha = 0.7) +
+  labs(title = "TBI Trends in Asia and Africa (2000–2020)",
+       x = "Year",
+       y = "TBI") +
+  theme_minimal() +
+  theme(legend.position = "none")  # Hide legend if there are many countries
 
@@
 
+It can be seen that Lesotho has very high observation and potentially become outlier. It is better to drop the observation.
+
 ```{r}
-ggplot(all_data, aes(x = year, y = tbi, group = country, color = country)) +
-  geom_line(alpha = 0.7) +  # Transparency to see overlapping lines
-  labs(title = "Spaghetti Plot of TBC Incidence",
-       x = "Year",
-       y = "TBC Incidence") +
-  theme_minimal() +
-  theme(legend.position = "none") 
+seaa <- asia_africa%>%
+  filter(country != "Lesotho")
+
+#save the final dataset 
+write.csv(seaa, here("data", "processed-data", "seaa.csv"), row.names = FALSE)
+saveRDS(seaa, here("data", "processed-data", "seaa.rds"))
 ```
@@
 ```{r}
-# convert variables into numeric
-
-all_data <- all_data %>%
-  mutate(across(c(pro_cleanfuel, pro_basicwater, pro_safewater, pro_slum, pro_basicsanitation, pro_safesanitation, mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, gdp), as.numeric))
-summary(all_data)
-str(all_data)
-
-saveRDS(all_data, here("data", "processed-data", "clean_data_withslum.rds")) # save the file 
-write.csv(all_data, here("data", "processed-data", "clean_data_withslum.csv"), row.names = FALSE)
-unique(all_data$country)
+# Line plot of TBI over time by country
+ggplot(seaa, aes(x = year, y = tbi, group = country, color = country)) +
+  geom_line(alpha = 0.7) +
+  labs(title = "TBI Trends in Asia and Africa (2000–2020)",
+       x = "Year",
+       y = "TBI") +
+  theme_minimal() +
+  theme(legend.position = "none")  # Hide legend if there are many countries
 ```
@@
 ```{r}
-all_asiaafrica <- all_data %>% 
-  filter(region %in% c("South-East Asia", "Africa"))
 
-dim(all_asiaafrica)
+ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
+  geom_line(alpha = 0.9) +
+  facet_wrap(~ region) +
+  scale_color_viridis_d(option = "turbo") +  # "plasma", "cividis", or "turbo" are good options
+  labs(title = "TBI Trends (2000–2020) by Region",
+       x = "Year",
+       y = "TBI",
+       color = "Country") +
+  theme_minimal() +
+  theme(legend.position = "right",
+        legend.key.height = unit(0.5, "cm"),
+        legend.text = element_text(size = 8))
 
-unique(all_asiaafrica$country)
-
-m2 <- lm(log(tbi) ~ pro_slum, data = all_asiaafrica)
-summary(m2)
-
-plot(m2)
 ```
 
-```{r}
-library(reshape2)  # To reshape data for ggplot
-
-# Compute Spearman correlation matrix
-cor_matrix_spearman_slum <- cor(all_asiaafrica[, c("tbi","gdp", "hdi", "pro_cleanfuel", "pro_safewater", 
-                                           "pro_basicsanitation", "pro_safesanitation", "pro_slum")], 
-                           method = "spearman", use = "complete.obs")
-
-# Melt the matrix into long format for ggplot
-cor_melted_slum <- melt(cor_matrix_spearman_slum)
-
-# Plot heatmap
-ggplot(cor_melted_slum, aes(Var1, Var2, fill = value)) +
-  geom_tile() +
-  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
-  geom_text(aes(label = round(value, 2)), size = 4) +  # Show correlation values
-  theme_minimal() +
-  labs(title = "Spearman Correlation Heatmap", x = "", y = "")
-```
+# Machine Learning Data 
+I will create machine learning data
 
 ```{r}
-ggplot(all_asiaafrica, aes(x = year, y = pro_slum, group = country)) +
-  geom_line() +
-  facet_wrap(~ country) +  # Creates separate plots for each country
-  theme_minimal()
+# create setseed for reproductibility 
+ml_seed <- set.seed(2345)
+
+# Split dataset into a 75% train set and a 25% test set
+
+splitted_ml <- initial_split(asia_africa, prop= .75)
+df_train <- training(splitted_ml)
+df_test <- testing(splitted_ml)
+
+# Save training set
+saveRDS(df_train, file = here("data", "processed-data", "df_train.rds"))
+
+# Save test set
+saveRDS(df_test, file = here("data", "processed-data", "df_test.rds"))
+
 ```

### 2025-04-08 products/manuscript/Manuscript.qmd
Edit manuscript

@@
 relationship between Tuberculosis Incidence and Predictors. Non linear
-relationship was discovered.Next step is to perform non-linear
-relationship to looking at form of relationship. Human Development Index
+relationship was discovered. Human Development Index
 (HDI), Access to basic water and well-managed water show curvature
 relationship, while access to basic sanitation and well-managed
-sanitation show polynomial. Data transformation was performed based.
-After data transformation, we fit model1, which only includes water and
-sanitation variables (as main variables in this study), model2 includes
-all predictors (HDI, access to water and sanitation), model3 only
-includes significant variables in the previous model. Model 1, 2 and 3
-use Generalized Linear Mixed Model (GLMM). We also performed VIF test to
-look at multicollinearity. Model 4 uses Generalized Additive Mixed
-Models. Moreover, BIC and AIC tests were performed to look at the best
-model. BIC and AIC tests found that Model 4 (GAMM) performs better with
-the smallest BIC and AIC. To be able to interpret GAMM better, we
-visualize the result into plots.
+sanitation show polynomial.
+There are several models fitted in the study, we start with linear model with all predictors included, followed by Generalized Addictive Models (GAM), Generalized Mixed Addictive Models (GAMM) without interaction, Generalized Mixed Addictive Models (GAMM) with two factor interaction, and Generalized Mixed Addictive Models (GAMM) with two factor interaction after drop interaction which are not significant.
+After fitting the models, RMSE and AIC were printed to chose the best models.
+Based on the RMSEs and AICs, GAMM with interaction after deleting unsigniuficant interactions performed best.
+The RMSE and AIC were visulized using plot for better understanding.
+
+To maximize test the performance of the model, machine learning was performed.
+We created splitted data with 75% train data and 25 % test data and setseed (2345).
+RMSEs and AICs were printed to see the performance of the models and plotted.
+The final model was fitted into a machine learning, XGBoost.
+The method is very plexible and can deal with the data conditions (non-linear relationship, some multicolinearity, and binomial distribution).
+Finally, the model was tunned.
 
@@
 
-There are some exploratory data analysis conducted and resulted table,
-and charts. Table 1 shows the summary statistics including min, mean,
-max, and standard deviation of all variables.
-
-NOTE: I still have issue in saving the tables. I am still working on it.
-I will show it next step.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS (here("code", "eda-code", "tables", "summary_table.rds"))
-knitr::kable(resulttable)
-```
-
-The line graph ( @fig-1 ) shows the trend of all variables over the
-period in each country. It can be seen that the majority of country
-showed the decreases trend in tuberculosis incidence over 21 years
-period. All countries showed the increase trend in percentage of
-population using using basic and safe drinking water, basic and safe
-sanitation, except Zimbabuwe showing the decrease in percentage of
-population using basic drinking water and safe sanitation.
+Firs of all, the pattern of the data was explored.
+The trend of Tuberculosis Incidence for each country during the period was assesed using spagethi plots.
+The ( @fig-1 ) shows that most countries shows a dicrease of tuberculosis incidence over the period of time.
 
@@
 #| label: fig-1
-#| fig-cap: "Trends of all variable over the period in each country"
+#| fig-cap: "Trends of Tuberculosis of each country from 2000 to 2020"
 #| echo: FALSE
-knitr::include_graphics(here("code","eda-code","figures","line-graph-tbivsfactors.png"))
-
+knitr::include_graphics(here("results", "figures", "eda", "tbi_spageti.png"))
 ```
 
-( @fig-2 ) shows spaghetti plots in each variables. We can see the
-trends of each country in every variables. It can be seen that generally
-most of the countries show the increase trends in water and sanitation,
-and human development index
+Figure ( @fig-2 ) shows the trend of percentage of population using safe water, basic water, safe sanitation, basic sanitation, tuberculosis incidence, and human development index over the prerion for each country.
+It can be seen that the percentage of population having access to water and sanitation and human development index increase over the time, while the tuberculosis incidence decrease.
+Figure ( @fig-3 ) show the trends of all variables based on region.
 
 ```{r}
-
 #| label: fig-2
-#| fig-cap: "Spaghetti plots in each variables"
+#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
 #| echo: FALSE
-knitr::include_graphics(here("code","eda-code","figures","spageti.png"))
-
+knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars.png"))
 ```
 
-Now lets take a more detail tuberculosis incidence in each country over
-the year using the following box plot. Lesotho leads the incidence
-followed by Myanmar and Tanzania.
-
 ```{r}
-
 #| label: fig-3
-#| fig-cap: "Boxplot plots for each Country"
+#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
 #| echo: FALSE
-knitr::include_graphics(here("code","eda-code","figures","boxplot.png"))
-
+knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars_region.png"))
 ```
@@
 
-As first step of statistical analysis. Using scatter plot to show the
-linearity between tuberculosis incidence and predictors. ( @fig-4 )
-shows the scatter plots. The scatter plots do not show linear
-relationship between tuberculosis Incidence and predictors.
+As first step of statistical analysis.
+linearity test was performed, the result shows that there is non-linear relationship between tuberculosis incidence and other predictors.
+Loess method is used to show the non-linear relationship betwen tuberculosis incidence and predictors.
+( @fig-4 ) shows the relationship between tuberculosis incidence and predictors and ( @fig-5 ) show the relationhsip after data transformation.
+It can be seen that after data transformed, non-linear relationship stil exists.
+Therefore, Generalized Addittive Mixed Model is used in this study.
 
@@
 #| label: fig-4
-#| fig-cap: "Scatter plots between Tuberculosis Incidence and Predictors "
+#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors "
 #| echo: FALSE
-knitr::include_graphics(here("code","eda-code","figures","scatterplot.png"))
+knitr::include_graphics(here("results","figures", "analysis", "plots_tbi_vs_allvars.png"))
 
+```
+```{r}
+#| label: fig-5
+#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors after transfirmation "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures", "analysis", "plot_transformed.png"))
 ```
@@
 tuberculosis Incidence and all predictors.
-
-Since they show non linearity, spearman's correlation performed and
-showing in the following correlation matrix. See ( @fig-5 )
-
-```{r}
-
-#| label: fig-5
-#| fig-cap: "Spearman Correlation Matrix "
-#| echo: FALSE
-knitr::include_graphics(here("code","eda-code","figures","correlation_matrix.png"))
-
-```
-
-Now, we want to assess non linear relationship.
+Since they show non linearity, spearman's correlation performed and showing in the following correlation matrix.
+See ( @fig-6 ) shows  that some variables have multi collinearity.
+The matrix shows that water and sanitation have low correlation with tuberculosis incidence.
 
@@
 #| label: fig-6
-#| fig-cap: "Nonlinear relationship"
+#| fig-cap: "Spearman Correlation Matrix "
 #| echo: FALSE
-knitr::include_graphics(here("code","analysis-code","figures","nonlinear.png"))
-
-```
-
-After performing nonlinear model, data transformation used based on the
-shave or distribution of the plots from nonlinear relationship plots (
-@fig-6 ) above. ( @fig-7 ) shows the data distribution after data
-transformation.
-
-```{r}
-
-#| label: fig-7
-#| fig-cap: "transformed data plots"
-#| echo: FALSE
-knitr::include_graphics(here("code","analysis-code","figures","transformed.png"))
+knitr::include_graphics(here("results","figures", "eda", "correlation_matrix.png"))
 
@@
 
-There are four models build; fitting model with only water and
-sanitation variables, then fitting models including co-variate (HDI).
-The third model is to fit only significant variables. The first three
-models use Generalized Linear Mixed Models (GLMM), while the last model
-used Generalized Additive Mixed Models (GAMM). Those models were
-compared using AIC and BIC criteria to select the best/ the final
-models.
+We start with linear model with all predictors included, followed by Generalized Addictive Models (GAM), Generalized Mixed Addictive Models (GAMM) without interaction, Generalized Mixed Addictive Models (GAMM) with two factor interaction, and Generalized Mixed Addictive Models (GAMM) with two factor interaction after drop interaction which are not significant.
+AIC and RMSE are used to select the best model.
+( @fig-7) show that  and Generalized Mixed Addictive Models (GAMM) with two factor interaction after dropin insignificant interaction performed better.
+GAMM with full interactions has low AIC and RMSE, but the degree of freedom shows negarive values, indicating inconsistency of the model.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Model Criteria"
+
+#| label: fig-7
+#| fig-cap: "AIC and RMSE "
 #| echo: FALSE
-resulttable2=readRDS (here("code", "analysis-code", "tables", "criteria_output.rds"))
-knitr::kable(resulttable2)
+knitr::include_graphics(here("results","figures", "analysis", "plot_aic_rmse.png"))
 ```
 
-Based on BIC and AIC criteria, the best model fit the data is
-Generalized Additive Mixed Models.
+***Result of Final Model***
 
-\$ \log(\text{TBI}) \sim s(\text{hdi}) + s(\text{pro\_basicwater}) +
-s(\text{pro\_safewater}) + s(\text{pro\_basicsanitation}) +
-s(\text{country}, \text{bs}="re") \$
+We fitted a generalized additive mixed model (GAMM) using a negative binomial distribution with a log link to model tuberculosis incidence (TBI) across countries in South East Asia and Africa.
+The model incorporated smooth functions of year, access to basic and safe water and sanitation services, and the Human Development Index (HDI), as well as interaction terms between key environmental and socioeconomic indicators.
+A random effect was included for country to account for unobserved heterogeneity.
+(@tbl-1 ) shows the result of the final model.
+```{r}
+#| label: tbl-1
+#| tbl-cap: "Summary of GAMM Parametric and Smooth Terms"
+#| echo: false
+#| message: false
+#| warning: false
+
+# Load and display the saved gt table
+gamm_tbl <- readRDS(here::here("results", "tables", "analysis", "gamm_summary_table.rds"))
+
+gamm_tbl %>%
+  gt(groupname_col = "type") %>%
+  fmt_number(columns = where(is.numeric), decimals = 3) %>%
+  tab_header(
+    title = "Summary of GAMM Estimates",
+    subtitle = "Parametric and Smooth Terms"
+  ) %>%
+  cols_label(
+    term = "Term", estimate = "Estimate", std.error = "Std. Error",
+    edf = "EDF", chi_sq = "Chi-square", p.value = "p-value"
+  ) %>%
+  fmt_missing(everything(), missing_text = "-")
+
+```
+
+The model demonstrated excellent fit, explaining 99.7% of the deviance (adjusted R² = 0.997).
+Among the smooth terms, year (edf = 3.38, p < 0.001), access to basic water (edf = 6.52, p = 0.013), access to safe water (edf = 1.00, p < 0.001), and HDI (edf = 1.00, p < 0.001) showed significant non-linear associations with TBI.
+In contrast, basic and safe sanitation services were not significantly associated with TBI after accounting for other variables.
+
+Notably, several two-dimensional interaction smooths were also statistically significant, including between basic water and HDI (p < 0.001), safe water and HDI (p < 0.001), safe sanitation and basic water (p < 0.001), and safe water and basic water (p < 0.001), suggesting complex joint effects of environmental infrastructure and development on TBI patterns.
+The random effect for country was highly significant, capturing substantial variation across national contexts.
+
+These findings underscore the importance of improving access to clean water and addressing socioeconomic disparities to reduce tuberculosis burden, while also highlighting the complex interplay between infrastructure and development indicators.
+
+### Machine Learning
+
+To test the consistency of the model, machine learning was used.
+
+Using Splitted / generated data shows that the model has a consistent result.
+(@fig-8 ) shows the observed and predicted values both original and generated data.
+It can be seen that both original data and generated data (both test and training data) have similiar pattern and the plots lay very closed or around the line.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Generalized Additive Mixed Models"
+#| label: fig-8
+#| fig-cap: "Predicted vs Observed TBI originalvs generated data "
 #| echo: FALSE
-resulttable3=readRDS (here("code", "analysis-code", "tables", "summary_Gamm.rds"))
-knitr::kable(resulttable3)
+knitr::include_graphics(here("results","figures", "analysis", "plot_ob_pre_ml_vs_ori.png"))
 ```
 
-This study employed a Generalized Additive Model (GAM) with a Gaussian
-family and identity link function to examine the association between
-tuberculosis incidence and key environmental and socioeconomic factors
-across multiple countries. The model incorporated smooth functions for
-continuous predictors and a random effect for country to account for
-within-country variability. The results indicate that HDI and basic
-sanitation are significant determinants of tuberculosis incidence, while
-access to basic and safe water did not show strong associations.
+Figure (@fig-9) shows the performance of machine learning (XGboost).
+The figure shows how well the model predicted tuberculosis incidence (TBI) by comparing the predicted values with the actual reported values for both the training and test data.
+The dotted line represents perfect prediction—where predicted and actual values would match exactly.
 
-The smooth function for Human Development Index (HDI) was highly
-significant (p \< 2e-16), with an estimated effective degrees of freedom
-(EDF) = 4.283, confirming a non-linear relationship between HDI and
-tuberculosis incidence. The effect was strongest at lower HDI values,
-suggesting that improvements in socioeconomic conditions may have the
-greatest impact in reducing tuberculosis burden in low-HDI countries.
+For the training data (right side), the dots are closely lined up along the diagonal line, which means the model predicted the TBI values very accurately for the data it was trained on.
 
-Among the water and sanitation variables, basic water access was not
-statistically significant (p = 0.73379), indicating no strong evidence
-of an association between water coverage and tuberculosis incidence.
-Similarly, safe water access was not significant at α = 0.05 (p =
-0.05605), suggesting only a weak relationship. However, basic sanitation
-access was statistically significant (p = 0.00542, EDF = 3.845),
-suggesting that higher access to sanitation is associated with a
-reduction in tuberculosis incidence. This highlights the importance of
-improving sanitation infrastructure as a potential intervention for
-tuberculosis control.
+For the test data (left side), most of the points are still close to the line, but some are spread out more, especially at higher TBI values.
+This means that while the model still performs well on new data, it slightly underestimates TBI in some cases when the actual rates are very high.
 
-The random effect of country (s(country)) was highly significant (p \<
-2e-16, EDF = 16.768), indicating substantial between-country variation
-in tuberculosis incidence. This justifies the inclusion of random
-effects to account for country-level heterogeneity and underscores the
-need for context-specific policy interventions tailored to national
-conditions.
-
-The model fit was strong, with an adjusted R² of 0.974, indicating that
-97.4% of the variance in tuberculosis incidence was explained by the
-model. The scale estimate was 0.014598, suggesting a low residual
-variance, which reflects a well-fitting model.
+Overall, the plot shows that the model is reliable and provides accurate predictions for TBI, even on data it hasn’t seen before.
+This suggests the model can be useful for forecasting TB burden in similar settings.
 
 ```{r}
-#| label: fig-7
-#| fig-cap: "Generalized Additive Mixed Models"
+#| label: fig-9
+#| fig-cap: "Predicted and observed value for Xgboost using train and test data"
 #| echo: FALSE
-knitr::include_graphics(here("code","analysis-code","figures","gam_diagnostics.jpg"))
+knitr::include_graphics(here("results","figures", "analysis", "plot_obs_pred_ml.png"))
+```
 
+***Tunning for XGBoost***
+
+```{r}
+#| label: fig-10
+#| fig-cap: "Tunning Result for XGBoost"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures", "analysis", "autoplot_xgb_tuning.png"))
 ```
+
+Figure (@fig-10) shows the results of hyperparameter tuning for the XGBoost model, showing how different parameter values affected the model’s performance, measured by root mean square error (RMSE).
+Each dot represents one combination of hyperparameters tested during cross-validation.
+These results highlight that careful tuning of parameters, especially learning rate, tree depth, and sampling rate, is essential to significantly improve model accuracy.
+The tuning process allowed us to identify a combination of settings that minimized prediction error and led to a more reliable model.
 
@@
 *Summarize what you did, what you found and what it means.*
+
+I am still working on it and read more paper to use in this discussion part. I will complete this part this week
 

### 2025-04-08 code/analysis-code/statistical-analysis.R

@@ -4,53 +4,3 @@
 #this script loads the processed, cleaned data, does a simple analysis
 #and saves the results to the results folder
 
-#load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
-
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
-
-#load data. 
-mydata <- readRDS(data_location)
-
-
-######################################
-#Data fitting/statistical analysis
-######################################
-
-############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
-
-lmfit1 <- lm(Height ~ Weight, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
-
-#look at fit results
-print(lmtable1)
-
-# save fit results table  
-table_file1 = here("results", "tables", "resulttable1.rds")
-saveRDS(lmtable1, file = table_file1)
-
-############################
-#### Second model fit
-# fit linear model using height as outcome, weight and gender as predictor
-
-lmfit2 <- lm(Height ~ Weight + Gender, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable2 <- broom::tidy(lmfit2)
-
-#look at fit results
-print(lmtable2)
-
-# save fit results table  
-table_file2 = here("results", "tables", "resulttable2.rds")
-saveRDS(lmtable2, file = table_file2)
-
-  

### 2025-04-17 products/manuscript/Manuscript.qmd
edit
Fix citation and reference lists

@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+bibliography: assets/dataanalysis-references.bib
+csl: assets/american-journal-of-epidemiology.csl
 editor: 
@@
 library(knitr)
+install.packages("dplyr")
+library(dplyr)
+install.packages("gt")
+library(gt)
 ```
@@
 2.  Faculty of Public Health, Universitas Tadulako, Palu, Indonesia.
-3.  Yet another Affiliation.
 
@@
 
-$\land$ Corresponding author: some\@email.com
+$\land$ Corresponding author: some\@muhammad.nasir@uga.edu
 
@@
 
-Tuberculosis is old infectious disease which still exists and become
-major public health in the world, particularly in developing countries.
-Tuberculosis is linked to environmental factors. This study will analyze
-the effects of multi environmental factors, specially water and
-sanitation on tuberculosis incidence in Asia and Africa regions, where
-the majority of tuberculosis coming from these regions. The water and
-sanitation will be the main exposures and Human Development Index as a
-control variable.
+Tuberculosis is an old infectious disease which still exists and become major public health in the world, particularly in developing countries.
+In 2023, World Health Organization (WHO) estimated 10.8 million people infected by TB globally with (95% uncertainty interval 10.1-11.7 million), which increased from 10.7 million in 2022.
+The number is equivalent to 134 incident cases per 100 000 population.
+45% and 24% of the cases were detected in South-East Asia and Africa repectively.[@who_data] Tuberculosis has been associated with socioeconomic status, housing conditions, nutrition and multiple environmental factors.
+Climate Change Phenomena also is suspected to affect tubercolusos both direct and indirect such as temperature, precipitation and humidity change [@Tosepu2024] and worseing air polution [@Andersen2023], agent survival[@Uwishema2023; @ElSayed2020], water resources destruption, crop destruption, and economic destruption, and nutrition [@Kharwadkar2022; @Hong2025; @Mirzabaev2023; @microbes2022; @Cai2021].
+In term of Environmental factors on Tuberculosis, most studies examined climatic indicators and air poluttion.
+There is alimited studies examine the effect of Water and Sanitation on tubercolusis.
+Althought Tuberculosis is an airborne diseases, lack of access to water and sanitation probably can affect the transmission of tubercolosis because it relates to personal hygiene, home cleaningness, and other infectious diseases and immune system.
+Therefore, the study aims to explore the effect of water and sanitation on Tuberculosis Incidence Rate in regions, which are major contributors on tuberculosis cases globally namely South-East Asia and Africa.
+The study also include Human Development Index in the analysis as a control variable.
 
@@
 RQ: Does access to water and sanitation effect Tuberculosis incidence in
-Asian and African Regions?
+SOuth-East Asia and African Regions?
 
@@
 tuberculosis incidence
-
-To cite other work (important everywhere, but likely happens first in
-introduction), make sure your references are in the bibtex file
-specified in the YAML header above and have the right bibtex key. Then
-you can include like this:
-
-Examples of reproducible research projects can for instance be found in
-[@mckay2020; @mckay2020a].
 
@@
 
-## Schematic of workflow
-
-NOTE: I AM STILL WORKING TO FIGUR THIS ONE OUT Sometimes you might want
-to show a schematic diagram/figure that was not created with code (if
-you can do it with code, do it). @fig-schematic is an example of some -
-completely random/unrelated - schematic that was generated with
-Biorender. We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
-
 ## Data acquisition
 
-The data was downloaded from Worlbank Dataset. However, those data are
+The data was downloaded from Worlbank Dataset. Those data are
 originally from many different courses/ organizations, such as World
-Bank, WHA, UNICEF, UN-HABITAT, EDGAR, and IHME.
+Bank, UNICEF, and UN-HABITAT
 
@@
 
-The study applied Generalized Additive Mixed Model (GAMM). The model can
-capture mixed effect of countries on Tuberculosis Incidence with non
-linear relationship. The data analysis started by looking at the
-relationship between Tuberculosis Incidence and Predictors. Non linear
-relationship was discovered. Human Development Index
-(HDI), Access to basic water and well-managed water show curvature
-relationship, while access to basic sanitation and well-managed
-sanitation show polynomial.
+The study applied Generalized Additive Mixed Model (GAMM).
+The model can capture mixed effect of countries on Tuberculosis Incidence with non linear relationship.
+The data analysis started by looking at the
+relationship between Tuberculosis Incidence and Predictors.
+Non linear relationship was discovered.
+Human Development Index (HDI), Access to basic water and well-managed water show curvature relationship, while access to basic sanitation and well-managed sanitation show polynomial.
 There are several models fitted in the study, we start with linear model with all predictors included, followed by Generalized Addictive Models (GAM), Generalized Mixed Addictive Models (GAMM) without interaction, Generalized Mixed Addictive Models (GAMM) with two factor interaction, and Generalized Mixed Addictive Models (GAMM) with two factor interaction after drop interaction which are not significant.
@@
 Firs of all, the pattern of the data was explored.
-The trend of Tuberculosis Incidence for each country during the period was assesed using spagethi plots.
+The trend of Tuberculosis Incidence rate for each country during the period was assesed using spagethi plots.
 The ( @fig-1 ) shows that most countries shows a dicrease of tuberculosis incidence over the period of time.
@@
 As first step of statistical analysis.
-linearity test was performed, the result shows that there is non-linear relationship between tuberculosis incidence and other predictors.
-Loess method is used to show the non-linear relationship betwen tuberculosis incidence and predictors.
+linearity test was performed, the result shows that there is non-linear relationship between tuberculosis incidence rate and other predictors.
+Loess method is used to show the non-linear relationship betwen tuberculosis incidence rate and predictors.
 ( @fig-4 ) shows the relationship between tuberculosis incidence and predictors and ( @fig-5 ) show the relationhsip after data transformation.
@@
 ```
+
 ```{r}
@@
 
-We fitted a generalized additive mixed model (GAMM) using a negative binomial distribution with a log link to model tuberculosis incidence (TBI) across countries in South East Asia and Africa.
+We fitted a generalized additive mixed model (GAMM) using a negative binomial distribution with a log link to model tuberculosis incidence (TBI) rate across countries in South East Asia and Africa.
 The model incorporated smooth functions of year, access to basic and safe water and sanitation services, and the Human Development Index (HDI), as well as interaction terms between key environmental and socioeconomic indicators.
@@
 Among the smooth terms, year (edf = 3.38, p < 0.001), access to basic water (edf = 6.52, p = 0.013), access to safe water (edf = 1.00, p < 0.001), and HDI (edf = 1.00, p < 0.001) showed significant non-linear associations with TBI.
-In contrast, basic and safe sanitation services were not significantly associated with TBI after accounting for other variables.
+In contrast, basic and safe sanitation services were not significantly associated with TBI rate after accounting for other variables.
 
@@
 
-I am still working on it and read more paper to use in this discussion part. I will complete this part this week
+This study highlights the significant impact of environmental infrastructure and socioeconomic development on tuberculosis incidence (TBI) in South East Asia and Africa.
+The use of a generalized additive mixed model (GAMM) with a negative binomial distribution and log link allowed us to flexibly assess both linear and non-linear associations between predictors and TBI rates, accounting for random variation between countries.
+Our model demonstrated excellent fit, explaining 99.7% of the deviance, and identified multiple significant associations between water access, development indices, and TBI incidence, consistent with and extending prior literature.
+
+The significant non-linear effect of year on TBI incidence underscores the ongoing temporal trends in tuberculosis control, shaped by global health policies and national-level interventions over time.
+Previous studies have documented a steady decline in TB incidence rates globally, albeit at a slower rate than needed to meet elimination targets, particularly in high-burden settings such as sub-Saharan Africa and Southeast Asia [@pai2016tuberculosis; @world2023global].
+Our findings align with these trends and further demonstrate the utility of flexible modeling approaches in capturing such non-linear patterns over extended time frames.
+
+The significant associations of access to basic and safe water with TBI reinforce the importance of environmental determinants in TB control.
+Limited access to clean water can contribute to a cycle of poor hygiene, malnutrition, and vulnerability to infectious diseases, including TB [@esrey1991effects; @pruitt2021water].
+Our identification of complex non-linear relationships suggests that the effects of water access on TB are not constant but vary depending on context, infrastructure level, and interaction with other factors.
+
+Moreover, the Human Development Index (HDI) was significantly associated with TBI rates in a non-linear fashion.
+This is consistent with research indicating that higher levels of human development—including education, life expectancy, and income—are linked to reduced susceptibility to TB through better living conditions, nutrition, and access to healthcare services [@siroka2016assessing; @wilkinson1999inequality].
+These findings reaffirm the critical role of social determinants in TB epidemiology and provide quantitative support for global policy frameworks that integrate development with infectious disease control.
+
+Notably, basic and safe sanitation access did not show a significant independent association with TB in our adjusted model.
+While sanitation is undeniably important for overall public health, its isolated effect on TB incidence may be attenuated once other covariates such as water access and HDI are considered.
+This finding echoes studies showing that while sanitation is crucial for preventing diseases like diarrhea and helminth infections, its direct effect on TB—which is primarily airborne—may be more indirect and mediated by broader health and development pathways [@bartram2010water; @pruss2002estimating].
+
+The interaction terms in our model revealed statistically significant joint effects between several pairs of variables, suggesting that infrastructure and development indicators do not act independently.
+For instance, the interaction between basic water and HDI supports prior findings that the protective effects of environmental infrastructure are magnified in settings with higher socioeconomic development [@lienhardt2001risk; @matos2013tuberculosis].
+Likewise, the interaction between safe water and HDI emphasizes the synergistic value of combining access to safe water with broader improvements in human development.
+
+Interactions between safe sanitation and basic water, as well as between safe water and basic water, also showed significant effects on TBI, implying that layered improvements in environmental infrastructure produce more than additive benefits.
+These patterns align with ecological frameworks emphasizing cumulative and contextual effects in public health interventions [@marmot2008closing; @aagaard2014integrated].
+
+Importantly, the random effects for country were highly significant, indicating unmeasured country-level heterogeneity in TB patterns.
+These may include factors such as political stability, health system capacity, HIV prevalence, urbanization patterns, or population density—all known modifiers of TB burden [@gupta2015addressing; @reid2019building].
+
+Our findings have several policy implications.
+First, they highlight the need for multisectoral strategies that address both environmental conditions and broader socioeconomic development to combat TB effectively.
+Investments in water infrastructure should be accompanied by programs targeting education, nutrition, housing, and healthcare access [@onozaki2015impact; @udwadia2014multidrug].
+Second, the complex interactions observed suggest that context-specific interventions are essential.
+A one-size-fits-all approach may fail to capture the unique synergy between environmental and developmental factors within each country.
 
@@
 
-*Discuss what you perceive as strengths and limitations of your
-analysis.*
+The strength of our study lies in the use of a flexible GAMM framework that accommodates non-linear relationships and interaction effects, offering a nuanced view of TB determinants.
+By including a random effect for country, we account for latent national characteristics that may bias fixed-effect estimates.
+However, our study has limitations.
+The ecological nature of the data precludes individual-level inferences, and potential measurement error in national statistics could affect the reliability of some estimates.
+
+Despite these limitations, this study adds to the growing evidence that TB incidence is deeply embedded in broader developmental and environmental systems.
+Our findings support the WHO End TB Strategy, which emphasizes tackling the social determinants of TB alongside biomedical interventions [@who2015end; @hargreaves2011social].
 
@@
 
-*What are the main take-home messages?*
-
-These findings underscore the significant role of socioeconomic
-conditions and sanitation infrastructure in tuberculosis control. While
-access to basic and safe water was not a significant predictor,
-improvements in sanitation and overall socioeconomic development (HDI)
-were strongly associated with lower tuberculosis incidence. Given the
-significant country-level differences, tuberculosis prevention
-strategies should be tailored to national contexts, emphasizing
-improvements in living standards and sanitation access. These results
-provide valuable insights for policymakers aiming to design targeted
-public health interventions in high-burden settings.
-
-*Include citations in your Rmd file using bibtex, the list of references
-will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers
-published using a fully reproducible setup similar to the one shown in
-this template.
-
-Note that this cited reference will show up at the end of the document,
-the reference formatting is determined by the CSL file specified in the
-YAML header. Many more style files for almost any journal [are
-available](https://www.zotero.org/styles). You also specify the location
-of your bibtex reference file in the YAML. You can call your reference
-file anything you like.
+This study reinforces that tuberculosis incidence in South East Asia and Africa is not solely a biomedical issue but is deeply influenced by environmental infrastructure and socioeconomic development.
+Our findings demonstrate that access to clean water and improvements in human development significantly reduce TB burden, with their effects magnified when considered jointly.
+The non-linear and interactive associations revealed by our model highlight the importance of tailored, multisectoral approaches that integrate public health with infrastructure and development planning.
+To achieve meaningful and sustained reductions in TB, especially in high-burden regions, global health strategies must move beyond treatment alone and address the root structural determinants of disease.
 
@@
 # References
-
-I will include reference for the sourse of data in the next step.

### 2025-04-21 products/manuscript/Manuscript.qmd
Fixing render issues

@@
 ---
-title: "The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countris"
+title: "The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countries"
 author: ""
@@
     highlight-style: github
-bibliography: assets/dataanalysis-references.bib
+execute:
+  echo: false
+bibliography: assets/citations.bib
 csl: assets/american-journal-of-epidemiology.csl
@@
 
-This uses MS Word as output format. [See
-here](https://quarto.org/docs/output-formats/ms-word.html) for more
-information. You can switch to other formats, like html or pdf. See [the
-Quarto documentation](https://quarto.org/) for other formats.
-
 ```{r, echo=FALSE, message=FALSE}
@@
 library(knitr)
-install.packages("dplyr")
+#install.packages("dplyr")
 library(dplyr)
-install.packages("gt")
+#install.packages("gt")
 library(gt)
@@
 -   Muhammad Nasir$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
--   Second Author$^{3, *}$
--   Third Author$^{3, \dagger}$
--   Last Author$^{1, 2, \land}$
 
@@
 
-# Summary/Abstract
+# Abstract
 
-*Write a summary of your project.*
+Tuberculosis (TB) remains a significant global health concern, particularly in low- and middle-income countries.
+In 2023, an estimated 10.8 million people contracted TB, with nearly 70% of cases reported in South-East Asia and Africa.
+While TB is primarily airborne, environmental and socioeconomic factors, including water and sanitation access, may influence transmission by affecting hygiene and immune function.
+This study aims to investigate the impact of access to water and sanitation services on TB incidence across South-East Asia and Africa, incorporating the Human Development Index (HDI) as a control for socioeconomic context.
+A Generalized Additive Mixed Model (GAMM) was applied to longitudinal data from 2000 to 2022.
+The model included smooth terms for year, water and sanitation indicators, and HDI, with country as a random effect.The model showed excellent fit (adjusted R² = 0.997), with significant non-linear effects observed for year, access to basic water (p = 0.013), safe water (p < 0.001), and HDI (p < 0.001).
+Sanitation variables were not independently significant.
+However, key interactions—such as between water access and HDI—were highly significant (p < 0.001), suggesting complex environmental-development dynamics influencing TB risk.In conclusion, expanding water access and addressing development disparities may be critical in reducing TB burden in high-incidence regions.
 
@@
 #| fig-cap: "Trends of Tuberculosis of each country from 2000 to 2020"
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results", "figures", "eda", "tbi_spageti.png"))
+
 ```
@@
 ```{r}
+
 #| label: fig-2
 #| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars.png"))
+
 ```
@@
 ```{r}
+
 #| label: fig-3
 #| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars_region.png"))
+
 ```
@@
 #| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors "
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "analysis", "plots_tbi_vs_allvars.png"))
@@
 ```{r}
+
 #| label: fig-5
 #| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors after transfirmation "
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "analysis", "plot_transformed.png"))
+
 ```
@@
 #| fig-cap: "Spearman Correlation Matrix "
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "eda", "correlation_matrix.png"))
@@
 #| fig-cap: "AIC and RMSE "
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "analysis", "plot_aic_rmse.png"))
+
 ```
@@
 ```{r}
+
 #| label: tbl-1
@@
   ) %>%
-  fmt_missing(everything(), missing_text = "-")
+  sub_missing(everything(), missing_text = "-")
 
@@
 #| fig-cap: "Predicted vs Observed TBI originalvs generated data "
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "analysis", "plot_ob_pre_ml_vs_ori.png"))
@@
 ```{r}
+
 #| label: fig-9
 #| fig-cap: "Predicted and observed value for Xgboost using train and test data"
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "analysis", "plot_obs_pred_ml.png"))
+
 ```
@@
 ```{r}
+
 #| label: fig-10
 #| fig-cap: "Tunning Result for XGBoost"
-#| echo: FALSE
+#| echo: false
 knitr::include_graphics(here("results","figures", "analysis", "autoplot_xgb_tuning.png"))
+
 ```
@@
 First, they highlight the need for multisectoral strategies that address both environmental conditions and broader socioeconomic development to combat TB effectively.
-Investments in water infrastructure should be accompanied by programs targeting education, nutrition, housing, and healthcare access [@onozaki2015impact; @udwadia2014multidrug].
+Investments in water infrastructure should be accompanied by programs targeting education, nutrition, housing, and healthcare access [@onozaki2015impact].
 Second, the complex interactions observed suggest that context-specific interventions are essential.
@@
 The ecological nature of the data precludes individual-level inferences, and potential measurement error in national statistics could affect the reliability of some estimates.
-
-Despite these limitations, this study adds to the growing evidence that TB incidence is deeply embedded in broader developmental and environmental systems.
-Our findings support the WHO End TB Strategy, which emphasizes tackling the social determinants of TB alongside biomedical interventions [@who2015end; @hargreaves2011social].
 

### 2025-04-27 project-review-Murtaza Yaqubi.qmd
Project review by Murtaza Yaqubi

---
title: Project Review Template 
author: Murtaza Yaqubi
date: 04/25/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countries

Name of project author(s): Muhammad Nasir

Name of project reviewer: Murtaza Yaqubi

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The context of the project is described very well. The manuscript provides a comprehensive background, explaining that tuberculosis (TB) remains a major global health issue, particularly in South-East Asia and Africa. It outlines the biological and social determinants of TB and highlights that while environmental factors like climate and air pollution have been studied extensively, the role of water and sanitation is less explored. The project is clearly placed within existing research, referencing multiple relevant studies and reports (including WHO data and other recent papers). The motivation for the project is strong and well-articulated: it aims to fill a notable gap in the literature by investigating how water and sanitation access might influence TB incidence, using updated and longitudinal data. It is also clear what new information the project hopes to provide—namely, a better understanding of environmental and development factors in TB burden.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The research question is explicitly stated: to assess whether access to water and sanitation impacts TB incidence. Clear null and alternative hypotheses are provided for each indicator. The link between the research questions and the data is well explained.

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data sources (World Bank, WHO, UNICEF) are reliable and clearly mentioned. Variables are listed and described, though a slightly more detailed data dictionary would further assist understanding.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Data wrangling is well performed, with careful cleaning, merging, and preparation steps. Exploratory Data Analysis (EDA) includes country-level trends, boxplots of TBI, and multivariable time trends. Some plots, like the correlation heatmap and spaghetti plots of TB incidence, could benefit from deeper interpretation and explicit linking back to the research questions. Tables and datasets can be accessed by making slight modifications to the file path in eda.qmd and edacode.R.

### Summary assessment
*  some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Analysis using Generalized Additive Mixed Models (GAMM) is highly appropriate, capturing non-linear trends and random country effects. Model selection using RMSE and AIC is properly performed. Testing machine learning models like XGBoost to validate results adds robustness. Overall, analysis methods are strong and well-justified.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Results are clearly presented using plots and tables. Figures are generally clean and well-labeled, though occasional improvements to plot titles and clearer faceting could further enhance clarity.

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion is thorough, connecting findings back to hypotheses and literature. Strengths, limitations, and policy implications are addressed thoughtfully. Interpretation of non-significant findings for sanitation is also appropriately cautious.

### Summary assessment
* strong, complete and clear discussion

## Further comments

This project demonstrates excellent understanding of TB determinants and appropriate use of complex modeling techniques.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

The project is very well-structured. Code, results, and data are organized into logical, clearly labeled folders. However, there certain file and folder that are not used and shold be removed.

### Summary assessment
* mostly clear, but some confusing parts (e.g. eda-v2.qmd and supplement folder)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Documentation is generally very strong, with clear explanations in the manuscript and good code comments. Some exploratory sections could use slightly more detailed annotations for clarity.

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

The project is fully reproducible. Paths are clearly managed with here(), and minimal manual intervention is required.

### Summary assessment
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The study is thorough, addressing multiple models, interaction effects, and machine learning validation. Alternative approaches and sensitivity checks (e.g., model tuning) are thoughtfully considered.

### Summary assessment
* strong level of thorougness

## Further comments

This is an outstanding project overall. The attention to model selection, reproducibility, and thorough discussion are commendable. Small additions like a fuller codebook and slightly deeper EDA interpretation would elevate it even more.

### 2025-04-30 project-review-Yufei Wu.qmd

---
title: Project Review_Yufei Wu
author: Yufei Wu
date: 04/24/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countries

Name of project author(s): Muhammad Nasir

Name of project reviewer: Yufei Wu

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments

The introduction part provides a clear description of global TB burden and regional focus of South-East Asia and Africa. The previous research gap is also stated. Only several sentences need editing for grammar.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description

### Feedback and Comments

The research question is clearly described.

### Summary assessment
* question/hypotheses fully clear

## Data description

### Feedback and Comments

The data source is provided. Although the codebook is not provided, the major variables are enumerated here.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis

### Feedback and Comments

The cleaning steps are outlined. The overall exploratory analysis is good. However, the method part mentioned EDA included boxplots, but it is not actually embedded. Besides, the figure captions and links to figures do not seem to work properly.

### Summary assessment
* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments

The analysis is detailed and thorough, including AIC/RMSE comparisons among LM, GAM, GAMM, and GAMM + interactions, which is a good practice. In addition, it will be better to provide cross-validated RMSE to show predictive validity if possible.

### Summary assessment
* strong and reasonable analysis

## Presentation

### Feedback and Comments

Some figure captions are not rendered correctly, and some figure references show “?@fig-X”.

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions

### Feedback and Comments

The discussion part can link the findings to policy implications, which highlight the application of the project. The strengths and limitations are also acknowledged.

### Summary assessment
* strong, complete and clear discussion

## Further comments

Further polish the figures and tables to meet the publication standard.

# Overall project content evaluation

## Structure

### Feedback and Comments

The project is generally well structured. However, some template files need to be removed, like the "eda-v2.qmd", and the blank Supplementary-Material.

### Summary assessment
* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation

### Feedback and Comments

The project is well documented, except for the interpretation paragraph at the end of the EDA file, which would be better if written in English.

### Summary assessment
* fully and well documented

## Reproducibility

### Feedback and Comments

The manuscript and processing file can be fully reproduced. However, there are some issues in other code files. There is an error in line 292 of EDA file saying "Error in readRDS(con, refhook = refhook) : cannot open the connection", which suggests the file pathway ("code", "eda-code", "tables", "summary_table.rds") is not correct. There's also an error in the code chunk starting at line 109 in the statistical-analysis file, which says "Error: object 'asia_africa_long' not found".

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness

### Feedback and Comments

The study is thorough, with multiple models tested. The model assessment shows robustness.

### Summary assessment
* strong level of thorougness

## Further comments

The overall analysis is impressive. The manuscript may need further polishing to make sure the figures are of publication-level quality.
